---
title: "Economic Hardship in Maricopa County"
format:
  dashboard:
    theme:
      - darkly
      - ../../assets/dashboard-theme.scss
    orientation: columns
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
knitr::opts_chunk$set( echo = FALSE, message = FALSE, warning = FALSE )
options( tigris_use_cache = TRUE )
options( tigris_progress_bar = FALSE )
```

```{r}
#| label: packages
# ── Package Setup ──────────────────────────────────────────────────────
# R packages update frequently and can break working code. We have saved
# the exact package versions used to build this lab. Run the block below
# ONCE the first time you open this file, then comment it out.
# if (!requireNamespace("renv", quietly = TRUE)) install.packages("renv")
# download.file(
#   url      = "https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab7/renv.lock",
#   destfile = "renv.lock", mode = "wb"
# )
# renv::restore(lockfile = "renv.lock", prompt = FALSE)
# After running once, comment out the four lines above.
# ───────────────────────────────────────────────────────────────────────

# ── Packages ─────────────────────────────────────────
library( tidyverse )
library( tidycensus )
library( sf )
library( tigris )
library( spdep )
library( psych )
library( viridis )
library( mapgl )
library( DT )
```

```{r}
#| label: dark-theme
# ── Dark dashboard theme for ggplot2 static plots ──────────────
theme_dark_dash <- function() {
  theme_void(base_size = 12) +
  theme(
    plot.background   = element_rect(fill = "#0d1117", color = NA),
    panel.background  = element_rect(fill = "#0d1117", color = NA),
    plot.title        = element_text(color = "#e6edf3", size = 14, face = "bold"),
    plot.subtitle     = element_text(color = "#8b949e", size = 11),
    plot.caption      = element_text(color = "#8b949e", size = 9),
    legend.background = element_rect(fill = "#0d1117", color = NA),
    legend.text       = element_text(color = "#e6edf3"),
    legend.title      = element_text(color = "#8b949e")
  )
}
```

```{r}
#| label: api-key
# Uncomment and add your Census API key if not already installed
# census_api_key("YOUR KEY GOES HERE", install = TRUE)
```

```{r}
#| label: fallback-data
#| eval: true

# ── FALLBACK: Load pre-saved datasets if Census API is down ──────
# Uncomment the lines below if you cannot get a Census API key.
#
# county_complete <- readRDS("../data/county_complete.rds")
# cbg_complete    <- readRDS("../data/cbg_2023.rds")
# cbg_2015_raw    <- readRDS("../data/cbg_2015.rds")
#
# county_raw      <- readRDS(url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab7/data/county_complete.rds"))
# cbg_raw         <- readRDS(url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab7/data/cbg_2023.rds"))
# cbg_2015_raw    <- readRDS(url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab7/data/cbg_2015.rds"))
```

```{r}
#| label: define-vars

# ── Define the 5 hardship indicator variables ────────────────
hardship_vars <- c(
  pov_below50  = "C17002_002",   # Income-to-poverty ratio under .50
  pov_50to99   = "C17002_003",   # Income-to-poverty ratio .50 to .99
  pov_den      = "C17002_001",   # Poverty universe total
  unemp_num    = "B23025_005",
  unemp_den    = "B23025_002",
  no_hs_num    = "B15003_002",
  no_hs_den    = "B15003_001",
  no_ins_num   = "B27010_017",
  no_ins_den   = "B27010_001",
  median_income = "B19013_001"
)
```

```{r}
#| label: build-function

# ── Reusable function to build the hardship index ────────────
build_hardship_index <- function( df ) {
  wide <- df %>%
    select( -moe ) %>%
    pivot_wider( names_from = variable, values_from = estimate )

  wide <- wide %>%
    mutate(
      poverty_rate = ( pov_below50 + pov_50to99 ) / pov_den,
      unemp_rate   = unemp_num / unemp_den,
      no_hs_rate   = no_hs_num / no_hs_den,
      no_ins_rate  = no_ins_num / no_ins_den
    )

  # Filter to complete cases FIRST (keeping geometry intact)
  complete <- wide %>%
    filter( !is.na( poverty_rate ) & !is.na( unemp_rate ) &
            !is.na( no_hs_rate )   & !is.na( no_ins_rate ) &
            !is.na( median_income ) )

  if ( nrow( complete ) == 0 ) stop( "No complete cases — check variable codes" )

  # Z-score standardize (drop geometry temporarily for scale())
  nums <- complete %>%
    st_drop_geometry() %>%
    select( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income ) %>%
    mutate( across( everything(), ~ as.numeric( scale(.) ) ) )

  # Reverse-code median income
  nums$median_income <- -1 * nums$median_income

  # Composite index
  complete$hardship_index <- rowMeans( nums, na.rm = TRUE )

  return( complete )  # Returns sf object with geometry intact
}
```

```{r}
#| label: pull-county
#| results: hide

# ── Pull US county-level data ────────────────────────────────
county_raw <- get_acs(
  geography    = "county",
  variables    = hardship_vars,
  year         = 2023,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
```

```{r}
#| label: build-county-index
county_complete <- build_hardship_index( county_raw )
```

```{r}
#| label: pull-tract-2023
#| results: hide

# ── Pull AZ census tract data (2023) ─────────────────────────
tract_raw_2023 <- get_acs(
  geography    = "tract",
  variables    = hardship_vars,
  state        = "AZ",
  year         = 2023,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
```

```{r}
#| label: build-tract-2023

tract_complete_2023 <- build_hardship_index(tract_raw_2023)

# Extract county name from NAME field:
# NAME format: "Census Tract XXXX, Maricopa County, Arizona"
tract_complete_2023 <- tract_complete_2023 |>
  mutate(
    county_name  = str_extract(NAME, "(?<=, )([^,]+)(?=, Arizona)"),
    tract_number = str_extract(NAME, "(?<=Census Tract )([0-9.]+)")
  )
```

```{r}
#| label: pull-tract-2015
#| results: hide

# ── Pull AZ census tract data (2015) ─────────────────────────
tract_raw_2015 <- get_acs(
  geography    = "tract",
  variables    = hardship_vars,
  state        = "AZ",
  year         = 2015,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
```

```{r}
#| label: build-tract-2015

tract_complete_2015 <- build_hardship_index(tract_raw_2015)
```

```{r}
#| label: pull-maricopa-cbg
#| results: hide

# ── Pull Maricopa block groups (2023) — for LISA only ────────
cbg_raw <- get_acs(
  geography    = "block group",
  variables    = hardship_vars,
  state        = "AZ",
  county       = "Maricopa",
  year         = 2023,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
```

```{r}
#| label: build-cbg-index

cbg_complete <- build_hardship_index(cbg_raw)
```

```{r}
#| label: reliability

# ── Cronbach's alpha for index reliability ────────────────────
alpha_data <- cbg_complete |>
  st_drop_geometry() |>
  select( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, med_income = median_income ) |>
  mutate( med_income = -med_income ) |>
  drop_na()

alpha_result <- psych::alpha( alpha_data )
alpha_val    <- round( alpha_result$total$raw_alpha, 3 )
```

```{r}
#| label: lisa-analysis

# ── LISA Hot Spot Analysis ───────────────────────────────────
cbg_complete <- st_make_valid( cbg_complete )

nb <- poly2nb( cbg_complete, queen = TRUE )
lw <- nb2listw( nb, style = "W", zero.policy = TRUE )

lisa <- localmoran( cbg_complete$hardship_index, lw, zero.policy = TRUE )

cbg_complete$lisa_i <- lisa[, "Ii"]
cbg_complete$lisa_z <- lisa[, "Z.Ii"]
cbg_complete$lisa_p <- lisa[, "Pr(z != E(Ii))"]

cbg_complete$lag_hardship <- lag.listw( lw, cbg_complete$hardship_index,
                                         zero.policy = TRUE )
```

```{r}
#| label: lisa-classify

# ── Classify LISA clusters ───────────────────────────────────
cbg_complete <- cbg_complete %>%
  mutate(
    val_centered = hardship_index - mean( hardship_index, na.rm = TRUE ),
    lag_centered = lag_hardship   - mean( lag_hardship, na.rm = TRUE ),

    quadrant = case_when(
      val_centered > 0 & lag_centered > 0 ~ "High-High",
      val_centered < 0 & lag_centered < 0 ~ "Low-Low",
      val_centered > 0 & lag_centered < 0 ~ "High-Low",
      val_centered < 0 & lag_centered > 0 ~ "Low-High",
      TRUE ~ "Not Significant"
    ),

    lisa_cluster = if_else( lisa_p <= 0.05, quadrant, "Not Significant" )
  )

cbg_complete$lisa_cluster <- factor(
  cbg_complete$lisa_cluster,
  levels = c( "High-High", "Low-Low", "High-Low", "Low-High", "Not Significant" )
)
```

```{r}
#| label: compute-moran

moran_global <- moran.test(cbg_complete$hardship_index, lw, zero.policy = TRUE)
moran_i_val  <- round(moran_global$estimate["Moran I statistic"], 3)
```

```{r}
#| label: temporal-build

# ── AZ tract-level temporal change (2015 → 2023) ────────────
# full_join: keeps all 2023 tracts, NA where no 2015 match
tract_change <- full_join(
  st_drop_geometry(tract_complete_2015) |>
    select(GEOID, hardship_2015 = hardship_index),
  tract_complete_2023 |>
    select(GEOID, hardship_2023 = hardship_index,
           county_name, tract_number, geometry,
           poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income),
  by = "GEOID"
) |>
  st_as_sf() |>
  mutate(hardship_change = hardship_2023 - hardship_2015)
```

```{r}
#| label: compute-stats

# ── Summary statistics ────────────────────────────────────────
n_tract        <- nrow(tract_complete_2023)
n_county       <- nrow(county_complete)
n_cbg          <- nrow(cbg_complete)
n_hotspot      <- sum(cbg_complete$lisa_cluster == "High-High", na.rm = TRUE)
n_coldspot     <- sum(cbg_complete$lisa_cluster == "Low-Low",   na.rm = TRUE)
min_hardship    <- round(min(county_complete$hardship_index,    na.rm = TRUE), 2)
median_hardship <- round(median(county_complete$hardship_index, na.rm = TRUE), 2)
mean_hardship   <- round(mean(county_complete$hardship_index,   na.rm = TRUE), 2)
max_hardship    <- round(max(county_complete$hardship_index,    na.rm = TRUE), 2)

county_display <- county_complete |>
  st_drop_geometry() |>
  mutate(
    county_label  = str_extract(NAME, "^[^,]+"),
    state_label   = str_extract(NAME, "(?<=, ).+"),
    hardship_disp = round(hardship_index, 2)
  ) |>
  select(
    `County`  = county_label,
    `State`   = state_label,
    `Hardship Index` = hardship_disp
  ) |>
  arrange(desc(`Hardship Index`))

n_matched      <- sum(!is.na(tract_change$hardship_2015) & !is.na(tract_change$hardship_2023))
n_worsened     <- sum(tract_change$hardship_change > 0, na.rm = TRUE)
n_improved     <- sum(tract_change$hardship_change < 0, na.rm = TRUE)
pct_worsened   <- round(100 * n_worsened / n_matched, 1)
pct_improved   <- round(100 * n_improved / n_matched, 1)
mean_change    <- round(mean(tract_change$hardship_change, na.rm = TRUE), 3)
```


# Overview

## Column {width=60%}

### About This Dashboard

This dashboard examines **economic hardship across multiple geographic scales**
using data from the American Community Survey (ACS) 5-Year Estimates.

**What we measure:** A composite Economic Hardship Index constructed from
5 Census variables --- poverty rate, unemployment rate, lack of high school
education, uninsured rate, and median income (reverse-coded). Variables are
z-score standardized and averaged into a single index where higher values
indicate greater hardship.

**Geographic scales analyzed:**

- **National:** ~3,100 counties across the United States (ACS 2019-2023)
- **State:** ~1,500 census tracts across Arizona (ACS 2019-2023)
- **Temporal:** AZ census tract change between 2011-2015 and 2019-2023
- **Neighborhood:** ~900 block groups within Maricopa County for hot spot detection

**Policy focus:** Identifying persistent hot spots of economic hardship within
Maricopa County to support targeted intervention by county government and
community partners.

## Column {width=40%}

### Scale Summary

```{r}
#| label: overview-vbox1
#| content: valuebox
list(
  title = "US Counties",
  value = scales::comma(n_county),
  icon  = "globe-americas",
  color = "primary"
)
```

### AZ Census Tracts

```{r}
#| label: overview-vbox2
#| content: valuebox
list(
  title = "AZ Census Tracts",
  value = scales::comma(n_tract),
  icon  = "map",
  color = "info"
)
```

### Maricopa Block Groups

```{r}
#| label: overview-vbox3
#| content: valuebox
list(
  title = "Maricopa Block Groups",
  value = scales::comma(n_cbg),
  icon  = "geo-alt-fill",
  color = "secondary"
)
```

### Index Variables

```{r}
#| label: overview-vbox4
#| content: valuebox
list(
  title = "Index Variables",
  value = 5,
  icon  = "bar-chart-fill",
  color = "success"
)
```

### Methods

**Index Construction:** Each of the 5 indicators is z-score standardized
(mean = 0, SD = 1) across all geographies in each dataset. Median income
is reverse-coded so that higher values indicate greater hardship. The
composite index is the row mean of the 5 standardized values.

**Temporal Comparison:** 2011--2015 ACS 5-year estimates vs. 2019--2023
ACS 5-year estimates. These are non-overlapping windows enabling
valid before/after comparison.

**Spatial Clustering:** Local Moran's I (LISA) with queen contiguity
spatial weights identifies statistically significant clusters of high
or low hardship at p < 0.05.


# National Context

## Column {width=60%}

### US County Hardship Map

```{r}
#| label: national-map

# ── National mapgl interactive choropleth ────────────────────
# mapgl handles projection natively — no shift_geometry() needed
county_map <- county_complete |>
  st_transform(4326) |>
  mutate(hardship_round = round(hardship_index, 2))

maplibre(
  style  = openfreemap_style("dark"),
  bounds = st_bbox(county_map)
) |>
  add_source(
    id   = "county-src",
    type = "geojson",
    data = county_map
  ) |>
  add_fill_layer(
    id     = "county-fill",
    source = "county-src",
    fill_color = interpolate(
      column = "hardship_round",
      values = c(-2, 0, 2, 4),
      stops  = c("#4575b4", "#fee090", "#d73027", "#a50026"),
      na_color = "#2a2a2a"
    ),
    fill_opacity = 0.85,
    tooltip = concat(
      get_column("NAME"),
      " | Hardship: ", number_format(get_column("hardship_round"))
    )
  ) |>
  add_line_layer(
    id         = "county-border",
    source     = "county-src",
    line_color = "#30363d",
    line_width  = 0.3
  )
```

## Column {width=40%}

### Min Hardship

```{r}
#| label: min-vbox
#| content: valuebox
list(
  title = "Min County Hardship",
  value = min_hardship,
  icon  = "arrow-down-circle-fill",
  color = "success"
)
```

### Median Hardship

```{r}
#| label: median-vbox
#| content: valuebox
list(
  title = "Median County Hardship",
  value = median_hardship,
  icon  = "dash-circle-fill",
  color = "info"
)
```

### Mean Hardship

```{r}
#| label: mean-vbox
#| content: valuebox
list(
  title = "Mean County Hardship",
  value = mean_hardship,
  icon  = "bar-chart-fill",
  color = "warning"
)
```

### Max Hardship

```{r}
#| label: max-vbox
#| content: valuebox
list(
  title = "Max County Hardship",
  value = max_hardship,
  icon  = "exclamation-triangle-fill",
  color = "danger"
)
```

### US County Rankings

```{r}
#| label: county-table

DT::datatable(
  county_display,
  extensions = "Scroller",
  options = list(
    scrollY    = 350,
    scroller   = TRUE,
    paging     = TRUE,
    pageLength = 10
  ),
  rownames = FALSE,
  filter   = "top"
)
```


# State Patterns

## Column {width=60%}

### Arizona Census Tracts --- Economic Hardship Index {.no-title}

```{r}
#| label: state-map
# ── AZ census tract hardship mapgl ───────────────────────────
tract_map <- tract_complete_2023 |>
  st_transform(4326) |>
  mutate(
    hardship_round = round(hardship_index, 3),
    poverty_pct    = round(poverty_rate * 100, 1),
    unemp_pct      = round(unemp_rate * 100, 1),
    no_hs_pct      = round(no_hs_rate * 100, 1),
    no_ins_pct     = round(no_ins_rate * 100, 1),
    income_fmt     = scales::dollar(median_income)
  )

maplibre(
  style  = openfreemap_style("dark"),
  bounds = st_bbox(tract_map)
) |>
  add_source(id = "tract-src", type = "geojson", data = tract_map) |>
  add_fill_layer(
    id     = "tract-fill",
    source = "tract-src",
    fill_color = interpolate(
      column = "hardship_round",
      values = c(-2, 0, 2, 4),
      stops  = c("#4575b4", "#fee090", "#d73027", "#a50026"),
      na_color = "#2a2a2a"
    ),
    fill_opacity = 0.8,
    hover_options = list(fill_opacity = 1),
    tooltip = concat(
      get_column("tract_number"), " — ", get_column("county_name"),
      " | Hardship: ", number_format(get_column("hardship_round")),
      " | Poverty: ", get_column("poverty_pct"), "%",
      " | Unemp: ",   get_column("unemp_pct"), "%",
      " | Uninsured: ", get_column("no_ins_pct"), "%"
    )
  ) |>
  add_line_layer(
    id = "tract-border", source = "tract-src",
    line_color = "#30363d", line_width = 0.3
  )
```

## Column {width=40%}

### Census Tract Details

```{r}
#| label: state-table

tract_display <- tract_complete_2023 |>
  st_drop_geometry() |>
  mutate(
    hardship_round = round(hardship_index, 3),
    poverty_pct    = round(poverty_rate * 100, 1),
    unemp_pct      = round(unemp_rate * 100, 1),
    no_hs_pct      = round(no_hs_rate * 100, 1),
    no_ins_pct     = round(no_ins_rate * 100, 1),
    income_fmt     = scales::dollar(median_income)
  ) |>
  select(
    `Tract`         = tract_number,
    `County`        = county_name,
    `Hardship`      = hardship_round,
    `Poverty %`     = poverty_pct,
    `Unemp %`       = unemp_pct,
    `No HS %`       = no_hs_pct,
    `Uninsured %`   = no_ins_pct,
    `Median Income` = income_fmt
  ) |>
  arrange(desc(Hardship))

DT::datatable(
  tract_display,
  extensions = "Scroller",
  options = list(
    scrollY    = 450,
    scroller   = TRUE,
    paging     = TRUE,
    pageLength = 50
  ),
  rownames = FALSE,
  filter   = "top"
)
```

### Index Reliability

```{r}
#| label: alpha-vbox
#| content: valuebox
list(
  title = "Cronbach's Alpha",
  value = alpha_val,
  icon  = "check-circle-fill",
  color = if (alpha_val >= 0.7) "success" else "warning"
)
```


# Change Over Time

## Column {width=60%}

### Arizona Census Tract Change: 2015 → 2023 {.no-title}

```{r}
#| label: change-map

change_map_sf <- tract_change |>
  filter(!is.na(geometry)) |>
  st_transform(4326) |>
  mutate(change_round = round(hardship_change, 2))

maplibre(
  style  = openfreemap_style("dark"),
  bounds = st_bbox(change_map_sf)
) |>
  add_source(id = "change-src", type = "geojson", data = change_map_sf) |>
  add_fill_layer(
    id     = "change-fill",
    source = "change-src",
    fill_color = interpolate(
      column   = "change_round",
      values   = c(-3, 0, 3),
      stops    = c("#4575b4", "#f7f7f7", "#d73027"),
      na_color = "#3a3a3a"
    ),
    fill_opacity = 0.85,
    tooltip = concat(
      get_column("tract_number"), " — ", get_column("county_name"),
      " | Change: ", number_format(get_column("change_round"))
    )
  ) |>
  add_line_layer(
    id = "change-border", source = "change-src",
    line_color = "#1a1a2e", line_width = 0.3
  )
```

## Row

### Hardship Change by County: 2015 vs 2023

```{r}
#| label: timeseries-plot
#| fig-height: 5

# ── County-level mean hardship for top 12 counties ───────────
county_ts <- tract_change |>
  st_drop_geometry() |>
  filter(!is.na(hardship_2015), !is.na(hardship_2023),
         !is.na(county_name)) |>
  group_by(county_name) |>
  summarise(
    mean_2015 = mean(hardship_2015, na.rm = TRUE),
    mean_2023 = mean(hardship_2023, na.rm = TRUE),
    n_tracts  = n(),
    .groups   = "drop"
  ) |>
  filter(n_tracts >= 3) |>           # exclude tiny counties
  arrange(desc(mean_2023)) |>
  slice_head(n = 12) |>
  pivot_longer(
    cols      = c(mean_2015, mean_2023),
    names_to  = "year",
    values_to = "hardship"
  ) |>
  mutate(
    year         = ifelse(year == "mean_2015", "2015", "2023"),
    county_short = str_remove(county_name, " County$")
  )

ggplot(county_ts,
       aes(x = year, y = hardship,
           color = county_short, group = county_short)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_color_manual(
    values = c(
      "Maricopa"  = "#d73027", "Pima"       = "#fc8d59",
      "Pinal"     = "#fee090", "Yuma"       = "#4575b4",
      "Cochise"   = "#91bfdb", "Mohave"     = "#e0f3f8",
      "Navajo"    = "#a50026", "Apache"     = "#f46d43",
      "Coconino"  = "#74add1", "Yavapai"    = "#abd9e9",
      "Graham"    = "#fdae61", "Greenlee"   = "#d9ef8b"
    )
  ) +
  labs(
    title    = "Mean Economic Hardship Index by County",
    subtitle = "Arizona census tracts · ACS 2011–2015 vs 2019–2023",
    x = NULL, y = "Mean Hardship Index", color = NULL
  ) +
  theme_dark_dash() +
  theme(
    legend.position  = "right",
    axis.text        = element_text(color = "#e6edf3"),
    axis.title       = element_text(color = "#8b949e"),
    panel.grid.major = element_line(color = "#21262d"),
    panel.grid.minor = element_blank()
  )
```

## Column {width=40%}

### Summary

Comparing 2011--2015 and 2019--2023 ACS estimates across Arizona census
tracts, most counties show modest changes. Persistently high-hardship
tracts concentrate in tribal areas (Apache, Navajo counties), border
communities (Yuma, Cochise), and urban cores of Maricopa and Pima counties.

The line chart shows mean hardship for the 12 largest counties. Lines
above zero indicate above-average hardship; lines sloping upward indicate
worsening conditions. This statewide view places Maricopa County's
neighborhood-level analysis (Hot Spots tab) in regional context.

### Tracts Matched

```{r}
#| label: match-vbox
#| content: valuebox
list(
  title = "Tracts in Both Periods",
  value = scales::comma(n_matched),
  icon  = "arrow-left-right",
  color = "secondary"
)
```

### Hardship Worsened

```{r}
#| label: worse-vbox
#| content: valuebox
list(
  title = "Tracts Where Hardship Increased",
  value = paste0(pct_worsened, "%"),
  icon  = "arrow-up-circle-fill",
  color = "danger"
)
```

### Hardship Improved

```{r}
#| label: improve-vbox
#| content: valuebox
list(
  title = "Tracts Where Hardship Decreased",
  value = paste0(pct_improved, "%"),
  icon  = "arrow-down-circle-fill",
  color = "success"
)
```


# Hot Spots

## Column {width=60%}

### LISA Cluster Map

*Neighborhood-level spatial clustering --- Maricopa County Census Block Groups.
**High-High** = hot spot (high hardship surrounded by high hardship).
**Low-Low** = cold spot. Moran's I = `r moran_i_val` indicates moderate positive clustering.*

```{r}
#| label: lisa-map

# ── LISA cluster mapgl interactive map ───────────────────────
cbg_lisa <- cbg_complete |>
  st_transform(4326) |>
  mutate(
    cluster_label = as.character(lisa_cluster),
    cluster_color = case_when(
      cluster_label == "High-High"       ~ "#d73027",
      cluster_label == "Low-Low"         ~ "#4575b4",
      cluster_label == "High-Low"        ~ "#fc8d59",
      cluster_label == "Low-High"        ~ "#91bfdb",
      cluster_label == "Not Significant" ~ "#6e7681",
      TRUE ~ "#6e7681"
    )
  )

maplibre(style = openfreemap_style("dark"),
         bounds = st_bbox(cbg_lisa)) |>
  add_source(
    id   = "lisa-src",
    type = "geojson",
    data = cbg_lisa
  ) |>
  add_fill_layer(
    id     = "lisa-fill",
    source = "lisa-src",
    fill_color = get_column("cluster_color"),
    fill_opacity = 0.85,
    tooltip = concat(
      "Cluster: ", get_column("cluster_label"),
      " | Hardship: ", number_format(get_column("hardship_index"))
    )
  ) |>
  add_line_layer(
    id         = "lisa-border",
    source     = "lisa-src",
    line_color = "#1a1a2e",
    line_width  = 0.3
  )
```

## Column {width=40%}

### Key Statistics

```{r}
#| label: hotspot-vbox
#| content: valuebox
list(
  title = "Hot Spot CBGs (High-High)",
  value = n_hotspot,
  icon  = "fire",
  color = "danger"
)
```

### Cold Spot Count

```{r}
#| label: coldspot-vbox
#| content: valuebox
list(
  title = "Cold Spot CBGs (Low-Low)",
  value = n_coldspot,
  icon  = "snow2",
  color = "info"
)
```

### Moran's I

```{r}
#| label: moran-vbox
#| content: valuebox
list(
  title = "Global Moran's I",
  value = moran_i_val,
  icon  = "diagram-3-fill",
  color = "warning"
)
```

### Hot Spot Block Groups

```{r}
#| label: hotspot-table

# ── Table of HH and LL cluster CBGs ─────────────────────────
cluster_tbl <- cbg_complete %>%
  st_drop_geometry() %>%
  filter( lisa_cluster %in% c( "High-High", "Low-Low" ) ) %>%
  select(
    GEOID,
    Cluster       = lisa_cluster,
    Hardship      = hardship_index,
    `Poverty %`   = poverty_rate,
    `Unemp %`     = unemp_rate,
    `No HS %`     = no_hs_rate,
    `Uninsured %` = no_ins_rate
  ) %>%
  mutate(
    Hardship      = round( Hardship, 3 ),
    `Poverty %`   = round( `Poverty %` * 100, 1 ),
    `Unemp %`     = round( `Unemp %` * 100, 1 ),
    `No HS %`     = round( `No HS %` * 100, 1 ),
    `Uninsured %` = round( `Uninsured %` * 100, 1 )
  ) %>%
  arrange( desc( Hardship ) )

DT::datatable(
  cluster_tbl,
  extensions = "Scroller",
  options    = list(
    scrollY    = 300,
    scroller   = TRUE,
    paging     = TRUE,
    pageLength = 50
  ),
  rownames = FALSE,
  filter   = "top"
)
```


# Policy

## Column {width=50%}

### Policy Indicators

```{r}
#| label: policy-vbox1
#| content: valuebox
list(
  title = "Persistent Hot Spots",
  value = n_hotspot,
  icon  = "exclamation-octagon-fill",
  color = "danger"
)
```

### Mean Change

```{r}
#| label: policy-vbox2
#| content: valuebox
list(
  title = "Mean Hardship Change",
  value = mean_change,
  icon  = "graph-up",
  color = "warning"
)
```

### Priority Areas for Intervention

```{r}
#| label: priority-table

# ── Priority areas table ─────────────────────────────────────
# Identifies the top 20 highest-hardship hot spot CBGs as
# priority targets for policy intervention.
priority <- cbg_complete %>%
  st_drop_geometry() %>%
  filter( lisa_cluster == "High-High" ) %>%
  select(
    GEOID,
    Hardship      = hardship_index,
    `Poverty %`   = poverty_rate,
    `Unemp %`     = unemp_rate,
    `No HS %`     = no_hs_rate,
    `Uninsured %` = no_ins_rate,
    `Median Income` = median_income
  ) %>%
  mutate(
    Hardship        = round( Hardship, 3 ),
    `Poverty %`     = round( `Poverty %` * 100, 1 ),
    `Unemp %`       = round( `Unemp %` * 100, 1 ),
    `No HS %`       = round( `No HS %` * 100, 1 ),
    `Uninsured %`   = round( `Uninsured %` * 100, 1 ),
    `Median Income` = scales::dollar( `Median Income` )
  ) %>%
  arrange( desc( Hardship ) ) %>%
  head( 20 )

DT::datatable(
  priority,
  extensions = "Scroller",
  options    = list(
    scrollY    = 350,
    scroller   = TRUE,
    paging     = TRUE,
    pageLength = 20
  ),
  rownames = FALSE,
  caption  = "Top 20 Highest-Hardship Hot Spot Block Groups"
)
```

## Column {width=50%}

### Recommendations

**Recommendation 1: Targeted Workforce Development in South Phoenix**

The south Phoenix corridor is the largest and most persistent hardship hot spot. Unemployment rates are 2-3x the county average, with no meaningful improvement between 2015 and 2023. Invest in a workforce development center co-located with light rail infrastructure, focused on healthcare, logistics, and construction sectors with wraparound services for childcare and transportation.

**Recommendation 2: Health Insurance Enrollment Outreach in West Phoenix**

The Maryvale hot spot shows elevated uninsurance rates that persist despite Medicaid expansion. Deploy targeted door-to-door enrollment campaigns in High-High cluster block groups, partnering with community health centers and bilingual navigators.

**Recommendation 3: Affordable Housing Preservation in Central Mesa**

The central Mesa hot spot shows signs of expansion between 2015 and 2023. Pursue an affordable housing preservation strategy including acquisition of naturally occurring affordable housing (NOAH) properties, rehabilitation investment, and zoning reforms allowing gentle density near light rail stations.

### Data Sources & Methods

- **Data**: American Community Survey 5-Year Estimates via tidycensus
- **Periods**: 2011--2015 (ACS 2015) and 2019--2023 (ACS 2023)
- **Geographies**: US counties (national); AZ census tracts (state/temporal);
  Maricopa County block groups (hot spot detection)
- **Index**: 5-variable composite (poverty, unemployment, education,
  insurance, income) using z-score standardization
- **Clustering**: Local Moran's I (LISA) with queen contiguity spatial weights,
  p < 0.05 significance threshold
- **Tools**: R, Quarto Dashboard, mapgl (maplibre), DT, spdep, psych

### Download the Example File

[Right-click → Save As: Final_Project_Example.qmd](https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab7/Solutions/Final_Project_Example.qmd)

Open in RStudio. Replace the `define-vars` chunk with your own 5--10 ACS
variables to build your own index. See Module 7 Guide for full instructions.
