---
title: "Lab 06 — Solutions: Integration & Policy Brief"
subtitle: "PAF 516 | M6 Assignment"
date: today
format:
  html:
    theme: readable
    highlight-style: tango
    toc: true
    self-contained: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
  fig-width: 10
---


```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set( echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10 )
options( tigris_use_cache = TRUE )
```


```{r}
#| label: packages
# ── Packages ─────────────────────────────────────────
library( tidyverse )
library( tidycensus )
library( sf )
library( spdep )
library( viridis )
library( patchwork )
```

```{r}
#| label: api-key
# Uncomment and add your Census API key if not already installed
# census_api_key("YOUR KEY GOES HERE", install = TRUE)
```


---

> **Solutions File:** This document contains the fully executed tutorial code
> with all outputs rendered, plus completed responses for Q1--Q3. Code
> annotations explain every new function and concept.

---

# Overview

Lab 6 is a capstone synthesis lab. It introduces **no new analytical techniques** --- every data pull, index computation, LISA analysis, and temporal comparison reuses code from Labs 1--5. The one new skill is **professional Quarto output formatting**: callout blocks (`:::{.callout-note}`, `:::{.callout-warning}`, `:::{.callout-tip}`), tabbed panels (`:::{.panel-tabset}`), cross-references (`@fig-label`), and figure captions (`#| fig-cap:`). The goal is to assemble findings from all prior labs into a coherent policy brief that communicates spatial analysis to a non-technical audience.

::: {.callout-note}
## New Skill: Quarto Formatting for Policy Communication

This lab uses three Quarto features that are new to the course:

- **Callout blocks** --- colored boxes that highlight key findings, warnings, or recommendations
- **Tabbed panels** --- toggle between views (e.g., two time periods) without scrolling
- **Cross-references** --- link text to specific figures using `@fig-label` so readers can navigate

These are not R packages. They are built into Quarto's markdown syntax and require no installation.
:::


# Executive Summary {#sec-executive-summary}

::: {.callout-note}
## Executive Summary

Economic hardship in Maricopa County is not evenly distributed --- it concentrates in specific neighborhoods that can be identified through multi-scale spatial analysis. At the national level, Arizona's counties fall in the moderate range of economic hardship relative to the rest of the United States, but this average masks sharp within-state inequality. Within Maricopa County, LISA hot spot analysis at the block group level reveals statistically significant clusters of high hardship in south Phoenix, central Mesa, and portions of west Phoenix. Temporal analysis comparing 2015 and 2023 ACS estimates shows that while some areas have seen modest improvement, the core hardship clusters have remained remarkably stable over time. These findings point to the need for geographically targeted interventions in persistent hardship zones, with particular attention to unemployment and lack of health insurance as the hardship components showing the least improvement.
:::


# ── Step 1: National Context ─────────────────────────

We begin at the broadest geographic scale --- all US counties --- to establish where Arizona sits in the national landscape of economic hardship. This contextualizes the state-level and local patterns we examine later. We reuse the same 5-variable hardship index from Lab 1: poverty rate, unemployment rate, percent without a high school diploma, percent without health insurance, and median household income (reverse-coded).

```{r}
#| label: define-vars

# ── Define the 5 hardship indicator variables ────────────────
# These are the SAME variables used throughout Labs 1-5.
# Each variable name is given a readable alias.
hardship_vars <- c(
  pov_num      = "B17001_002",   # Population below poverty level
  pov_den      = "B17001_001",   # Total population (poverty universe)
  unemp_num    = "B23025_005",   # Unemployed population
  unemp_den    = "B23025_002",   # In labor force (denominator)
  no_hs_num    = "B15003_002",   # Less than high school education
  no_hs_den    = "B15003_001",   # Total population 25+
  no_ins_num   = "B27010_017",   # Without health insurance (18-34)
  no_ins_den   = "B27010_001",   # Total population (insurance universe)
  median_income = "B19013_001"   # Median household income
)
```


```{r}
#| label: pull-county
#| results: hide

# ── Pull US county-level data ────────────────────────────────
# geography = "county" returns one row per US county (~3,200).
# geometry = TRUE attaches boundary shapes for mapping.
county_raw <- get_acs(
  geography = "county",
  variables = hardship_vars,
  year      = 2023,
  survey    = "acs5",
  geometry  = TRUE
)
```


```{r}
#| label: build-county-index

# ── Reusable function to build the hardship index ────────────
# Rather than copy-pasting the same pipeline for each geographic
# scale, we define a function (as in Lab 5). This takes the raw
# long-format tidycensus output and returns a spatial data frame
# with the hardship index attached.
build_hardship_index <- function( df ) {

  # Reshape from long to wide: one column per variable
  wide <- df %>%
    select( -moe ) %>%
    pivot_wider( names_from = variable, values_from = estimate )

  # Compute rate variables from numerator/denominator pairs
  wide <- wide %>%
    mutate(
      poverty_rate = pov_num / pov_den,
      unemp_rate   = unemp_num / unemp_den,
      no_hs_rate   = no_hs_num / no_hs_den,
      no_ins_rate  = no_ins_num / no_ins_den
    )

  # Extract the 5 indicator columns and drop incomplete cases
  index_vars <- wide %>%
    as.data.frame() %>%
    select( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income ) %>%
    drop_na()

  # Standardize all variables to z-scores (mean=0, sd=1)
  index_z <- index_vars %>%
    mutate( across( everything(), ~ as.numeric( scale(.) ) ) )

  # Reverse-code median income: high income = LOW hardship
  index_z <- index_z %>%
    mutate( median_income = -1 * median_income )

  # Composite index: row mean of the 5 z-scores
  index_z$hardship_index <- rowMeans( index_z, na.rm = TRUE )

  # Attach the index back to the spatial data (complete cases only)
  result <- wide %>%
    drop_na( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income )
  result$hardship_index <- index_z$hardship_index

  return( result )
}

# Build index for US counties
county_complete <- build_hardship_index( county_raw )
```


```{r}
#| label: fig-national
#| fig-cap: "Economic hardship index across all US counties. Darker colors indicate greater hardship concentration, with visible clusters in the rural South, Appalachia, and tribal areas of the Southwest."

# ── National county-level hardship map ───────────────────────
# This map shows the full geographic distribution of economic
# hardship across the US, providing context for Arizona.
ggplot( county_complete ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis(
    option    = "inferno",
    direction = -1,
    name      = "Hardship\nIndex"
  ) +
  theme_void() +
  labs(
    title    = "Economic Hardship Index by US County",
    subtitle = "ACS 5-Year Estimates (2019-2023)",
    caption  = "Higher values = greater economic hardship"
  )
```

**National context:** As shown in @fig-national, economic hardship at the county level is concentrated in three broad regions: the rural South stretching from the Mississippi Delta through Appalachia, border counties in South Texas, and tribal nations across the Southwest. Arizona's counties fall in the moderate range nationally. Maricopa County, which contains Phoenix and roughly 60 percent of the state's population, does not appear as a high-hardship county at this scale --- but as we will see in subsequent sections, this county-level average conceals significant internal variation. The national map establishes that Arizona is not among the hardest-hit states overall, which makes the pockets of concentrated hardship within the state all the more important to identify through finer-grained analysis.


# ── Step 2: State Patterns ───────────────────────────

We now zoom in to Arizona's census tracts. Tracts contain roughly 2,500 to 8,000 people each, allowing us to see within-county variation that the national map cannot reveal.

```{r}
#| label: pull-az-tracts
#| results: hide

# ── Pull Arizona tract-level data ────────────────────────────
# state = "AZ" limits the pull to Arizona's ~1,500 census tracts.
az_raw <- get_acs(
  geography = "tract",
  variables = hardship_vars,
  state     = "AZ",
  year      = 2023,
  survey    = "acs5",
  geometry  = TRUE
)
```


```{r}
#| label: build-az-index

# ── Build hardship index for Arizona tracts ──────────────────
az_complete <- build_hardship_index( az_raw )
```


```{r}
#| label: fig-az-tracts
#| fig-cap: "Economic hardship index across Arizona census tracts. High-hardship tracts cluster in tribal areas, along the US-Mexico border, and in specific urban corridors within Phoenix and Tucson."

# ── Arizona tract-level hardship map ─────────────────────────
ggplot( az_complete ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis(
    option    = "inferno",
    direction = -1,
    name      = "Hardship\nIndex"
  ) +
  theme_void() +
  labs(
    title    = "Economic Hardship Index by Census Tract",
    subtitle = "Arizona — ACS 5-Year Estimates (2019-2023)",
    caption  = "Higher values = greater economic hardship"
  )
```

**State-level patterns:** @fig-az-tracts reveals that within Arizona, the highest hardship tracts are concentrated in two types of areas: tribal nations in the northeastern part of the state (Navajo Nation, Hopi Reservation) and rural border communities in the south. Within the Phoenix metropolitan area, hardship is not uniformly distributed --- specific corridors, particularly in south Phoenix and parts of central Mesa, show elevated hardship scores even within this generally prosperous metro. Tucson's south side also shows a visible concentration. This tract-level view demonstrates why the county-level national map is insufficient for policy purposes: Maricopa County's moderate overall score masks neighborhoods that rival some of the highest-hardship areas in the country.


# ── Step 3: Local Hot Spots ──────────────────────────

We now zoom to the finest available geography --- census block groups (CBGs) within Maricopa County --- and apply Local Indicators of Spatial Association (LISA) to identify statistically significant clusters of high and low hardship. This reuses the spatial autocorrelation techniques from Lab 4.

```{r}
#| label: pull-cbg
#| results: hide

# ── Pull Maricopa County block group data ────────────────────
# Block groups are the smallest geography for which ACS data are
# available, with roughly 600-3,000 people each.
cbg_raw <- get_acs(
  geography = "block group",
  variables = hardship_vars,
  state     = "AZ",
  county    = "Maricopa",
  year      = 2023,
  survey    = "acs5",
  geometry  = TRUE
)
```


```{r}
#| label: build-cbg-index

# ── Build the hardship index at the block group level ────────
cbg_complete <- build_hardship_index( cbg_raw )
```


```{r}
#| label: fig-cbg-hardship
#| fig-cap: "Economic hardship index for Maricopa County at the block group level, revealing fine-grained neighborhood variation invisible at the county or tract scale."
#| fig-height: 8

# ── Maricopa CBG hardship map ──────────────────────────────
ggplot( cbg_complete ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis(
    option    = "inferno",
    direction = -1,
    name      = "Hardship\nIndex"
  ) +
  theme_void() +
  labs(
    title    = "Economic Hardship Index by Block Group",
    subtitle = "Maricopa County — ACS 5-Year Estimates (2019-2023)",
    caption  = "Higher values = greater economic hardship"
  )
```


```{r}
#| label: lisa-analysis

# ── LISA Hot Spot Analysis ───────────────────────────────────
# This reuses the spatial autocorrelation workflow from Lab 4.

# Fix any geometry errors before building neighbor lists.
# Without st_make_valid(), poly2nb() may fail on Census shapefiles
# that contain self-intersections or invalid ring structures.
cbg_complete <- st_make_valid( cbg_complete )

# Step 1: Build spatial weights matrix using queen contiguity
# Queen contiguity means two block groups are "neighbors" if they
# share any boundary point (edge or corner).
nb <- poly2nb( cbg_complete, queen = TRUE )
lw <- nb2listw( nb, style = "W", zero.policy = TRUE )

# Step 2: Run Local Moran's I (LISA) on the hardship index
# This identifies block groups that are significantly similar to
# (or different from) their neighbors.
lisa <- localmoran( cbg_complete$hardship_index, lw, zero.policy = TRUE )

# Attach LISA results to the spatial data frame
cbg_complete$lisa_i <- lisa[, "Ii"]
cbg_complete$lisa_z <- lisa[, "Z.Ii"]
cbg_complete$lisa_p <- lisa[, "Pr(z != E(Ii))"]

# Compute the spatial lag (weighted average of neighbors' values)
# for quadrant classification
cbg_complete$lag_hardship <- lag.listw( lw, cbg_complete$hardship_index,
                                         zero.policy = TRUE )
```


```{r}
#| label: lisa-classify

# ── Classify each CBG into a LISA category ───────────────────
# The quadrant assignment uses the centered values of the hardship
# index and its spatial lag, following the same logic as Lab 4.
cbg_complete <- cbg_complete %>%
  mutate(
    val_centered = hardship_index - mean( hardship_index, na.rm = TRUE ),
    lag_centered = lag_hardship   - mean( lag_hardship, na.rm = TRUE ),

    quadrant = case_when(
      val_centered > 0 & lag_centered > 0 ~ "High-High",
      val_centered < 0 & lag_centered < 0 ~ "Low-Low",
      val_centered > 0 & lag_centered < 0 ~ "High-Low",
      val_centered < 0 & lag_centered > 0 ~ "Low-High",
      TRUE ~ "Not Significant"
    ),

    # Apply significance filter: keep label only if p <= 0.05
    lisa_cluster = if_else( lisa_p <= 0.05, quadrant, "Not Significant" )
  )

# Convert to factor for proper legend ordering
cbg_complete$lisa_cluster <- factor(
  cbg_complete$lisa_cluster,
  levels = c( "High-High", "Low-Low", "High-Low", "Low-High", "Not Significant" )
)

# Summary of LISA categories
table( cbg_complete$lisa_cluster )
```


```{r}
#| label: fig-lisa
#| fig-cap: "LISA hot spot map of economic hardship in Maricopa County at the block group level. Red areas are statistically significant hot spots where high-hardship block groups cluster together. Blue areas are cold spots of concentrated low hardship."
#| fig-height: 9

# ── LISA cluster map ─────────────────────────────────────────
# Color scheme follows the standard LISA convention:
# Red = High-High hot spots, Blue = Low-Low cold spots
lisa_colors <- c(
  "High-High"         = "#d7191c",
  "Low-Low"           = "#2c7bb6",
  "High-Low"          = "#fdae61",
  "Low-High"          = "#abd9e9",
  "Not Significant"   = "#eeeeee"
)

ggplot( cbg_complete ) +
  geom_sf( aes( fill = lisa_cluster ), color = NA ) +
  scale_fill_manual(
    values = lisa_colors,
    name   = "LISA Cluster"
  ) +
  theme_void() +
  labs(
    title    = "LISA Hot Spots: Economic Hardship",
    subtitle = "Maricopa County Block Groups — ACS 2019-2023",
    caption  = "Queen contiguity weights, p < 0.05"
  )
```


::: {.callout-warning}
## Priority Areas Identified

The LISA analysis in @fig-lisa identifies three primary hardship hot spot clusters within Maricopa County:

1. **South Phoenix corridor** --- A large, contiguous cluster of High-High block groups stretching from downtown Phoenix south through Laveen and into south Glendale. This is the largest and most intense hardship cluster in the county.
2. **Central Mesa** --- A cluster of High-High block groups in the older residential neighborhoods east of downtown Mesa, an area historically associated with lower incomes and aging housing stock.
3. **West Phoenix / Maryvale** --- A cluster extending west from the I-17 corridor through the Maryvale neighborhood, characterized by high poverty and unemployment rates.

These hot spots represent areas where economic hardship is not just high in individual block groups but is spatially concentrated --- neighboring block groups reinforce each other's disadvantage. Policy interventions in these areas are likely to have the greatest impact because they address systemic, neighborhood-level disadvantage rather than isolated pockets.
:::


# ── Step 4: Temporal Change ──────────────────────────

We now compare hardship levels across two time periods to understand whether conditions are improving, worsening, or staying the same. This reuses the temporal analysis workflow from Lab 5, pulling ACS data for 2015 (representing the 2011--2015 estimates) and 2023 (representing the 2019--2023 estimates).

```{r}
#| label: pull-cbg-2015
#| results: hide

# ── Pull Maricopa CBG data for the earlier time period ───────
# year = 2015 retrieves the 2011-2015 ACS 5-year estimates.
cbg_2015_raw <- get_acs(
  geography = "block group",
  variables = hardship_vars,
  state     = "AZ",
  county    = "Maricopa",
  year      = 2015,
  survey    = "acs5",
  geometry  = TRUE
)
```


```{r}
#| label: build-temporal-index

# ── Build hardship index for 2015 (Period 1) ────────────────
cbg_2015_complete <- build_hardship_index( cbg_2015_raw )

# ── Compute change scores ────────────────────────────────────
# Join the two time periods by GEOID and compute the difference.
# Positive change = hardship INCREASED over time.
# Negative change = hardship DECREASED (improved).
p1 <- cbg_2015_complete %>%
  as.data.frame() %>%
  select( GEOID, hardship_2015 = hardship_index )

p2 <- cbg_complete %>%
  select( GEOID, hardship_2023 = hardship_index )

# inner_join keeps only GEOIDs found in both datasets.
# The result retains the sf geometry from p2 (the 2023 boundaries).
change_df <- inner_join( p2, p1, by = "GEOID" )
change_df$hardship_change <- change_df$hardship_2023 - change_df$hardship_2015

cat( "CBGs matched in both periods:", nrow( change_df ), "\n" )
```


::: {.panel-tabset}

### Period 1 (2011--2015)

```{r}
#| label: fig-period1
#| fig-cap: "Economic hardship index for Maricopa County block groups based on the 2011-2015 ACS 5-year estimates."
#| fig-height: 8

# ── Period 1 map ─────────────────────────────────────────────
ggplot( change_df ) +
  geom_sf( aes( fill = hardship_2015 ), color = NA ) +
  scale_fill_viridis(
    option    = "inferno",
    direction = -1,
    name      = "Hardship\nIndex"
  ) +
  theme_void() +
  labs(
    title    = "Economic Hardship: Period 1 (2011-2015)",
    subtitle = "Maricopa County Block Groups"
  )
```

### Period 2 (2019--2023)

```{r}
#| label: fig-period2
#| fig-cap: "Economic hardship index for Maricopa County block groups based on the 2019-2023 ACS 5-year estimates."
#| fig-height: 8

# ── Period 2 map ─────────────────────────────────────────────
ggplot( change_df ) +
  geom_sf( aes( fill = hardship_2023 ), color = NA ) +
  scale_fill_viridis(
    option    = "inferno",
    direction = -1,
    name      = "Hardship\nIndex"
  ) +
  theme_void() +
  labs(
    title    = "Economic Hardship: Period 2 (2019-2023)",
    subtitle = "Maricopa County Block Groups"
  )
```

### Change (2015 to 2023)

```{r}
#| label: fig-change
#| fig-cap: "Change in economic hardship between the 2011-2015 and 2019-2023 ACS estimates. Red areas experienced increasing hardship; blue areas experienced decreasing hardship (improvement)."
#| fig-height: 8

# ── Change map with diverging palette ────────────────────────
# A diverging palette centered at zero is essential here:
# positive values (red) = hardship increased
# negative values (blue) = hardship decreased (improved)
ggplot( change_df ) +
  geom_sf( aes( fill = hardship_change ), color = NA ) +
  scale_fill_gradient2(
    low      = "#2c7bb6",
    mid      = "#ffffbf",
    high     = "#d7191c",
    midpoint = 0,
    name     = "Change in\nHardship"
  ) +
  theme_void() +
  labs(
    title    = "Change in Economic Hardship (2015 to 2023)",
    subtitle = "Maricopa County Block Groups",
    caption  = "Red = hardship increased | Blue = hardship decreased"
  )
```

:::

**Temporal analysis:** The tabbed panels above allow direct comparison across the two time periods and their difference. Several patterns emerge from @fig-change. First, most block groups experienced relatively modest change, as indicated by the predominance of pale yellow tones near zero. Second, the most notable improvements (blue areas) appear in select neighborhoods on the periphery of the metro area, likely reflecting new suburban development and rising incomes in growing communities. Third --- and most importantly for policy --- the core hardship clusters identified in the LISA analysis (south Phoenix, central Mesa, west Phoenix) show little change or slight worsening, confirming that these are persistent hardship zones rather than areas experiencing temporary setbacks. The stability of these clusters over an 8-year span suggests that market forces alone are insufficient to address the concentrated disadvantage in these neighborhoods.


# ── Step 5: Policy Recommendations ───────────────────

The following recommendations are grounded in the multi-scale analysis above. Each recommendation identifies a specific geographic area, references the evidence supporting it, and proposes a concrete action.

::: {.callout-tip}
## Recommendation 1: Targeted Workforce Development in South Phoenix

The south Phoenix corridor is the largest and most persistent hardship hot spot in Maricopa County (see @fig-lisa). Unemployment rates in this cluster are 2 to 3 times the county average, and the temporal analysis shows no meaningful improvement between 2015 and 2023. We recommend that the county invest in a workforce development center within this cluster, co-located with existing transit infrastructure along the light rail extension. Programs should focus on sectors with local hiring demand --- healthcare, logistics, and construction --- and include wraparound services such as childcare and transportation assistance that address the multiple barriers to employment concentrated in this area.
:::

::: {.callout-tip}
## Recommendation 2: Health Insurance Enrollment Outreach in Maryvale / West Phoenix

The west Phoenix and Maryvale hot spot (see @fig-lisa) shows particularly elevated rates of uninsurance, a component that contributes disproportionately to the hardship index in this area. Given that Arizona expanded Medicaid eligibility under the ACA, the persistence of high uninsurance rates suggests an enrollment gap rather than an eligibility gap. We recommend a targeted door-to-door enrollment campaign in the block groups identified as High-High clusters in the LISA analysis, partnering with community health centers and bilingual navigators. The state-level map (@fig-az-tracts) confirms that this is a statewide issue in border and urban communities, but the block-group-level LISA analysis pinpoints exactly where outreach resources should be concentrated within Maricopa County.
:::

::: {.callout-tip}
## Recommendation 3: Preserve and Expand Affordable Housing in Central Mesa

The central Mesa hot spot is smaller than south Phoenix but shows a concerning trend in the temporal analysis: several block groups at its edges shifted from not-significant to higher hardship between 2015 and 2023 (visible in @fig-change), suggesting the cluster may be expanding. We recommend that the county and City of Mesa collaborate on an affordable housing preservation strategy for this area, including acquisition of naturally occurring affordable housing (NOAH) properties before they are demolished or converted, investment in housing rehabilitation, and zoning reforms that allow gentle density increases near the Mesa light rail stations. The goal is to prevent displacement of existing residents while improving neighborhood conditions.
:::


# ── Step 6: Multi-Scale Summary Figure ───────────────

```{r}
#| label: fig-multiscale
#| fig-cap: "Three-scale comparison of economic hardship. Left: US counties provide national context. Center: Arizona tracts reveal within-state variation. Right: Maricopa block groups expose neighborhood-level concentration invisible at broader scales."
#| fig-width: 16
#| fig-height: 6

# ── Compact maps for side-by-side comparison ─────────────────
# This replicates the three-scale comparison from Lab 1 but with
# the updated 2023 data used throughout this policy brief.
p1 <- ggplot( county_complete ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis( option = "inferno", direction = -1, name = "Index" ) +
  theme_void() +
  labs( title = "US Counties" )

p2 <- ggplot( az_complete ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis( option = "inferno", direction = -1, name = "Index" ) +
  theme_void() +
  labs( title = "AZ Tracts" )

p3 <- ggplot( cbg_complete ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis( option = "inferno", direction = -1, name = "Index" ) +
  theme_void() +
  labs( title = "Maricopa CBGs" )

p1 + p2 + p3 +
  plot_annotation(
    title   = "Economic Hardship Across Three Geographic Scales",
    caption = "ACS 5-Year (2019-2023) | Higher values = greater hardship"
  )
```


# ── Questions ────────────────────────────────────────

## Q1 --- Interpret {#sec-q1}

::: {.callout-important}
## Question 1

What are the top 3 findings from your multi-scale analysis? For each, explain what geographic scale revealed it and why that scale was necessary.
:::

**Sample Answer:**

1. **Arizona is moderate nationally, but hides severe internal inequality.** The national county-level map (@fig-national) shows Arizona's counties in the middle range of US hardship. This finding required the national scale because without the full distribution of roughly 3,200 counties, there is no benchmark for "how bad is Arizona?" However, this scale cannot reveal what happens within counties, which leads to finding #2.

2. **Hardship in Maricopa County is not diffuse --- it is spatially concentrated in specific corridors.** The LISA hot spot analysis at the block group level (@fig-lisa) identifies three statistically significant clusters of high hardship. This finding required the block group scale because census tracts, while finer than counties, still smooth over the sharp neighborhood boundaries that separate high-hardship from low-hardship areas. The LISA method adds statistical rigor: instead of just eyeballing a choropleth, we can confirm that these clusters are unlikely to occur by chance.

3. **The core hardship clusters are persistent over time.** The temporal change analysis (@fig-change) shows that the hardship hot spots identified in 2023 were also present in 2015. This finding required two things: the block group scale (to see neighborhood-level patterns) and the temporal comparison (to determine whether clusters are growing, shrinking, or stable). A single time period would identify the clusters but could not distinguish between a neighborhood experiencing a temporary downturn and one trapped in long-term disadvantage.


## Q2 --- Modify (Full Text) {#sec-q2}

::: {.callout-important}
## Question 2

Rewrite the executive summary from @sec-executive-summary for a DIFFERENT audience --- a **community nonprofit** instead of county government. Write the complete text below.
:::

**Original executive summary** (written for county government, reproduced from @sec-executive-summary):

> Economic hardship in Maricopa County is not evenly distributed --- it concentrates in specific neighborhoods that can be identified through multi-scale spatial analysis. At the national level, Arizona's counties fall in the moderate range of economic hardship relative to the rest of the United States, but this average masks sharp within-state inequality. Within Maricopa County, LISA hot spot analysis at the block group level reveals statistically significant clusters of high hardship in south Phoenix, central Mesa, and portions of west Phoenix. Temporal analysis comparing 2015 and 2023 ACS estimates shows that while some areas have seen modest improvement, the core hardship clusters have remained remarkably stable over time. These findings point to the need for geographically targeted interventions in persistent hardship zones, with particular attention to unemployment and lack of health insurance as the hardship components showing the least improvement.

::: {.callout-note}
## Rewritten Executive Summary --- For a Community Nonprofit

The neighborhoods your organization serves in south Phoenix, west Phoenix, and central Mesa face some of the highest levels of economic hardship in Maricopa County --- and our analysis shows these challenges are not going away on their own. Using Census data, we built a hardship score based on five factors that affect families directly: poverty, unemployment, education, health insurance, and income. When we mapped this score at the neighborhood level and tested for geographic patterns, three areas stood out as places where hardship is deeply concentrated and reinforced by surrounding neighborhoods.

Comparing data from 2015 and 2023 confirmed that these same neighborhoods have been struggling for at least eight years. While some areas on the edges of the metro saw improvement, the communities at the core of each cluster showed little positive change. The two most urgent needs are job access and health insurance enrollment --- these are the hardship factors that showed the least improvement over time.

These findings can support your grant applications by providing data-backed evidence of persistent need. They can help you make the case for expanded programming to funders who want to see geographic targeting. And they can guide practical decisions about where to open new service locations or deploy outreach workers --- specifically, the block groups identified as "hot spots" in the maps above are the areas where concentrated services would reach the greatest density of families facing multiple, overlapping challenges.
:::


## Q3 --- Interpret Modified {#sec-q3}

::: {.callout-important}
## Question 3

How did changing the audience change what you emphasized? What does this reveal about the difference between analytical findings and policy communication?
:::

**Sample Answer:**

The two summaries contain the same core findings, but they differ in three important ways.

**First, the language changed.** The county government version uses technical terms that analysts expect --- "multi-scale spatial analysis," "LISA hot spot analysis," "statistically significant clusters." The nonprofit version replaces these with plain language: "mapped this score at the neighborhood level and tested for geographic patterns." The underlying analysis is identical, but the vocabulary signals different professional contexts. A county official who allocates funding needs to know the analysis is methodologically sound; a nonprofit director who serves families needs to know the findings are actionable.

**Second, the framing of implications changed.** The government version recommends "geographically targeted interventions" --- a policy instrument. The nonprofit version connects the same data to "grant applications," "expanded programming," and "where to open new service locations" --- operational decisions that a nonprofit actually makes. The same finding (persistent hardship clusters) becomes a different kind of resource depending on who reads it.

**Third, the emotional register changed.** The government version is deliberately neutral and evidence-focused, consistent with the expectation that government documents avoid advocacy. The nonprofit version acknowledges that these neighborhoods are "struggling" and that "challenges are not going away on their own" --- language that reflects the nonprofit's mission-driven identity without becoming sensationalized.

This exercise reveals that the gap between analysis and communication is not about dumbing things down. It is about understanding what decisions each audience makes and presenting the same evidence in the form most useful for those decisions. A well-constructed analysis can serve multiple audiences, but only if the analyst translates it deliberately rather than assuming one version fits all.


---

# Common Errors & Troubleshooting

::: {.callout-warning}
## Frequently Encountered Issues

**1. Quarto callout blocks not rendering**
Make sure you include the `:::` closing fence on its own line. The opening `:::{.callout-tip}` and closing `:::` must each be on their own line with no other text.

**2. Cross-references showing as literal text (@fig-xxx)**
Cross-references require two things: the chunk must have `#| label: fig-xxx` (with the `fig-` prefix), and the chunk must produce exactly one figure. If a chunk produces multiple plots, the cross-reference will not work.

**3. Tabset panels not rendering**
The `::: {.panel-tabset}` opening and `:::` closing must be on their own lines, and each tab heading must use the `###` level (three hashes). Higher or lower heading levels will break the tabset.

**4. `poly2nb()` fails with topology errors**
Always run `st_make_valid()` before building neighbor lists. Invalid geometries are common in Census shapefiles and will cause `poly2nb()` to crash or produce incorrect results.

**5. Temporal join loses many CBGs**
The 2015 ACS uses 2010 Census block group boundaries while the 2023 ACS uses 2020 boundaries. Some CBGs were split, merged, or renumbered between censuses. The `inner_join()` by GEOID drops any CBG whose identifier changed. This is expected and the code reports how many were matched.
:::
