---
title: "Lab 7 | PAF 516"
format:
  dashboard:
    theme:
      - darkly
      - ../../assets/dashboard-theme.scss
    orientation: columns
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
knitr::opts_chunk$set( echo = FALSE, message = FALSE, warning = FALSE )
options( tigris_use_cache = TRUE )
options( tigris_progress_bar = FALSE )
```

```{r}
#| label: packages
# ── Package Setup ──────────────────────────────────────────────────────
# R packages update frequently and can break working code. We have saved
# the exact package versions used to build this lab. Run the block below
# ONCE the first time you open this file, then comment it out.
if (!requireNamespace("renv", quietly = TRUE)) install.packages("renv")
download.file(
  url      = "https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab7/renv.lock",
  destfile = "renv.lock", mode = "wb"
)
renv::restore(lockfile = "renv.lock", prompt = FALSE)
# After running once, comment out the four lines above.
# ───────────────────────────────────────────────────────────────────────

# ── Packages ─────────────────────────────────────────
library( tidyverse )
library( tidycensus )
library( sf )
library( tigris )
library( spdep )
library( psych )
library( viridis )
library( mapgl )
library( DT )
library( crosstalk )
library( bslib )
```

```{r}
#| label: dark-theme
# ── Dark dashboard theme for ggplot2 static plots ──────────────
theme_dark_dash <- function() {
  theme_void(base_size = 12) +
  theme(
    plot.background   = element_rect(fill = "#0d1117", color = NA),
    panel.background  = element_rect(fill = "#0d1117", color = NA),
    plot.title        = element_text(color = "#e6edf3", size = 14, face = "bold"),
    plot.subtitle     = element_text(color = "#8b949e", size = 11),
    plot.caption      = element_text(color = "#8b949e", size = 9),
    legend.background = element_rect(fill = "#0d1117", color = NA),
    legend.text       = element_text(color = "#e6edf3"),
    legend.title      = element_text(color = "#8b949e")
  )
}
```

```{r}
#| label: api-key
# Uncomment and add your Census API key if not already installed
# census_api_key("YOUR KEY GOES HERE", install = TRUE)
```

```{r}
#| label: fallback-data
#| eval: true

# ── FALLBACK: Load pre-saved datasets if Census API is down ──────
# Uncomment the lines below if you cannot get a Census API key.
#
# county_complete <- readRDS("../data/county_complete.rds")
# cbg_complete    <- readRDS("../data/cbg_2023.rds")
# cbg_2015_raw    <- readRDS("../data/cbg_2015.rds")
#
# county_raw      <- readRDS(url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab7/data/county_complete.rds"))
# cbg_raw         <- readRDS(url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab7/data/cbg_2023.rds"))
# cbg_2015_raw    <- readRDS(url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab7/data/cbg_2015.rds"))
```

```{r}
#| label: define-vars

# ── Define the 5 hardship indicator variables ────────────────
hardship_vars <- c(
  pov_below50  = "C17002_002",   # Income-to-poverty ratio under .50
  pov_50to99   = "C17002_003",   # Income-to-poverty ratio .50 to .99
  pov_den      = "C17002_001",   # Poverty universe total
  unemp_num    = "B23025_005",
  unemp_den    = "B23025_002",
  no_hs_num    = "B15003_002",
  no_hs_den    = "B15003_001",
  no_ins_num   = "B27010_017",
  no_ins_den   = "B27010_001",
  median_income = "B19013_001"
)
```

```{r}
#| label: build-function

# ── Reusable function to build the hardship index ────────────
build_hardship_index <- function( df ) {
  wide <- df %>%
    select( -moe ) %>%
    pivot_wider( names_from = variable, values_from = estimate )

  wide <- wide %>%
    mutate(
      poverty_rate = ( pov_below50 + pov_50to99 ) / pov_den,
      unemp_rate   = unemp_num / unemp_den,
      no_hs_rate   = no_hs_num / no_hs_den,
      no_ins_rate  = no_ins_num / no_ins_den
    )

  # Filter to complete cases FIRST (keeping geometry intact)
  complete <- wide %>%
    filter( !is.na( poverty_rate ) & !is.na( unemp_rate ) &
            !is.na( no_hs_rate )   & !is.na( no_ins_rate ) &
            !is.na( median_income ) )

  if ( nrow( complete ) == 0 ) stop( "No complete cases — check variable codes" )

  # Z-score standardize (drop geometry temporarily for scale())
  nums <- complete %>%
    st_drop_geometry() %>%
    select( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income ) %>%
    mutate( across( everything(), ~ as.numeric( scale(.) ) ) )

  # Reverse-code median income
  nums$median_income <- -1 * nums$median_income

  # Composite index
  complete$hardship_index <- rowMeans( nums, na.rm = TRUE )

  return( complete )  # Returns sf object with geometry intact
}
```

```{r}
#| label: pull-county
#| results: hide

# ── Pull US county-level data ────────────────────────────────
county_raw <- get_acs(
  geography    = "county",
  variables    = hardship_vars,
  year         = 2023,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
```

```{r}
#| label: build-county-index
county_complete <- build_hardship_index( county_raw )
```

```{r}
#| label: pull-cbg
#| results: hide

# ── Pull Maricopa County block group data ────────────────────
cbg_raw <- get_acs(
  geography    = "block group",
  variables    = hardship_vars,
  state        = "AZ",
  county       = "Maricopa",
  year         = 2023,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
```

```{r}
#| label: build-cbg-index
cbg_complete <- build_hardship_index( cbg_raw )
```

```{r}
#| label: reliability

# ── Cronbach's alpha for index reliability ────────────────────
alpha_data <- cbg_complete |>
  st_drop_geometry() |>
  select( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, med_income = median_income ) |>
  mutate( med_income = -med_income ) |>
  drop_na()

alpha_result <- psych::alpha( alpha_data )
alpha_val    <- round( alpha_result$total$raw_alpha, 3 )
```

```{r}
#| label: lisa-analysis

# ── LISA Hot Spot Analysis ───────────────────────────────────
cbg_complete <- st_make_valid( cbg_complete )

nb <- poly2nb( cbg_complete, queen = TRUE )
lw <- nb2listw( nb, style = "W", zero.policy = TRUE )

lisa <- localmoran( cbg_complete$hardship_index, lw, zero.policy = TRUE )

cbg_complete$lisa_i <- lisa[, "Ii"]
cbg_complete$lisa_z <- lisa[, "Z.Ii"]
cbg_complete$lisa_p <- lisa[, "Pr(z != E(Ii))"]

cbg_complete$lag_hardship <- lag.listw( lw, cbg_complete$hardship_index,
                                         zero.policy = TRUE )
```

```{r}
#| label: lisa-classify

# ── Classify LISA clusters ───────────────────────────────────
cbg_complete <- cbg_complete %>%
  mutate(
    val_centered = hardship_index - mean( hardship_index, na.rm = TRUE ),
    lag_centered = lag_hardship   - mean( lag_hardship, na.rm = TRUE ),

    quadrant = case_when(
      val_centered > 0 & lag_centered > 0 ~ "High-High",
      val_centered < 0 & lag_centered < 0 ~ "Low-Low",
      val_centered > 0 & lag_centered < 0 ~ "High-Low",
      val_centered < 0 & lag_centered > 0 ~ "Low-High",
      TRUE ~ "Not Significant"
    ),

    lisa_cluster = if_else( lisa_p <= 0.05, quadrant, "Not Significant" )
  )

cbg_complete$lisa_cluster <- factor(
  cbg_complete$lisa_cluster,
  levels = c( "High-High", "Low-Low", "High-Low", "Low-High", "Not Significant" )
)
```

```{r}
#| label: temporal-pull
#| results: hide

# ── Pull 2015 data for temporal comparison ───────────────────
cbg_2015_raw <- get_acs(
  geography    = "block group",
  variables    = hardship_vars,
  state        = "AZ",
  county       = "Maricopa",
  year         = 2015,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
```

```{r}
#| label: temporal-build

# ── Build temporal change data ───────────────────────────────
cbg_2015_complete <- build_hardship_index( cbg_2015_raw )

p1 <- cbg_2015_complete %>%
  st_drop_geometry() %>%
  select( GEOID, hardship_2015 = hardship_index )

p2 <- cbg_complete %>%
  select( GEOID, hardship_2023 = hardship_index )

change_df <- inner_join( p2, p1, by = "GEOID" )
change_df$hardship_change <- change_df$hardship_2023 - change_df$hardship_2015
```

```{r}
#| label: compute-stats

# ── Precompute summary statistics for value boxes ────────────
n_cbg          <- nrow( cbg_complete )
n_hotspot      <- sum( cbg_complete$lisa_cluster == "High-High", na.rm = TRUE )
n_coldspot     <- sum( cbg_complete$lisa_cluster == "Low-Low", na.rm = TRUE )
mean_hardship  <- round( mean( cbg_complete$hardship_index, na.rm = TRUE ), 3 )
max_hardship   <- round( max( county_complete$hardship_index, na.rm = TRUE ), 2 )
n_county       <- nrow( county_complete )

n_matched      <- nrow( change_df )
n_worsened     <- sum( change_df$hardship_change > 0, na.rm = TRUE )
n_improved     <- sum( change_df$hardship_change < 0, na.rm = TRUE )
pct_worsened   <- round( 100 * n_worsened / n_matched, 1 )
pct_improved   <- round( 100 * n_improved / n_matched, 1 )
mean_change    <- round( mean( change_df$hardship_change, na.rm = TRUE ), 3 )

# Global Moran's I
moran_global <- moran.test( cbg_complete$hardship_index, lw, zero.policy = TRUE )
moran_i_val  <- round( moran_global$estimate["Moran I statistic"], 3 )
```


# National Context

## Column {width=65%}

### US County Hardship Map

```{r}
#| label: national-map
#| fig-width: 12
#| fig-height: 7

# ── National choropleth (static ggplot) ──────────────────────
# Rendered as an image to keep file size manageable for RPubs.
# shift_geometry() moves AK and HI to inset positions so they
# appear as proper insets rather than at true lat/lon.
county_shifted <- shift_geometry( county_complete )

ggplot( county_shifted ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis(
    option    = "inferno",
    direction = -1,
    name      = "Hardship\nIndex"
  ) +
  theme_dark_dash() +
  labs(
    title    = "Economic Hardship Index by US County",
    subtitle = "ACS 5-Year Estimates (2019-2023)",
    caption  = "Higher values = greater economic hardship"
  )
```

## Column {width=35%}

### Summary Statistics

```{r}
#| label: national-vbox
#| content: valuebox
list(
  title = "US Counties Analyzed",
  value = scales::comma( n_county ),
  icon  = "globe-americas",
  color = "primary"
)
```

### Maricopa CBGs

```{r}
#| label: cbg-vbox
#| content: valuebox
list(
  title = "Maricopa Block Groups",
  value = scales::comma( n_cbg ),
  icon  = "geo-alt-fill",
  color = "info"
)
```

### Peak County Hardship

```{r}
#| label: max-vbox
#| content: valuebox
list(
  title = "Peak County Hardship",
  value = max_hardship,
  icon  = "exclamation-triangle-fill",
  color = "danger"
)
```

### About

Economic hardship concentrates in the rural South, Appalachia, border counties in South Texas, and tribal areas of the Southwest. Arizona's counties fall in the moderate range nationally, but this county-level average conceals sharp within-county inequality that becomes visible only at finer geographic scales.


# Local Patterns

## Column {width=60%}

### Interactive Map: Maricopa County Block Groups {.no-title}

```{r}
#| label: local-mapgl

# ── Prepare data for display ─────────────────────────────────
# Round values for tooltip display.
cbg_map <- cbg_complete %>%
  st_transform( 4326 ) %>%
  mutate(
    hardship_round = round( hardship_index, 3 ),
    poverty_pct    = round( poverty_rate * 100, 1 ),
    unemp_pct      = round( unemp_rate * 100, 1 ),
    no_hs_pct      = round( no_hs_rate * 100, 1 ),
    no_ins_pct     = round( no_ins_rate * 100, 1 ),
    income_fmt     = scales::dollar( median_income )
  )

# ── Create SharedData for crosstalk linking ──────────────────
# SharedData wraps the sf object so that DT datatable supports
# filtering and selection.  Note: mapgl's add_source() requires
# a plain sf object, so bidirectional map-to-table linking is
# not natively supported.  The DT table uses SharedData for its
# own selection/filter features; full bidirectional linking would
# require Observable JS or Shiny.
sd <- crosstalk::SharedData$new(
  cbg_map,
  key = ~GEOID,
  group = "cbg_link"
)

# ── Build mapgl interactive map using GeoJSON source ─────────
# Uses add_source() with type = "geojson" instead of PMTiles/serve_tiles()
# so the map is fully self-contained and works without a local tile server.
# add_source() expects an sf object (not SharedData).
maplibre( style = openfreemap_style( "dark" ),
          bounds = st_bbox( cbg_map ) ) |>
  add_source( id = "hardship-src",
              type = "geojson",
              data = cbg_map ) |>
  add_fill_layer( id = "hardship-fill",
                  source = "hardship-src",
                  fill_color = interpolate(
                    column = "hardship_round",
                    values = c( -1, 0, 1, 2 ),
                    stops  = c( "#1a9850", "#fee08b", "#d73027", "#67001f" )
                  ),
                  fill_opacity = 0.7,
                  hover_options = list( fill_opacity = 1 ),
                  tooltip = concat(
                    "GEOID: ", get_column( "GEOID" ),
                    " | Hardship: ", number_format( get_column( "hardship_round" ) ),
                    " | Poverty: ", get_column( "poverty_pct" ), "%",
                    " | Unemp: ", get_column( "unemp_pct" ), "%",
                    " | No HS: ", get_column( "no_hs_pct" ), "%",
                    " | Uninsured: ", get_column( "no_ins_pct" ), "%"
                  ) )
```

## Column {width=40%}

### Block Group Details (linked to map)

```{r}
#| label: local-table

# ── DT datatable linked via crosstalk ────────────────────────
# Selecting rows here highlights the corresponding polygon on the
# map, and clicking a polygon on the map selects the table row.
DT::datatable(
  sd,
  extensions = "Scroller",
  options    = list(
    scrollY    = 400,
    scroller   = TRUE,
    columnDefs = list(
      list( visible = FALSE, targets = c( 0 ) )
    )
  ),
  colnames   = c(
    "GEOID", "Name",
    "Hardship", "Poverty %", "Unemp %",
    "No HS %", "Uninsured %", "Income"
  ),
  selection  = "multiple",
  rownames   = FALSE
) %>%
  DT::formatRound( columns = c( "hardship_round" ), digits = 3 ) %>%
  DT::formatRound( columns = c( "poverty_pct", "unemp_pct", "no_hs_pct", "no_ins_pct" ), digits = 1 )
```

### Index Reliability

```{r}
#| label: alpha-vbox
#| content: valuebox
list(
  title = "Cronbach's Alpha",
  value = alpha_val,
  icon  = "check-circle-fill",
  color = if ( alpha_val >= 0.7 ) "success" else "warning"
)
```


# Hot Spots

## Column {width=60%}

### LISA Cluster Map

```{r}
#| label: lisa-map
#| fig-width: 10
#| fig-height: 9

# ── LISA cluster map (static ggplot) ─────────────────────────
lisa_colors <- c(
  "High-High"       = "#d73027",
  "Low-Low"         = "#4575b4",
  "High-Low"        = "#fc8d59",
  "Low-High"        = "#91bfdb",
  "Not Significant" = "#4a4a4a"
)

ggplot( cbg_complete ) +
  geom_sf( aes( fill = lisa_cluster ), color = NA ) +
  scale_fill_manual(
    values = lisa_colors,
    name   = "LISA Cluster"
  ) +
  theme_dark_dash() +
  labs(
    title    = "LISA Hot Spots: Economic Hardship",
    subtitle = "Maricopa County Block Groups | Queen contiguity, p < 0.05",
    caption  = "Red = High-High hot spots | Blue = Low-Low cold spots"
  )
```

## Column {width=40%}

### Key Statistics

```{r}
#| label: hotspot-vbox
#| content: valuebox
list(
  title = "Hot Spot CBGs (High-High)",
  value = n_hotspot,
  icon  = "fire",
  color = "danger"
)
```

### Cold Spot Count

```{r}
#| label: coldspot-vbox
#| content: valuebox
list(
  title = "Cold Spot CBGs (Low-Low)",
  value = n_coldspot,
  icon  = "snow2",
  color = "info"
)
```

### Moran's I

```{r}
#| label: moran-vbox
#| content: valuebox
list(
  title = "Global Moran's I",
  value = moran_i_val,
  icon  = "diagram-3-fill",
  color = "warning"
)
```

### Hot Spot Block Groups

```{r}
#| label: hotspot-table

# ── Table of HH and LL cluster CBGs ─────────────────────────
cluster_tbl <- cbg_complete %>%
  st_drop_geometry() %>%
  filter( lisa_cluster %in% c( "High-High", "Low-Low" ) ) %>%
  select(
    GEOID,
    Cluster       = lisa_cluster,
    Hardship      = hardship_index,
    `Poverty %`   = poverty_rate,
    `Unemp %`     = unemp_rate,
    `No HS %`     = no_hs_rate,
    `Uninsured %` = no_ins_rate
  ) %>%
  mutate(
    Hardship      = round( Hardship, 3 ),
    `Poverty %`   = round( `Poverty %` * 100, 1 ),
    `Unemp %`     = round( `Unemp %` * 100, 1 ),
    `No HS %`     = round( `No HS %` * 100, 1 ),
    `Uninsured %` = round( `Uninsured %` * 100, 1 )
  ) %>%
  arrange( desc( Hardship ) )

DT::datatable(
  cluster_tbl,
  options  = list(
    scrollY  = 300,
    scroller = TRUE,
    pageLength = 50
  ),
  rownames = FALSE,
  filter   = "top"
)
```


# Change Over Time

## Column {width=60%}

### Hardship Change (2015 to 2023)

```{r}
#| label: change-map
#| fig-width: 10
#| fig-height: 9

# ── Diverging change map ────────────────────────────────────
ggplot( change_df ) +
  geom_sf( aes( fill = hardship_change ), color = NA ) +
  scale_fill_gradient2(
    low      = "#4575b4",
    mid      = "#3d4450",
    high     = "#d73027",
    midpoint = 0,
    name     = "Change in\nHardship"
  ) +
  theme_dark_dash() +
  labs(
    title    = "Change in Economic Hardship (2015 to 2023)",
    subtitle = "Maricopa County Block Groups",
    caption  = "Red = hardship increased | Blue = hardship decreased"
  )
```

## Column {width=40%}

### CBGs Matched

```{r}
#| label: match-vbox
#| content: valuebox
list(
  title = "CBGs in Both Periods",
  value = scales::comma( n_matched ),
  icon  = "arrow-left-right",
  color = "secondary"
)
```

### Hardship Worsened

```{r}
#| label: worse-vbox
#| content: valuebox
list(
  title = "CBGs Where Hardship Increased",
  value = paste0( pct_worsened, "%" ),
  icon  = "arrow-up-circle-fill",
  color = "danger"
)
```

### Hardship Improved

```{r}
#| label: improve-vbox
#| content: valuebox
list(
  title = "CBGs Where Hardship Decreased",
  value = paste0( pct_improved, "%" ),
  icon  = "arrow-down-circle-fill",
  color = "success"
)
```

### Summary

Comparing 2011-2015 and 2019-2023 ACS estimates, most block groups experienced modest change. The core hardship clusters in south Phoenix, central Mesa, and west Phoenix show little improvement, confirming these are persistent hardship zones. Peripheral suburban areas show the most improvement, likely reflecting new development and rising incomes in growing communities.


# Policy

## Column {width=50%}

### Key Metrics

```{r}
#| label: policy-vbox1
#| content: valuebox
list(
  title = "Persistent Hot Spots",
  value = n_hotspot,
  icon  = "exclamation-octagon-fill",
  color = "danger"
)
```

### Mean Change

```{r}
#| label: policy-vbox2
#| content: valuebox
list(
  title = "Mean Hardship Change",
  value = mean_change,
  icon  = "graph-up",
  color = "warning"
)
```

### Priority Areas for Intervention

```{r}
#| label: priority-table

# ── Priority areas table ─────────────────────────────────────
# Identifies the top 20 highest-hardship hot spot CBGs as
# priority targets for policy intervention.
priority <- cbg_complete %>%
  st_drop_geometry() %>%
  filter( lisa_cluster == "High-High" ) %>%
  select(
    GEOID,
    Hardship      = hardship_index,
    `Poverty %`   = poverty_rate,
    `Unemp %`     = unemp_rate,
    `No HS %`     = no_hs_rate,
    `Uninsured %` = no_ins_rate,
    `Median Income` = median_income
  ) %>%
  mutate(
    Hardship        = round( Hardship, 3 ),
    `Poverty %`     = round( `Poverty %` * 100, 1 ),
    `Unemp %`       = round( `Unemp %` * 100, 1 ),
    `No HS %`       = round( `No HS %` * 100, 1 ),
    `Uninsured %`   = round( `Uninsured %` * 100, 1 ),
    `Median Income` = scales::dollar( `Median Income` )
  ) %>%
  arrange( desc( Hardship ) ) %>%
  head( 20 )

DT::datatable(
  priority,
  options  = list(
    scrollY    = 350,
    scroller   = TRUE,
    pageLength = 20
  ),
  rownames = FALSE,
  caption  = "Top 20 Highest-Hardship Hot Spot Block Groups"
)
```

## Column {width=50%}

### Recommendations

**Recommendation 1: [YOUR TITLE]**

[Write 3-5 sentences identifying a specific geographic area from the LISA analysis, describing the problem using data from the dashboard, and proposing a targeted policy intervention. Reference specific statistics.]

**Recommendation 2: [YOUR TITLE]**

[Write 3-5 sentences for a second recommendation targeting a different hot spot area or issue revealed by the temporal analysis.]

**Recommendation 3: [YOUR TITLE]**

[Write 3-5 sentences for a third recommendation. Consider what the change-over-time data reveals about emerging needs.]

### Data Sources & Methods

- **Data**: American Community Survey 5-Year Estimates via tidycensus
- **Periods**: 2011-2015 and 2019-2023
- **Geography**: Census block groups within Maricopa County, AZ
- **Index**: 5-variable composite (poverty, unemployment, education, insurance, income) using z-score standardization
- **Clustering**: Local Moran's I (LISA) with queen contiguity weights, p < 0.05
- **Tools**: R, Quarto Dashboard, crosstalk, mapgl, DT
