---
title: "Economic Hardship Dashboard — [YOUR NAME]"
format:
  dashboard:
    theme: readable
    orientation: columns
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
knitr::opts_chunk$set( echo = FALSE, message = FALSE, warning = FALSE )
options( tigris_use_cache = TRUE )
```

```{r}
#| label: packages
# ── Packages ─────────────────────────────────────────
# All packages needed for the dashboard. Install any you
# do not already have with install.packages("package_name").
library( tidyverse )
library( tidycensus )
library( sf )
library( spdep )
library( viridis )
library( mapgl )
library( freestiler )
library( DT )
library( crosstalk )
library( bslib )
```

```{r}
#| label: api-key
# Uncomment and add your Census API key if not already installed
# census_api_key("YOUR KEY GOES HERE", install = TRUE)
```

```{r}
#| label: define-vars

# ── Define the 5 hardship indicator variables ────────────────
# These are the SAME variables from Labs 1-6.
hardship_vars <- c(
  pov_num      = "B17001_002",
  pov_den      = "B17001_001",
  unemp_num    = "B23025_005",
  unemp_den    = "B23025_002",
  no_hs_num    = "B15003_002",
  no_hs_den    = "B15003_001",
  no_ins_num   = "B27010_017",
  no_ins_den   = "B27010_001",
  median_income = "B19013_001"
)
```

```{r}
#| label: build-function

# ── Reusable function to build the hardship index ────────────
# This is the same build_hardship_index() function from Labs 5-6.
# Copy it exactly — it works for any geography or time period.
build_hardship_index <- function( df ) {
  wide <- df %>%
    select( -moe ) %>%
    pivot_wider( names_from = variable, values_from = estimate )

  wide <- wide %>%
    mutate(
      poverty_rate = pov_num / pov_den,
      unemp_rate   = unemp_num / unemp_den,
      no_hs_rate   = no_hs_num / no_hs_den,
      no_ins_rate  = no_ins_num / no_ins_den
    )

  # Filter to complete cases FIRST (keeping geometry intact)
  complete <- wide %>%
    filter( !is.na( poverty_rate ) & !is.na( unemp_rate ) &
            !is.na( no_hs_rate )   & !is.na( no_ins_rate ) &
            !is.na( median_income ) )

  if ( nrow( complete ) == 0 ) stop( "No complete cases — check variable codes" )

  # Z-score standardize (drop geometry temporarily for scale())
  nums <- complete %>%
    st_drop_geometry() %>%
    select( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income ) %>%
    mutate( across( everything(), ~ as.numeric( scale(.) ) ) )

  # Reverse-code income
  nums$median_income <- -1 * nums$median_income

  # Composite index
  complete$hardship_index <- rowMeans( nums, na.rm = TRUE )

  return( complete )  # Returns sf object with geometry intact
}
```

```{r}
#| label: pull-data
#| results: hide

# TODO: Pull your data for all three scales.
# 1. US county data (for Tab 1 — National Context)
# 2. Maricopa CBG data for 2023 (for Tabs 2, 3, 5)
# 3. Maricopa CBG data for 2015 (for Tab 4 — Change Over Time)
#
# Example for county data:
# county_raw <- get_acs(
#   geography = "county",
#   variables = hardship_vars,
#   year      = 2023,
#   survey    = "acs5",
#   geometry  = TRUE
# )
#
# Example for Maricopa CBG data:
# cbg_raw <- get_acs(
#   geography = "block group",
#   variables = hardship_vars,
#   state     = "AZ",
#   county    = "Maricopa",
#   year      = 2023,
#   survey    = "acs5",
#   geometry  = TRUE
# )
```

```{r}
#| label: build-indices

# TODO: Build hardship indices using the function above.
# county_complete    <- build_hardship_index( county_raw )
# cbg_complete       <- build_hardship_index( cbg_raw )
# cbg_2015_complete  <- build_hardship_index( cbg_2015_raw )
```

```{r}
#| label: lisa-analysis

# TODO: Run the LISA hot spot analysis from Lab 4.
# 1. st_make_valid() on your CBG data
# 2. poly2nb() to build neighbor list
# 3. nb2listw() to create spatial weights
# 4. localmoran() to compute local Moran's I
# 5. Classify into HH/LL/HL/LH/NS using the centered-values method
# 6. Apply the p <= 0.05 significance filter
```

```{r}
#| label: temporal-change

# TODO: Build the temporal change dataset from Lab 5.
# 1. Extract GEOID + hardship_index from each period
# 2. inner_join() by GEOID
# 3. Compute hardship_change = hardship_2023 - hardship_2015
```

```{r}
#| label: compute-stats

# TODO: Precompute summary statistics for your value boxes.
# These variables will be referenced in the value box chunks below.
# Examples:
# n_cbg       <- nrow( cbg_complete )
# n_hotspot   <- sum( cbg_complete$lisa_cluster == "High-High", na.rm = TRUE )
# n_coldspot  <- sum( cbg_complete$lisa_cluster == "Low-Low", na.rm = TRUE )
# max_hardship <- round( max( cbg_complete$hardship_index, na.rm = TRUE ), 2 )
# n_worsened  <- sum( change_df$hardship_change > 0, na.rm = TRUE )
# pct_worsened <- round( 100 * n_worsened / nrow( change_df ), 1 )
```


# National Context

## Column {width=65%}

### US County Hardship Map

```{r}
#| label: national-map
#| fig-width: 12
#| fig-height: 7

# TODO: Create a ggplot choropleth of the national county hardship index.
# Use scale_fill_viridis(option = "inferno", direction = -1).
# Include labs() with title, subtitle, and caption.
```

## Column {width=35%}

### Summary Statistics

```{r}
#| label: national-vbox
#| content: valuebox

# TODO: Create a value box showing the number of counties analyzed.
# list(
#   title = "US Counties Analyzed",
#   value = scales::comma( n_county ),
#   icon  = "globe-americas",
#   color = "primary"
# )
```

### Block Groups

```{r}
#| label: cbg-vbox
#| content: valuebox

# TODO: Create a value box showing the number of local block groups.
# list(
#   title = "Maricopa Block Groups",
#   value = scales::comma( n_cbg ),
#   icon  = "geo-alt-fill",
#   color = "info"
# )
```

### About

<!-- TODO: Write 2-3 sentences placing your study area in national context. -->
<!-- What does the national map reveal? Where does your county fall? -->


# Local Patterns

## Column {width=60%}

### Interactive Map

```{r}
#| label: local-mapgl

# TODO: Create a crosstalk-linked mapgl map of your CBG hardship data.
#
# Steps:
# 1. Transform CBG data to WGS84: st_transform( cbg_complete, 4326 )
# 2. Round values for clean tooltip display
# 3. Wrap in SharedData: crosstalk::SharedData$new( data, key = ~GEOID )
# 4. Create PMTiles: freestile( data, "hardship.pmtiles", layer_name = "hardship" )
# 5. Serve tiles: serve_tiles( "hardship.pmtiles" )
# 6. Build map: maplibre( style = openfreemap_style("positron"), bounds = ... )
# 7. Add source: add_pmtiles_source( id = "hardship-src", url = ... )
# 8. Add fill layer: add_fill_layer( id = "hardship-fill", ... )
```

## Column {width=40%}

### Block Group Details (linked to map)

```{r}
#| label: local-table

# TODO: Create a DT::datatable using the SAME SharedData object
# from the mapgl map above. This links selections between
# the map and the table via crosstalk.
#
# DT::datatable(
#   sd,
#   options  = list( scrollY = 400, scroller = TRUE ),
#   rownames = FALSE
# )
```


# Hot Spots

## Column {width=60%}

### LISA Cluster Map

```{r}
#| label: lisa-map
#| fig-width: 10
#| fig-height: 9

# TODO: Create the LISA cluster map from Lab 4.
# Use scale_fill_manual() with the standard LISA colors:
#   "High-High" = "#d7191c", "Low-Low" = "#2c7bb6",
#   "High-Low" = "#fdae61", "Low-High" = "#abd9e9",
#   "Not Significant" = "#eeeeee"
```

## Column {width=40%}

### Hot Spot Count

```{r}
#| label: hotspot-vbox
#| content: valuebox

# TODO: Value box for the number of High-High hot spot CBGs.
# list(
#   title = "Hot Spot CBGs",
#   value = n_hotspot,
#   icon  = "fire",
#   color = "danger"
# )
```

### Cold Spot Count

```{r}
#| label: coldspot-vbox
#| content: valuebox

# TODO: Value box for the number of Low-Low cold spot CBGs.
```

### Cluster Table

```{r}
#| label: hotspot-table

# TODO: Create a DT::datatable filtered to only HH and LL clusters.
# Show GEOID, Cluster type, Hardship, and the 5 component rates.
# Sort by descending hardship.
```


# Change Over Time

## Column {width=60%}

### Hardship Change Map (2015 to 2023)

```{r}
#| label: change-map
#| fig-width: 10
#| fig-height: 9

# TODO: Create a diverging change map using scale_fill_gradient2().
# Blue = improvement (hardship decreased)
# White/yellow = no change
# Red = worsening (hardship increased)
# Set midpoint = 0.
```

## Column {width=40%}

### Hardship Worsened

```{r}
#| label: worse-vbox
#| content: valuebox

# TODO: Value box showing % of CBGs where hardship increased.
```

### Hardship Improved

```{r}
#| label: improve-vbox
#| content: valuebox

# TODO: Value box showing % of CBGs where hardship decreased.
```

### Summary

<!-- TODO: Write 2-3 sentences interpreting the temporal change patterns. -->
<!-- Are the hot spots from Tab 3 getting better, worse, or staying the same? -->


# Policy

## Column {width=50%}

### Key Metrics

```{r}
#| label: policy-vbox1
#| content: valuebox

# TODO: Value box for a key metric (e.g., number of persistent hot spots).
```

### Priority Areas for Intervention

```{r}
#| label: priority-table

# TODO: Create a table of the top 15-20 highest-hardship hot spot CBGs.
# These are the priority areas for policy intervention.
# Filter to lisa_cluster == "High-High", sort by hardship_index descending.
```

## Column {width=50%}

### Recommendations

<!-- TODO: Write 3 policy recommendations. Each should: -->
<!-- 1. Name a specific geographic area identified in your analysis -->
<!-- 2. Reference the evidence (which tab/analysis supports it) -->
<!-- 3. Propose a concrete, actionable intervention -->

<!-- **Recommendation 1: [Title]** -->
<!-- [Your recommendation text here] -->

<!-- **Recommendation 2: [Title]** -->
<!-- [Your recommendation text here] -->

<!-- **Recommendation 3: [Title]** -->
<!-- [Your recommendation text here] -->

### Data Sources & Methods

<!-- TODO: List your data sources, time periods, geography, index construction -->
<!-- method, clustering approach, and tools used. -->
