---
title: "Policy Brief | PAF 516"
subtitle: "Economic Hardship in [Your County]"
author: "[Your Name]"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-title: "Sections"
    self-contained: true
    code-fold: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
  fig-width: 10
---

::: {.callout-important}
## Before You Begin

1. **This policy brief accompanies your interactive dashboard.** Both deliverables use the **same custom economic hardship index** --- the variables you define in the `define-vars` chunk below should match the variables in your dashboard template.

2. **Customize this document:**
   - Replace `[Your Name]` in the YAML header with your name
   - Replace `[Your County]` in the subtitle with your chosen county
   - Update the `define-vars` chunk with your own ACS variable codes (matching your dashboard)
   - Update the `pull-data` chunk with your state and county

3. **Two static figures** are generated by the pre-filled code below. These are static versions of figures that also appear in your dashboard (LISA cluster map and top-10 hardship bar chart). The `code-fold: true` setting hides code by default so the document reads like a professional report.

4. **Your task:** Write the narrative sections marked `[YOUR ANSWER HERE]`. Each section tells you what to write and approximately how long your response should be. The quality of a policy brief depends on clear, audience-appropriate writing --- not code.

5. **Submit** both `Policy_Brief_YourLastName.qmd` and `Policy_Brief_YourLastName.html` to Canvas, along with your dashboard files.
:::


```{r}
#| label: packages
# ── One-time course environment setup ─────────────────────────────────────
if (!file.exists("~/.paf516_ready")) {
  if (!requireNamespace("renv", quietly = TRUE)) install.packages("renv")
  tmp <- tempfile(fileext = ".lock")
  download.file(
    url      = "https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/renv.lock",
    destfile = tmp, mode = "wb"
  )
  renv::restore(lockfile = tmp, prompt = FALSE)
  file.create("~/.paf516_ready")
}

library(tidyverse)
library(tidycensus)
library(sf)
library(spdep)
library(viridis)
library(patchwork)
```

```{r}
#| label: api-key
# Get a free Census API key at: https://api.census.gov/data/key_signup.html
# Run ONCE in RStudio console: census_api_key("YOUR_KEY_HERE", install = TRUE)
# Then restart R.
```

```{r}
#| label: define-vars

# ── Define your hardship indicator variables ─────────────────
# UPDATE these to match your dashboard template variables.
# The example below uses the same 5 variables as the course example.
hardship_vars <- c(
  pov_below50   = "C17002_002",
  pov_50to99    = "C17002_003",
  pov_den       = "C17002_001",
  unemp_num     = "B23025_005",
  unemp_den     = "B23025_002",
  no_hs_num     = "B15003_002",
  no_hs_den     = "B15003_001",
  no_ins_num    = "B27010_017",
  no_ins_den    = "B27010_001",
  median_income = "B19013_001"
)
```

```{r}
#| label: build-function

# ── Reusable function to build the economic hardship index ────
build_hardship_index <- function(df) {
  wide <- df %>%
    select(-moe) %>%
    pivot_wider(names_from = variable, values_from = estimate) %>%
    mutate(
      poverty_rate = (pov_below50 + pov_50to99) / pov_den,
      unemp_rate   = unemp_num / unemp_den,
      no_hs_rate   = no_hs_num / no_hs_den,
      no_ins_rate  = no_ins_num / no_ins_den
    )

  complete <- wide %>%
    filter(!is.na(poverty_rate), !is.na(unemp_rate),
           !is.na(no_hs_rate), !is.na(no_ins_rate),
           !is.na(median_income))

  nums <- complete %>%
    st_drop_geometry() %>%
    select(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income) %>%
    mutate(across(everything(), ~ as.numeric(scale(.))))

  nums$median_income <- -1 * nums$median_income
  complete$hardship_index <- rowMeans(nums, na.rm = TRUE)

  return(complete)
}
```

```{r}
#| label: pull-data
#| results: hide

# ── Pull census tract data for your county ───────────────────
# UPDATE state and county to match your dashboard.
tract_raw <- get_acs(
  geography    = "tract",
  variables    = hardship_vars,
  state        = "AZ",
  county       = "Maricopa",
  year         = 2023,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
```

```{r}
#| label: build-index

tract_complete <- build_hardship_index(tract_raw)

tract_complete <- tract_complete %>%
  mutate(
    tract_number = str_extract(NAME, "(?<=Census Tract )[0-9.]+")
  )

cat("Tracts with valid index:", nrow(tract_complete), "\n")
```

```{r}
#| label: lisa-analysis

# ── LISA hot spot analysis ───────────────────────────────────
tract_complete <- st_make_valid(tract_complete)
nb <- poly2nb(tract_complete, queen = TRUE)
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)

# Global Moran's I
moran_global <- moran.test(
  tract_complete$hardship_index, lw, zero.policy = TRUE
)

# Local Moran's I (LISA)
lisa <- localmoran(
  tract_complete$hardship_index, lw, zero.policy = TRUE
)

tract_complete$lisa_i <- lisa[, "Ii"]
tract_complete$lisa_p <- lisa[, "Pr(z != E(Ii))"]
tract_complete$lag_hardship <- lag.listw(
  lw, tract_complete$hardship_index, zero.policy = TRUE
)

# Classify clusters
tract_complete <- tract_complete %>%
  mutate(
    val_centered = hardship_index - mean(hardship_index, na.rm = TRUE),
    lag_centered = lag_hardship   - mean(lag_hardship, na.rm = TRUE),
    quadrant = case_when(
      val_centered > 0 & lag_centered > 0 ~ "High-High",
      val_centered < 0 & lag_centered < 0 ~ "Low-Low",
      val_centered > 0 & lag_centered < 0 ~ "High-Low",
      val_centered < 0 & lag_centered > 0 ~ "Low-High",
      TRUE ~ "Not Significant"
    ),
    lisa_cluster = if_else(lisa_p <= 0.05, quadrant, "Not Significant")
  )

tract_complete$lisa_cluster <- factor(
  tract_complete$lisa_cluster,
  levels = c("High-High", "Low-Low", "High-Low", "Low-High", "Not Significant")
)

n_hotspot  <- sum(tract_complete$lisa_cluster == "High-High", na.rm = TRUE)
n_coldspot <- sum(tract_complete$lisa_cluster == "Low-Low",   na.rm = TRUE)
```


# Executive Summary

::: {.callout-note}
## Executive Summary

[YOUR ANSWER HERE --- Write 1 paragraph (4--6 sentences) summarizing the most important findings from your spatial analysis and your primary policy recommendation. A busy decision-maker should be able to read only this paragraph and understand what your analysis found and what action you recommend. Lead with the finding, not the method.]
:::


# Background

[YOUR ANSWER HERE --- Write 2--3 paragraphs providing context for your analysis. Include: (1) why neighborhood-level economic hardship matters for policy, (2) a brief description of your study area and why it was chosen, and (3) what your composite index measures --- list the variables and explain what they capture. Do not describe R code or statistical methods --- write for a non-technical audience.]


# Key Findings

## LISA Hot Spot Analysis

```{r}
#| label: fig-lisa
#| fig-cap: "LISA cluster map identifying statistically significant hot spots (High-High) and cold spots (Low-Low) of economic hardship. Queen contiguity spatial weights, p < 0.05."
#| fig-height: 9

lisa_colors <- c(
  "High-High"       = "#d7191c",
  "Low-Low"         = "#2c7bb6",
  "High-Low"        = "#fdae61",
  "Low-High"        = "#abd9e9",
  "Not Significant" = "#eeeeee"
)

ggplot(tract_complete) +
  geom_sf(aes(fill = lisa_cluster), color = "white", linewidth = 0.05) +
  scale_fill_manual(values = lisa_colors, name = "LISA Cluster") +
  labs(
    title    = "LISA Hot Spots: Economic Hardship",
    subtitle = paste0(
      "Moran's I = ",
      round(moran_global$estimate["Moran I statistic"], 3),
      " | ", n_hotspot, " hot spot tracts, ",
      n_coldspot, " cold spot tracts"
    ),
    caption  = "Source: ACS 2019-2023 5-Year Estimates"
  ) +
  theme_void() +
  theme(legend.position = "right")
```

[YOUR ANSWER HERE --- Write 2--3 paragraphs interpreting @fig-lisa for a non-technical audience. Address the following:

- **Identify the clusters:** Where are the major hot spots (High-High areas)? Name specific geographic areas (e.g., "south Phoenix corridor," "central Mesa"). Where are the cold spots?
- **Translate the statistics:** What does the Global Moran's I value mean in plain language? (e.g., "Hardship is strongly concentrated in specific neighborhoods rather than spread evenly across the county.")
- **Policy significance:** Why does spatial clustering matter for policy? How should knowing *where* hardship concentrates change how resources are allocated?]


## Priority Hardship Areas

```{r}
#| label: fig-top10
#| fig-cap: "Top 10 highest-hardship census tracts within the study area, ranked by composite economic hardship index score."
#| fig-height: 5

top_tracts <- tract_complete %>%
  st_drop_geometry() %>%
  arrange(desc(hardship_index)) %>%
  slice_head(n = 10) %>%
  mutate(
    label = paste0("Tract ", tract_number),
    label = fct_reorder(label, hardship_index)
  )

ggplot(top_tracts, aes(x = hardship_index, y = label)) +
  geom_col(fill = "#d7191c", alpha = 0.85) +
  geom_text(aes(label = round(hardship_index, 2)),
            hjust = -0.1, size = 3.5) +
  labs(
    title    = "Top 10 Highest-Hardship Census Tracts",
    subtitle = "ACS 2019-2023 5-Year Estimates",
    x        = "Economic Hardship Index",
    y        = NULL
  ) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank())
```

[YOUR ANSWER HERE --- Write 1--2 paragraphs discussing @fig-top10. What do the highest-hardship tracts have in common? Are they geographically concentrated in the same hot spot cluster, or are they scattered? How do their scores compare to the county average (which is approximately zero for a z-score composite)?]


# Recommendations

::: {.callout-tip}
## Recommendation 1

[YOUR ANSWER HERE --- Write 3--5 sentences. Follow this structure: (1) State the finding --- what did the LISA analysis reveal about this area? (2) Explain the implication --- why does this matter for residents? (3) Propose the action --- what specifically should be done, by whom, and where? Name a responsible entity (e.g., city council, transit authority, health department) and a specific geographic target.]
:::

::: {.callout-tip}
## Recommendation 2

[YOUR ANSWER HERE --- Write 3--5 sentences for a second recommendation targeting a different hot spot area or hardship dimension (e.g., if Recommendation 1 addressed unemployment, this one might address health insurance coverage or educational attainment).]
:::

::: {.callout-tip}
## Recommendation 3

[YOUR ANSWER HERE --- Write 3--5 sentences for a third recommendation. Consider what spatial outliers (High-Low or Low-High areas) or cold spots might reveal about successful interventions that could be replicated in hot spot neighborhoods.]
:::


# Limitations

[YOUR ANSWER HERE --- Write 1 paragraph (3--5 sentences) acknowledging the limitations of your analysis. Consider:

- ACS data has margins of error, especially at the census tract level
- Your composite index captures only the dimensions you measured --- other factors (housing quality, environmental hazards, social cohesion) also matter
- Spatial patterns describe *where* hardship clusters, not *why* --- correlation is not causation
- The index weights all variables equally, which may not reflect local priorities]
