---
title: "Lab 3 Assignment | PAF 516"
subtitle: "Spatial Data Integration & Environmental Justice"
format:
  html:
    theme: readable
    highlight-style: tango
    toc: true
    toc-title: "Steps & Questions"
    self-contained: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
  fig-width: 10
---

::: {.callout-important}
## Before you begin
1. **This file contains the same pre-filled code as the tutorial.** Steps 1–9 below produce the same output you saw in the tutorial. If this file is in the same folder as your tutorial QMD, you can knit immediately (Cmd+Shift+K / Ctrl+Shift+K) and see the same results. If you placed it in a different location, run the `packages` chunk first (Cmd+Return / Ctrl+Enter), then knit.
3. **The assignment questions are at the top of this document.** Here is the workflow:
   - **Q1** asks you to interpret results from the pre-filled steps. Knit first, then review the ICC output (Step 3), the hardship + facility map (Step 7), and the distance-decay charts (Steps 8–9), then write your answers.
   - **Q2** asks you to re-run the distance-decay analysis with finer distance bands by adapting the existing code. Fill in the scaffold chunks labeled `q2-...` at the end of this document, then re-knit.
   - **Q3** asks you to connect your buffer sensitivity findings to the Uncertain Geographic Context Problem (UGCoP) and make a defended policy recommendation.
4. Submit both `Lab3_Assignment_YourLastName.qmd` and `Lab3_Assignment_YourLastName.html` to Canvas.
:::

# Assignment Questions

## Q1 — Interpret the ICC and Distance-Decay Pattern {.question}

> **Text answer — no code required.** Knit the file first, then review the Step 3 ICC output and the Step 8–9 charts before writing your answers.

- **ICC:** What is the ICC value? Using the interpretation guide in Step 3 (>0.75 / 0.40–0.75 / <0.40), what does it suggest about aggregating to tract level? Is block group geography justified for this analysis?

[YOUR ANSWER HERE]

- **Distance-decay:** Describe the pattern in Step 8. Does mean hardship decrease as distance from facilities increases? Is the pattern strong or weak? Are there any unexpected reversals?

[YOUR ANSWER HERE]

- **EJ interpretation:** Do the maps and distance-decay chart support the environmental justice hypothesis — that lower-income communities bear disproportionate proximity to environmental hazards? Cite specific numbers from the decay table in your answer.

[YOUR ANSWER HERE]


## Q2 — Modify the Distance Bands {.question}

> **Code + output required.** Re-run the distance-decay analysis (Step 8) with **finer distance bands**:
> `< 0.25 mi`, `0.25–0.5 mi`, `0.5–1 mi`, `1–2 mi`, `> 2 mi`
>
> Fill in the scaffold chunks at the end of this document to re-generate the decay summary table and bar charts with the new bands.

::: {.callout-tip}
**Hint:** Change the `breaks` in `cut()` to `c(0, 0.25, 0.5, 1.0, 2.0, Inf)` and update the `labels` accordingly.
:::


## Q3 — Interpret Buffer Sensitivity (UGCoP) {.question}

> **Text answer — no code required.**

- **Decay comparison:** How does the finer-band pattern (Q2) compare to the original 5-band version (Step 8)? Did breaking `< 0.5 mi` into two sub-bands reveal anything new about the very nearest block groups?

[YOUR ANSWER HERE]

- **UGCoP:** Look at the multi-buffer comparison table from Step 9. Is the gap between inside and outside buffer consistent across the three distances (0.5, 1.0, 2.0 miles), or does it shrink/change at one threshold? What does this tell you about the Uncertain Geographic Context Problem?

[YOUR ANSWER HERE]

- **Policy recommendation:** If you were advising a city agency on which block groups to prioritize for environmental remediation funding, what buffer distance would you recommend — and why? Drawing on Kwan (2012) and Maantay (2002), how would you defend that choice?

[YOUR ANSWER HERE]


---



```{r}
#| label: packages
if (!file.exists(".renv_done")) {
  if (!requireNamespace("renv", quietly = TRUE)) install.packages("renv")
  download.file(
    url      = "https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab3/renv.lock",
    destfile = "renv.lock", mode = "wb"
  )
  renv::restore(lockfile = "renv.lock", prompt = FALSE)
  file.create(".renv_done")
}

library(tidyverse)
library(tidycensus)
library(sf)
library(tigris)
library(psych)
library(patchwork)
library(viridis)
```

```{r}
#| label: api-key
# Get a free Census API key at: https://api.census.gov/data/key_signup.html
# Run ONCE in RStudio console: census_api_key("YOUR_KEY_HERE", install = TRUE)
# Then restart R.
```


# Step 1: Pull Maricopa County Block Group Data

```{r}
#| label: step.01-vars

hardship_vars <- c(
  pov_below50    = "C17002_002",  pov_50to99  = "C17002_003",
  pov_den        = "C17002_001",  median_income = "B19013_001",
  pct_unemp_num  = "B23025_005",  pct_unemp_den = "B23025_002",
  pct_no_hs_num  = "B15003_002",  pct_no_hs_den = "B15003_001",
  pct_no_ins_num = "B27010_017",  pct_no_ins_den= "B27010_001"
)
```

```{r}
#| label: step.01-pull
#| results: hide

cbg_raw <- get_acs(
  geography = "block group", variables = hardship_vars,
  state = "AZ", county = "Maricopa",
  year = 2023, survey = "acs5", geometry = TRUE, progress_bar = FALSE
)
```


# Step 2: Build the 5-Variable Hardship Index

```{r}
#| label: step.02-wrangle

cbg_wide <- cbg_raw %>%
  select(-moe) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(
    poverty_rate = (pov_below50 + pov_50to99) / pov_den,
    unemp_rate   = pct_unemp_num / pct_unemp_den,
    no_hs_rate   = pct_no_hs_num / pct_no_hs_den,
    no_ins_rate  = pct_no_ins_num / pct_no_ins_den
  )

cbg_complete <- cbg_wide %>%
  drop_na(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income)

index_vars <- cbg_complete %>%
  st_drop_geometry() %>%
  select(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income)

index_z <- index_vars %>%
  mutate(across(everything(), ~ as.numeric(scale(.)))) %>%
  mutate(median_income = -1 * median_income)

index_z$hardship_index <- rowMeans(index_z[, 1:5], na.rm = TRUE)
cbg_complete$hardship_index <- index_z$hardship_index

cat("CBGs with valid index:", nrow(cbg_complete), "\n")
```


# Step 3: Aggregation Sensitivity — ICC Test

Does geography choice matter? The ICC measures how much within-tract variation we would lose by aggregating block groups to tracts.

```{r}
#| label: step.03-icc

cbg_complete <- cbg_complete |>
  mutate(
    tract_id = str_sub(GEOID, 1, 11),
    area     = st_area(geometry) |> as.numeric()
  )

# ICC via one-way ANOVA — no heavy packages required
# Formula: (MS_between - MS_within) / (MS_between + (n0 - 1) * MS_within)
# where n0 is the effective average group size (adjusted for unequal groups)
aov_icc  <- aov(hardship_index ~ factor(tract_id), data = cbg_complete)
ms       <- summary(aov_icc)[[1]][, "Mean Sq"]
ms_b     <- ms[1]   # mean square between tracts
ms_w     <- ms[2]   # mean square within tracts (residual)

# Effective group size (Shrout & Fleiss, 1979) — accounts for unequal CBG counts per tract
k        <- length(unique(cbg_complete$tract_id))
n_total  <- nrow(cbg_complete)
n0       <- (n_total - sum(table(cbg_complete$tract_id)^2) / n_total) / (k - 1)

icc_val  <- round((ms_b - ms_w) / (ms_b + (n0 - 1) * ms_w), 3)
cat("ICC:", icc_val, "\n")
cat("(Between-tract MS:", round(ms_b, 3), "| Within-tract MS:", round(ms_w, 3), "| Avg group size:", round(n0, 1), ")\n")
```

```{r}
#| label: step.03-maps
#| fig-width: 14
#| fig-height: 6

tract_agg <- cbg_complete |>
  st_drop_geometry() |>
  group_by(tract_id) |>
  summarise(hardship_tract = weighted.mean(hardship_index, w = area, na.rm = TRUE),
            .groups = "drop")

tract_geo <- get_acs(
  geography = "tract", variables = "B01001_001",
  state = "AZ", county = "Maricopa", year = 2023,
  geometry = TRUE, progress_bar = FALSE
) |>
  select(GEOID, geometry) |>
  left_join(tract_agg, by = c("GEOID" = "tract_id"))

p_cbg   <- ggplot(cbg_complete) +
  geom_sf(aes(fill = hardship_index), color = NA) +
  scale_fill_viridis_c(option = "inferno", name = "Hardship\n(BG)") +
  theme_void() + labs(title = "Block Group Level")

p_tract <- ggplot(tract_geo) +
  geom_sf(aes(fill = hardship_tract), color = NA) +
  scale_fill_viridis_c(option = "inferno", name = "Hardship\n(Tract)") +
  theme_void() + labs(title = "Tract Level (Aggregated)")

p_cbg + p_tract +
  plot_annotation(title = paste0("Scale Sensitivity  |  ICC = ", icc_val))
```

::: {.callout-note}
## What the ICC tells us — and why it matters
The ICC is the fraction of **total hardship variance** that falls *between* tracts (as opposed to *within* them). It answers: "how much does it matter which tract a block group is in?"

- **ICC > 0.75**: tracts are very homogeneous — aggregating to tract loses little
- **ICC 0.40–0.75**: moderate within-tract variation — tract analysis is usable but imprecise  
- **ICC < 0.40**: most variation is within tracts — block group geography is strongly preferred

An ICC around 0.5 means roughly half the variance in block group hardship is *between* tracts and half is *within* them — a direct empirical demonstration of the MAUP concern from Module 2. Aggregating to tracts would suppress meaningful neighborhood-level variation.

**Extension for Lab 4:** Once you cover Moran's I, run it on the within-tract residuals (block group hardship minus tract mean) to test whether the within-tract variation is spatially structured. If the residuals cluster, MAUP is introducing systematic spatial bias — not just random noise.
:::


# Step 4: Load External Point Data (Environmental Facilities)

```{r}
#| label: step.04-facilities

facilities <- tibble(
  name = c("APS Ocotillo Power Plant","Intel Chandler Fab 32",
            "Honeywell Aerospace","Univar Solutions",
            "Goodyear Airport Industrial Park","Former Motorola 52nd Street Site",
            "19th Avenue Landfill","Arizona Plating Works",
            "Phoenix Sky Harbor Area (TRI cluster)","Mesa NW Industrial (TRI cluster)"),
  type = c("Power Generation","Semiconductor Mfg","Aerospace Mfg",
            "Chemical Distribution","Industrial/Airport","Superfund",
            "Superfund/Landfill","Superfund","TRI Cluster","TRI Cluster"),
  lat  = c(33.3919,33.2974,33.4352,33.4423,33.4350,
            33.4162,33.5070,33.4445,33.4372,33.4512),
  lon  = c(-111.8828,-111.8156,-112.0694,-111.9738,-112.3797,
            -111.9763,-112.1171,-112.0738,-112.0167,-111.8935)
)
cat("Facilities:", nrow(facilities), "\n")
```

::: {.callout-warning}
## Note on facility data
This is a **representative sample** of 10 major EPA-regulated and Superfund sites in Maricopa County — not a comprehensive listing. The full EPA TRI database contains hundreds of facilities. Using only 10 sites produces sparse inner buffers (roughly 14 block groups within 0.5 miles, ~59 within 1 mile), so treat the distance-decay pattern as illustrative of the method rather than a definitive EJ finding. For a real analysis, download the full TRI dataset from `https://www.epa.gov/toxics-release-inventory-tri-program`.
:::

```{r}
#| label: step.04-note
#| include: false
# Placeholder — note displayed above
```


# Step 5: Convert to sf and Align CRS

```{r}
#| label: step.05-crs

facilities_sf   <- st_as_sf(facilities, coords = c("lon", "lat"), crs = 4326)
facilities_proj <- st_transform(facilities_sf, crs = 2868)
cbg_proj        <- st_transform(cbg_complete, crs = 2868) |> st_make_valid()
cat("Both layers in EPSG:2868 (Arizona Central, US survey feet)\n")
```


# Step 6: Compute Distance to Nearest Facility

Rather than just asking "is this block group inside a buffer?" we compute the **exact distance from each block group's centroid to the nearest EPA facility**. This gives a continuous proximity measure and enables the distance-decay analysis in Step 7.

```{r}
#| label: step.06-distance

# Compute centroids of each block group
cbg_centroids <- st_centroid(cbg_proj)

# Distance matrix: every CBG centroid to every facility (feet)
dist_matrix <- st_distance(cbg_centroids, facilities_proj)

# Keep only the distance to the NEAREST facility
cbg_proj$dist_nearest_ft <- apply(dist_matrix, 1, min) |> as.numeric()
cbg_proj$dist_nearest_mi <- cbg_proj$dist_nearest_ft / 5280

cat("Median distance to nearest facility:", round(median(cbg_proj$dist_nearest_mi), 2), "miles\n")
cat("% of CBGs within 1 mile:", round(100 * mean(cbg_proj$dist_nearest_mi < 1), 1), "%\n")
cat("% of CBGs within 0.5 miles:", round(100 * mean(cbg_proj$dist_nearest_mi < 0.5), 1), "%\n")
```


# Step 7: Facility Density Map + Hardship Overlay

The compelling EJ visual: where are facilities located, and does hardship follow the same pattern?

```{r}
#| label: step.07-map
#| fig-width: 14
#| fig-height: 6

cbg_wgs84 <- st_transform(cbg_proj, 4326)
fac_wgs84 <- st_transform(facilities_proj, 4326)

# Hardship choropleth with facility locations overlaid
p_hardship_fac <- ggplot() +
  geom_sf(data = cbg_wgs84, aes(fill = hardship_index), color = NA) +
  scale_fill_viridis_c(option = "inferno", name = "Hardship\nIndex",
                       na.value = "grey80") +
  geom_sf(data = fac_wgs84, color = "white", fill = "#ffc627",
          shape = 21, size = 3.5, stroke = 1.2) +
  theme_void() +
  labs(
    title    = "Economic Hardship & EPA Facility Locations",
    subtitle = "Maricopa County Block Groups — Gold dots = EPA-regulated / Superfund facilities",
    caption  = "ACS 5-Year (2023) + EPA TRI / Superfund data"
  )

# Distance to nearest facility choropleth
p_distance <- ggplot(cbg_wgs84) +
  geom_sf(aes(fill = dist_nearest_mi), color = NA) +
  scale_fill_viridis_c(option = "plasma", direction = -1,
                       name = "Miles to\nNearest\nFacility",
                       na.value = "grey80") +
  geom_sf(data = fac_wgs84, color = "white", fill = "black",
          shape = 21, size = 3, stroke = 1) +
  theme_void() +
  labs(
    title    = "Distance to Nearest EPA Facility",
    subtitle = "Darker = closer to facility",
    caption  = "CRS: EPSG:2868 → displayed in WGS84"
  )

p_hardship_fac + p_distance +
  plot_annotation(
    title = "Do High-Hardship Block Groups Cluster Near EPA Facilities?"
  )
```


# Step 8: Distance-Decay Analysis — The EJ Finding

**The core environmental justice question:** Does hardship *decrease* as you move farther from EPA facilities? If yes, proximity to environmental burden compounds socioeconomic disadvantage — exactly what EJ research predicts.

```{r}
#| label: step.08-decay

# Create distance bands (the exposure zones you are comparing)
cbg_decay <- cbg_proj |>
  st_drop_geometry() |>
  mutate(
    dist_band = cut(
      dist_nearest_mi,
      breaks = c(0, 0.5, 1.0, 2.0, 5.0, Inf),
      labels = c("< 0.5 mi", "0.5–1 mi", "1–2 mi", "2–5 mi", "> 5 mi"),
      right  = TRUE, include.lowest = TRUE
    )
  )

# Compute mean hardship and poverty rate within each distance band
decay_summary <- cbg_decay |>
  group_by(dist_band) |>
  summarise(
    n_cbg           = n(),
    mean_hardship   = round(mean(hardship_index,   na.rm = TRUE), 3),
    mean_poverty    = round(mean(poverty_rate * 100, na.rm = TRUE), 1),
    mean_unemp      = round(mean(unemp_rate * 100,   na.rm = TRUE), 1),
    .groups = "drop"
  )

print(decay_summary)
```

```{r}
#| label: step.08-decay-plot
#| fig-width: 12
#| fig-height: 5

# Hardship distance-decay bar chart
p_decay_hardship <- ggplot(decay_summary, aes(x = dist_band, y = mean_hardship)) +
  geom_col(aes(fill = mean_hardship), color = "white") +
  scale_fill_viridis_c(option = "inferno", direction = 1, guide = "none") +
  geom_text(aes(label = round(mean_hardship, 2)),
            vjust = -0.4, fontface = "bold", size = 3.5) +
  theme_minimal(base_size = 12) +
  labs(
    title    = "Economic Hardship Decreases with Distance from EPA Facilities",
    subtitle = "Mean hardship index by distance band | Maricopa County Block Groups",
    x        = "Distance to Nearest EPA Facility",
    y        = "Mean Hardship Index",
    caption  = "Distance-decay pattern: UGCoP predicts this relationship changes with buffer threshold choice"
  )

# Poverty rate distance-decay
p_decay_poverty <- ggplot(decay_summary, aes(x = dist_band, y = mean_poverty)) +
  geom_col(aes(fill = mean_poverty), color = "white") +
  scale_fill_viridis_c(option = "rocket", direction = 1, guide = "none") +
  geom_text(aes(label = paste0(round(mean_poverty, 1), "%")),
            vjust = -0.4, fontface = "bold", size = 3.5) +
  theme_minimal(base_size = 12) +
  labs(
    title    = "Poverty Rate by Distance from EPA Facilities",
    subtitle = "Higher poverty near facilities — lower near clean suburbs",
    x        = "Distance to Nearest EPA Facility",
    y        = "Mean Poverty Rate (%)"
  )

p_decay_hardship + p_decay_poverty +
  plot_annotation(
    caption = "ACS 5-Year (2023) + EPA TRI data | Maricopa County Block Groups"
  )
```

**Reading the chart:** If hardship and poverty are highest in the "< 0.5 mi" bar and decrease across the distance bands, the data support the EJ hypothesis: **poorer, more disadvantaged communities are disproportionately located near EPA-regulated facilities**. Wealthier neighborhoods are farther away.


# Step 9: Multi-Buffer UGCoP Comparison

**The UGCoP question:** How sensitive is this finding to the buffer threshold? We run the analysis at three distances — 0.5, 1.0, and 2.0 miles — and compare results.

```{r}
#| label: step.09-ugcop

# Compute mean hardship for CBGs inside vs. outside each buffer distance
buffer_comparison <- tibble(
  buffer_mi = c(0.5, 1.0, 2.0)
) |>
  rowwise() |>
  mutate(
    inside  = mean(cbg_proj$hardship_index[cbg_proj$dist_nearest_mi <= buffer_mi], na.rm = TRUE),
    outside = mean(cbg_proj$hardship_index[cbg_proj$dist_nearest_mi >  buffer_mi], na.rm = TRUE),
    n_inside  = sum(cbg_proj$dist_nearest_mi <= buffer_mi, na.rm = TRUE),
    n_outside = sum(cbg_proj$dist_nearest_mi >  buffer_mi, na.rm = TRUE),
    difference = inside - outside
  ) |>
  ungroup() |>
  mutate(across(c(inside, outside, difference), ~ round(., 3)))

cat("=== UGCoP: How buffer distance changes the EJ finding ===\n\n")
print(buffer_comparison |>
        select(buffer_mi, n_inside, inside, outside, difference) |>
        rename(`Buffer (mi)` = buffer_mi,
               `N inside`   = n_inside,
               `Mean hardship inside` = inside,
               `Mean hardship outside`= outside,
               `Difference (in - out)` = difference))
```

```{r}
#| label: step.09-ugcop-plot
#| fig-width: 11
#| fig-height: 5

# Reshape for grouped bar chart
buf_long <- buffer_comparison |>
  select(buffer_mi, inside, outside) |>
  pivot_longer(cols = c(inside, outside),
               names_to  = "group",
               values_to = "mean_hardship") |>
  mutate(
    group = factor(group, levels = c("inside","outside"),
                   labels = c("Inside buffer","Outside buffer"))
  )

ggplot(buf_long,
       aes(x = factor(buffer_mi), y = mean_hardship, fill = group)) +
  geom_col(position = "dodge", color = "white", width = 0.65) +
  geom_text(aes(label = round(mean_hardship, 2)),
            position = position_dodge(width = 0.65),
            vjust = -0.4, size = 3.5, fontface = "bold") +
  scale_fill_manual(
    values = c("Inside buffer" = "#d73027", "Outside buffer" = "#4575b4"),
    name   = NULL
  ) +
  scale_x_discrete(labels = paste0(c(0.5, 1.0, 2.0), "-mile buffer")) +
  theme_minimal(base_size = 12) +
  labs(
    title    = "UGCoP: Mean Hardship Inside vs. Outside Buffer — Three Distances",
    subtitle = "Does the 'proximity penalty' shrink or grow as the buffer expands?",
    x        = "Buffer Distance",
    y        = "Mean Hardship Index",
    caption  = "Red = inside buffer | Blue = outside | ACS 2023 + EPA TRI data"
  )
```

**Reading the chart:** If the gap between inside and outside is consistent across all three buffer distances, the finding is **robust** — it doesn't depend on where you draw the line. If the gap disappears or reverses at one distance, the finding is **threshold-sensitive** and the UGCoP is producing a misleading result at that threshold. This is exactly what Kwan (2012) warned about.

---

# Scaffold Chunks

```{r}
#| label: q2-fine-bands

# YOUR CODE HERE
# Hint: copy the cbg_decay and decay_summary code from Step 8 above,
# then change the breaks in cut() to:
#   breaks = c(0, 0.25, 0.5, 1.0, 2.0, Inf)
#   labels = c("< 0.25 mi", "0.25–0.5 mi", "0.5–1 mi", "1–2 mi", "> 2 mi")
```

```{r}
#| label: q2-decay-plot
#| fig-width: 12
#| fig-height: 5

# YOUR CODE HERE — regenerate the hardship and poverty decay bar charts
# using your new fine-band decay_summary
```
