---
title: "Lab 3 Assignment | PAF 516"
subtitle: "Spatial Data Integration & Environmental Justice"
format:
  html:
    theme: readable
    highlight-style: tango
    toc: true
    toc-title: "Steps & Questions"
    self-contained: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
  fig-width: 10
---

::: {.callout-important}
## Before you begin
1. Rename this file: `Lab3_Assignment_YourLastName.qmd`
2. Open the **renamed copy** — never edit the original
3. Run the `packages` chunk first before knitting
4. Submit both your `.qmd` and rendered `.html` to Canvas
:::


```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10)
options(tigris_use_cache = TRUE)
options(tigris_progress_bar = FALSE)
```

```{r}
#| label: packages
# ── Package Setup (runs automatically on first knit) ─────────────────
# The if-guard ensures this only runs once — no need to comment out.
if (!file.exists("renv.lock")) {
  if (!requireNamespace("renv", quietly = TRUE)) install.packages("renv")
  download.file(
    url      = "https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab3/renv.lock",
    destfile = "renv.lock", mode = "wb"
  )
  renv::restore(lockfile = "renv.lock", prompt = FALSE)
}

library(tidyverse)
library(tidycensus)
library(sf)
library(tigris)
library(psych)
library(patchwork)
library(viridis)
library(lme4)
library(performance)
```

```{r}
#| label: api-key
# Get a free Census API key at: https://api.census.gov/data/key_signup.html
# After receiving your key by email, run ONCE in the RStudio console:
#   census_api_key("YOUR_KEY_HERE", install = TRUE)
# Then restart R: Session -> Restart R
```

```{r}
#| label: fallback-data
# Option A (local): cbg_raw <- readRDS("../data/cbg_hardship_maricopa.rds")
# Option B (GitHub):
# cbg_raw <- readRDS(url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab3/data/cbg_hardship_maricopa.rds"))
```


# Step 1: Pull Maricopa County Block Group Data

```{r}
#| label: step.01-vars

hardship_vars <- c(
  pov_below50    = "C17002_002",
  pov_50to99     = "C17002_003",
  pov_den        = "C17002_001",
  median_income  = "B19013_001",
  pct_unemp_num  = "B23025_005",
  pct_unemp_den  = "B23025_002",
  pct_no_hs_num  = "B15003_002",
  pct_no_hs_den  = "B15003_001",
  pct_no_ins_num = "B27010_017",
  pct_no_ins_den = "B27010_001"
)
```

```{r}
#| label: step.01-pull
#| results: hide

cbg_raw <- get_acs(
  geography    = "block group",
  variables    = hardship_vars,
  state        = "AZ",
  county       = "Maricopa",
  year         = 2023,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
```


# Step 2: Build the 5-Variable Hardship Index

```{r}
#| label: step.02-wrangle

cbg_wide <- cbg_raw %>%
  select(-moe) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(
    poverty_rate = (pov_below50 + pov_50to99) / pov_den,
    unemp_rate   = pct_unemp_num / pct_unemp_den,
    no_hs_rate   = pct_no_hs_num / pct_no_hs_den,
    no_ins_rate  = pct_no_ins_num / pct_no_ins_den
  )

cbg_complete <- cbg_wide %>%
  drop_na(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income)

index_vars <- cbg_complete %>%
  st_drop_geometry() %>%
  select(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income)

index_z <- index_vars %>%
  mutate(across(everything(), ~ as.numeric(scale(.)))) %>%
  mutate(median_income = -1 * median_income)

alpha_original <- psych::alpha(index_z)
cat("Original 5-variable alpha:", round(alpha_original$total$raw_alpha, 3), "\n")
```

```{r}
#| label: step.02-index

index_z$hardship_index <- rowMeans(index_z[, 1:5], na.rm = TRUE)
cbg_complete$hardship_index <- index_z$hardship_index
```


# Step 3: Aggregation Sensitivity — ICC Test

```{r}
#| label: step.03-icc

cbg_complete <- cbg_complete |>
  mutate(
    tract_id = str_sub(GEOID, 1, 11),
    area     = st_area(geometry) |> as.numeric()
  )

model_icc <- lmer(hardship_index ~ 1 + (1 | tract_id),
                  data = cbg_complete, REML = TRUE)
icc_result <- performance::icc(model_icc)
icc_val    <- round(icc_result$ICC_adjusted, 3)

cat("ICC:", icc_val, "\n")
```

```{r}
#| label: step.03-maps
#| fig-width: 14
#| fig-height: 6

tract_aggregated <- cbg_complete |>
  st_drop_geometry() |>
  group_by(tract_id) |>
  summarise(hardship_tract = weighted.mean(hardship_index, w = area, na.rm = TRUE),
            .groups = "drop")

tract_geo <- get_acs(
  geography = "tract", variables = "B01001_001",
  state = "AZ", county = "Maricopa", year = 2023,
  geometry = TRUE, progress_bar = FALSE
) |>
  select(GEOID, geometry) |>
  left_join(tract_aggregated, by = c("GEOID" = "tract_id"))

p_cbg   <- ggplot(cbg_complete) +
  geom_sf(aes(fill = hardship_index), color = NA) +
  scale_fill_viridis_c(option = "inferno", name = "Hardship\n(BG)") +
  theme_void() + labs(title = "Block Group Level")

p_tract <- ggplot(tract_geo) +
  geom_sf(aes(fill = hardship_tract), color = NA) +
  scale_fill_viridis_c(option = "inferno", name = "Hardship\n(Tract)") +
  theme_void() + labs(title = "Tract Level (Aggregated)")

p_cbg + p_tract +
  plot_annotation(title = paste0("Scale Sensitivity  |  ICC = ", icc_val))
```


# Step 4: Load External Point Data (Environmental Facilities)

```{r}
#| label: step.04-facilities

facilities <- tibble(
  name = c("APS Ocotillo Power Plant","Intel Chandler Fab 32",
            "Honeywell Aerospace","Univar Solutions",
            "Goodyear Airport Industrial Park","Former Motorola 52nd Street Site",
            "19th Avenue Landfill","Arizona Plating Works",
            "Phoenix Sky Harbor Area (TRI cluster)","Mesa NW Industrial (TRI cluster)"),
  type = c("Power Generation","Semiconductor Mfg","Aerospace Mfg",
            "Chemical Distribution","Industrial/Airport","Superfund",
            "Superfund/Landfill","Superfund","TRI Cluster","TRI Cluster"),
  lat  = c(33.3919,33.2974,33.4352,33.4423,33.4350,
            33.4162,33.5070,33.4445,33.4372,33.4512),
  lon  = c(-111.8828,-111.8156,-112.0694,-111.9738,-112.3797,
            -111.9763,-112.1171,-112.0738,-112.0167,-111.8935)
)
```


# Step 5: Convert to sf and Align CRS

```{r}
#| label: step.05-crs

facilities_sf   <- st_as_sf(facilities, coords = c("lon", "lat"), crs = 4326)
facilities_proj <- st_transform(facilities_sf, crs = 2868)
cbg_proj        <- st_transform(cbg_complete, crs = 2868) |> st_make_valid()
```


# Step 6: Point-in-Polygon Spatial Join

```{r}
#| label: step.06-join

cbg_with_fac <- st_join(cbg_proj, facilities_proj, join = st_intersects)

facility_counts <- cbg_with_fac |>
  st_drop_geometry() |>
  group_by(GEOID) |>
  summarise(facility_count = sum(!is.na(name)), .groups = "drop")

cbg_proj <- cbg_proj |>
  left_join(facility_counts, by = "GEOID") |>
  mutate(facility_count = replace_na(facility_count, 0))

cat("CBGs with at least one facility:", sum(cbg_proj$facility_count > 0), "\n")
```


# Step 7: Buffer Analysis (1-Mile Proximity Zone)

```{r}
#| label: step.07-buffer

buffer_dist_ft  <- 5280  # 1 mile in US survey feet
facility_buffers <- st_buffer(facilities_proj, dist = buffer_dist_ft)
buffer_union     <- st_union(facility_buffers)
cbg_proj$near_facility <- as.integer(
  lengths(st_intersects(cbg_proj, buffer_union)) > 0
)

cat("CBGs within 1-mile buffer:", sum(cbg_proj$near_facility), "\n")
```


# Step 8: Enriched Hardship Index

```{r}
#| label: step.08-enriched

spatial_vars <- cbg_proj |>
  st_drop_geometry() |>
  select(GEOID, facility_count, near_facility)

cbg_enriched_data <- cbg_proj |>
  st_drop_geometry() |>
  select(GEOID) |>
  bind_cols(index_z |> select(-hardship_index)) |>
  left_join(spatial_vars, by = "GEOID") |>
  mutate(
    facility_z = as.numeric(scale(facility_count)),
    near_fac_z = as.numeric(scale(near_facility))
  )

enriched_vars <- cbg_enriched_data |>
  select(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate,
         median_income, facility_z, near_fac_z)

alpha_enriched    <- psych::alpha(enriched_vars, check.keys = TRUE)
alpha_val_enrich  <- round(alpha_enriched$total$raw_alpha, 3)
cbg_proj$hardship_enriched <- rowMeans(enriched_vars, na.rm = TRUE)

cat("Original alpha:", round(alpha_original$total$raw_alpha, 3), "\n")
cat("Enriched alpha:", alpha_val_enrich, "\n")
```


# Step 9: Comparative Maps

```{r}
#| label: step.09-maps
#| fig-width: 16
#| fig-height: 6

cbg_map <- st_transform(cbg_proj, 4326)

p_orig   <- ggplot(cbg_map) +
  geom_sf(aes(fill = hardship_index),    color = NA) +
  scale_fill_viridis_c(option = "inferno", name = "Hardship\n(Original)") +
  theme_void() + labs(title = "Original 5-Variable Index")

p_enrich <- ggplot(cbg_map) +
  geom_sf(aes(fill = hardship_enriched), color = NA) +
  scale_fill_viridis_c(option = "inferno", name = "Hardship\n(Enriched)") +
  theme_void() + labs(title = "Enriched 7-Variable Index")

p_orig + p_enrich +
  plot_annotation(title = "Does Spatial Integration Change the Picture?",
                  caption = "ACS 2023 + EPA TRI | Maricopa County Block Groups")
```

---

# Questions

## Q1 — Interpret the ICC and Scale Sensitivity {.question}

> **Text answer — no code required.**

- **ICC value:** What is the ICC for the Maricopa County hardship index? Based on the interpretation table in Module 3 Section 3, does this suggest that aggregating to tract level is acceptable, or does it lose meaningful variation?

[YOUR ANSWER HERE]

- **Map comparison:** Looking at the block group vs. tract maps in Step 3, where do you see the biggest differences? Name a specific area of Maricopa County where the block group map reveals something the tract map hides.

[YOUR ANSWER HERE]

- **Policy implication:** If you were advising a city agency on where to target services, would you recommend using block group or tract geography for this analysis? Justify your answer using the ICC result.

[YOUR ANSWER HERE]


## Q2 — Modify: Change Buffer Distance to 0.5 Miles {.question}

> **Code + output required.** Change the buffer distance from 1 mile (5,280 feet) to **0.5 miles (2,640 feet)**. Recompute the enriched index and regenerate the comparative maps. Show complete code and output.

```{r}
#| label: q2-half-mile-buffer

# YOUR CODE HERE
# Hint: change buffer_dist_ft to 2640
# Re-run steps 7, 8, and 9 with the new buffer distance
```

```{r}
#| label: q2-enriched-index

# YOUR CODE HERE
```

```{r}
#| label: q2-maps
#| fig-width: 16
#| fig-height: 6

# YOUR CODE HERE
```


## Q3 — Interpret the Buffer Sensitivity {.question}

> **Text answer — no code required.**

- **Alpha comparison:** Did Cronbach's alpha improve or weaken when you changed from 1 mile to 0.5 miles? What does this tell you about how well the 0.5-mile proximity variable fits the hardship construct?

[YOUR ANSWER HERE]

- **Map comparison:** How sensitive are the results to buffer distance? Where do the 1-mile and 0.5-mile enriched index maps disagree most?

[YOUR ANSWER HERE]

- **MAUP connection:** Buffer distance sensitivity is a form of MAUP — results change depending on how you define the "unit" of proximity. Drawing on the Module 3 readings (Maantay, 2002; Mohai & Saha, 2015), what steps would you take to make your buffer distance choice more defensible in a published analysis?

[YOUR ANSWER HERE]
