---
title: "Lab 03 — Lab Guide: Spatial Data Integration & Overlay Analysis"
subtitle: "PAF 516 | M3 Assignment"
format:
  html:
    theme: readable
    highlight-style: tango
    toc: true
    self-contained: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
  fig-width: 10
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set( echo = TRUE, message=F, warning=F, fig.width=10 )
options( tigris_use_cache = TRUE )
```

```{r}
#| label: packages
# Core packages for this lab
library( tidyverse )
library( tidycensus )
library( sf )          # Spatial operations: st_join, st_buffer, st_transform
library( viridis )     # Color scales for maps
library( patchwork )   # Side-by-side plot composition
library( psych )       # Cronbach's alpha
```

```{r}
#| label: api-key
#| eval: false
# Uncomment and add your Census API key
# census_api_key("YOUR KEY GOES HERE")
```

```{r}
#| label: fallback-data
#| eval: false

# FALLBACK: Load pre-saved dataset if Census API is down
# If the Census API is unavailable or you cannot get an API key,
# uncomment ONE of the lines below to load pre-saved Maricopa
# County CBG data. This contains raw Census variables in long
# format — you will still build the index yourself.
#
# Local fallback:
# cbg_raw <- readRDS( "../data/cbg_hardship_maricopa.rds" )
#
# GitHub URL fallback:
# cbg_raw <- readRDS( url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab3/data/cbg_hardship_maricopa.rds") )
#
# If using fallback data, skip the get_acs() call and proceed
# directly to Step 2 (Build the Hardship Index).
```


# Step 1: Pull Maricopa County CBG Data

In Labs 1 and 2 we worked at the tract and county level. Here we move to **census block groups (CBGs)** — the finest geography for which ACS data are published — and pull the same five hardship variables from Lab 1.

```{r}
#| label: step.01-vars
#| echo: true
#| eval: false

# Define our 5-variable economic hardship indicator set (same as Lab 1)
hardship_vars <- c(
  pov_below50    = "C17002_002",   # Income-to-poverty ratio under .50
  pov_50to99     = "C17002_003",   # Income-to-poverty ratio .50 to .99
  pov_den        = "C17002_001",   # Total population (poverty universe)
  median_income  = "B19013_001",   # Median household income
  pct_unemp_num  = "B23025_005",   # Unemployed population
  pct_unemp_den  = "B23025_002",   # In labor force (denominator)
  pct_no_hs_num  = "B15003_002",   # No high school diploma
  pct_no_hs_den  = "B15003_001",   # Total population 25+
  pct_no_ins_num = "B27010_017",   # Without health insurance (18-34)
  pct_no_ins_den = "B27010_001"    # Total population (insurance universe)
)
```

```{r}
#| label: step.01-pull
#| echo: true
#| eval: false
#| results: hide

# Pull CBG-level data with geometry for Maricopa County
cbg_raw <- get_acs(
  geography = "block group",
  variables = hardship_vars,
  state     = "AZ",
  county    = "Maricopa",
  year      = 2023,
  survey    = "acs5",
  geometry  = TRUE
)

# Fallback — uncomment if Census API is unavailable:
# cbg_raw <- readRDS( "../data/cbg_hardship_maricopa.rds" )
# OR via URL:
# cbg_raw <- readRDS( url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab3/data/cbg_hardship_maricopa.rds") )
```

Examine the dimensions of the resulting data frame to verify it loaded correctly. The data is in long format, with one row per variable per CBG.


# Step 2: Build the 5-Variable Hardship Index

This mirrors Lab 1: pivot to wide format, compute rates, standardize to z-scores, reverse-code income, and average into a composite index.

```{r}
#| label: step.02-wrangle
#| echo: true
#| eval: false

# Pivot from long to wide
cbg_wide <- cbg_raw %>%
  select( -moe ) %>%
  pivot_wider( names_from = variable, values_from = estimate )

# Compute derived rate variables
cbg_wide <- cbg_wide %>%
  mutate(
    poverty_rate = ( pov_below50 + pov_50to99 ) / pov_den,
    unemp_rate   = pct_unemp_num / pct_unemp_den,
    no_hs_rate   = pct_no_hs_num / pct_no_hs_den,
    no_ins_rate  = pct_no_ins_num / pct_no_ins_den
  )

# Keep only complete cases for the index
cbg_complete <- cbg_wide %>%
  drop_na( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income )

# Extract numeric columns, standardize to z-scores
index_vars <- cbg_complete %>%
  st_drop_geometry() %>%
  select( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income )

index_z <- index_vars %>%
  mutate( across( everything(), ~ as.numeric( scale(.) ) ) )

# Reverse-code median income (high income = LOW hardship)
index_z <- index_z %>%
  mutate( median_income = -1 * median_income )

# Compute Cronbach's alpha on the original 5-variable index
alpha_original <- psych::alpha( index_z )
alpha_val_orig <- alpha_original$total$raw_alpha
cat( "Original 5-variable alpha:", round( alpha_val_orig, 3 ), "\n" )
```

```{r}
#| label: step.02-index
#| echo: true
#| eval: false

# Create the composite hardship index (row mean of z-scores)
index_z$hardship_index <- rowMeans( index_z[ , 1:5 ], na.rm = TRUE )

# Attach back to spatial data
cbg_complete$hardship_index <- index_z$hardship_index

cat( "CBGs with valid index:", nrow( cbg_complete ), "\n" )
summary( cbg_complete$hardship_index )
```


# Step 3: Load External Point Data (Environmental Facilities)

This is the **new skill** in Lab 3. We bring in data from outside the Census — specifically, locations of major environmental facilities in Maricopa County drawn from EPA's Toxic Release Inventory (TRI) and Superfund programs. These are real facilities with known latitude/longitude coordinates.

```{r}
#| label: step.03-facilities
#| echo: true
#| eval: false

# Create a tibble of known environmental facilities in Maricopa County
# Sources: EPA TRI Explorer, EPA Superfund site listings
facilities <- tibble(
  name = c(
    "Palo Verde Nuclear Generating Station",
    "Salt River Project - Santan Station",
    "APS West Phoenix Power Plant",
    "APS Ocotillo Power Plant",
    "Motorola 52nd Street (Superfund)",
    "Indian Bend Wash (Superfund)",
    "ENSCO/Maryvale (Superfund)",
    "Hassayampa Landfill (Superfund)",
    "Phoenix Goodyear Airport North (Superfund)",
    "Phoenix Goodyear Airport South (Superfund)",
    "19th Avenue Landfill",
    "SR Recycling (Superfund)",
    "Agua Fria Wastewater Treatment",
    "91st Avenue Wastewater Treatment",
    "Tres Rios Wastewater Treatment",
    "Arizona Chemical Company",
    "Freeport McMoRan - Chandler",
    "Southwest Gas - Tempe Facility",
    "Arizona Public Service - Redhawk",
    "Honeywell Aerospace - Phoenix",
    "Raytheon Missile Systems - Mesa",
    "Boeing Mesa",
    "Luke Air Force Base",
    "Williams Gateway (Phoenix-Mesa)",
    "Intel Chandler Campus"
  ),
  lat = c(
    33.388, 33.316, 33.455, 33.420,
    33.435, 33.490, 33.491, 33.377,
    33.420, 33.410, 33.485, 33.432,
    33.531, 33.399, 33.387, 33.448,
    33.270, 33.425, 33.505, 33.442,
    33.373, 33.380, 33.535, 33.306,
    33.340
  ),
  lon = c(
    -112.862, -111.725, -112.125, -111.918,
    -111.978, -111.930, -112.146, -112.600,
    -112.373, -112.378, -112.100, -112.080,
    -112.201, -112.275, -112.280, -112.065,
    -111.860, -111.940, -112.720, -112.070,
    -111.700, -111.720, -112.383, -111.655,
    -111.940
  ),
  type = c(
    "Power Plant", "Power Plant", "Power Plant", "Power Plant",
    "Superfund", "Superfund", "Superfund", "Superfund",
    "Superfund", "Superfund", "Landfill", "Superfund",
    "Wastewater", "Wastewater", "Wastewater", "Industrial",
    "Industrial", "Industrial", "Power Plant", "Industrial",
    "Industrial", "Industrial", "Military", "Military",
    "Industrial"
  )
)

cat( "Facility records loaded:", nrow( facilities ), "\n" )
```


# Step 4: Convert to sf and Align CRS

The geometry column is the key that allows us to **join** data from completely different sources. First, we convert our facility table to an `sf` point object. Then we project both layers to **EPSG:2868** (NAD83 / Arizona Central, units in feet) so that distance calculations are accurate.

```{r}
#| label: step.04-crs
#| echo: true
#| eval: false

# Convert facility tibble to sf point layer
# Coordinates are in WGS 84 (EPSG:4326 — standard lat/lon)
facilities_sf <- st_as_sf(
  facilities,
  coords = c( "lon", "lat" ),
  crs    = 4326
)

# Project BOTH layers to EPSG:2868 (Arizona Central, feet)
# This is required for accurate distance/buffer calculations
cbg_proj    <- st_transform( cbg_complete, crs = 2868 )
facil_proj  <- st_transform( facilities_sf, crs = 2868 )

# Validate geometries before spatial operations
cbg_proj <- st_make_valid( cbg_proj )

cat( "CBG CRS:", st_crs( cbg_proj )$input, "\n" )
cat( "Facility CRS:", st_crs( facil_proj )$input, "\n" )
```

**Why EPSG:2868?** Geographic coordinates (lat/lon) use degrees — computing a 1-mile buffer in degrees gives distorted ellipses that vary with latitude. A projected CRS uses linear units (feet for 2868), so `st_buffer( x, dist = 5280 )` gives a true 1-mile circle.


# Step 5: Point-in-Polygon Spatial Join

A **spatial join** works just like `left_join()` or `inner_join()`, except the key is geometry rather than a shared column name. `st_join()` matches each CBG polygon to any facility points that fall inside it.

```{r}
#| label: step.05-spatial-join
#| echo: true
#| eval: false

# Spatial join: attach facility info to CBGs that contain them
cbg_with_pts <- st_join( cbg_proj, facil_proj, join = st_contains )

# Count facilities per CBG
facility_counts <- cbg_with_pts %>%
  st_drop_geometry() %>%
  group_by( GEOID ) %>%
  summarise( facility_count = sum( !is.na( name ) ) )

# Merge counts back to the projected CBG layer
cbg_proj <- cbg_proj %>%
  left_join( facility_counts, by = "GEOID" ) %>%
  mutate( facility_count = replace_na( facility_count, 0 ) )

cat( "CBGs with at least 1 facility:",
     sum( cbg_proj$facility_count > 0 ), "\n" )
cat( "Total facilities matched:",
     sum( cbg_proj$facility_count ), "\n" )
```


# Step 6: Buffer Analysis (1-Mile Proximity Zone)

Not all impacts stop at a polygon boundary. Pollution, noise, and truck traffic extend outward. We draw a **1-mile buffer** (5280 feet in EPSG:2868) around every facility, then identify which CBGs intersect those buffers.

```{r}
#| label: step.06-buffer
#| echo: true
#| eval: false

# 1 mile = 5280 feet (EPSG:2868 units are US survey feet)
buffer_dist <- 5280

# Create buffer polygons around each facility
facil_buffers <- st_buffer( facil_proj, dist = buffer_dist )

# Dissolve overlapping buffers into a single zone
buffer_zone <- st_union( facil_buffers )

# Identify CBGs that intersect the buffer zone
cbg_proj$in_buffer <- as.integer(
  st_intersects( cbg_proj, buffer_zone, sparse = FALSE )[ , 1 ]
)

cat( "CBGs within 1-mile buffer zone:",
     sum( cbg_proj$in_buffer == 1 ), "of", nrow( cbg_proj ), "\n" )
```

```{r}
#| label: step.06-buffer-map
#| echo: true
#| eval: false

# Visualize the buffer zones with facilities and CBGs
ggplot() +
  geom_sf( data = cbg_proj, fill = "grey90", color = "grey70", linewidth = 0.1 ) +
  geom_sf( data = facil_buffers, fill = "#ffc627", alpha = 0.3, color = "#ffc627" ) +
  geom_sf( data = facil_proj, color = "#8c1d40", size = 2 ) +
  theme_void() +
  labs(
    title    = "Environmental Facility Locations and 1-Mile Buffer Zones",
    subtitle = "Maricopa County CBGs | 25 TRI / Superfund / Industrial Facilities",
    caption  = "Gold = 1-mile buffer zone | Maroon points = facility locations"
  )
```


# Step 7: Enriched Hardship Index

We now add two spatially-derived variables to the original 5-variable index:

- **facility_count**: number of environmental facilities within the CBG
- **in_buffer**: binary indicator (1/0) for whether the CBG falls within 1 mile of any facility

Both are standardized and added to the index. We then recompute Cronbach's alpha to assess whether the enriched index remains internally consistent.

```{r}
#| label: step.07-enriched
#| echo: true
#| eval: false

# Build the enriched index with 7 variables
enriched_vars <- cbg_proj %>%
  st_drop_geometry() %>%
  select( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate,
          median_income, facility_count, in_buffer )

enriched_z <- enriched_vars %>%
  mutate( across( everything(), ~ as.numeric( scale(.) ) ) )

# Reverse-code median income
enriched_z <- enriched_z %>%
  mutate( median_income = -1 * median_income )

# Compute Cronbach's alpha for the 7-variable enriched index
alpha_enriched <- psych::alpha( enriched_z )
alpha_val_enr  <- alpha_enriched$total$raw_alpha

cat( "Original 5-variable alpha:", round( alpha_val_orig, 3 ), "\n" )
cat( "Enriched 7-variable alpha:", round( alpha_val_enr, 3 ), "\n" )
```

```{r}
#| label: step.07-alpha-detail
#| echo: true
#| eval: false

# Print full alpha output to see item-level diagnostics
print( alpha_enriched )
```

```{r}
#| label: step.07-compute-enriched
#| echo: true
#| eval: false

# Create the enriched composite index
enriched_z$enriched_index <- rowMeans( enriched_z[ , 1:7 ], na.rm = TRUE )

# Attach to the spatial data
cbg_proj$enriched_index <- enriched_z$enriched_index

summary( cbg_proj$enriched_index )
```


# Step 8: Comparative Maps

Side-by-side comparison of the original census-only hardship index vs. the spatially enriched index. We use `patchwork` to compose the two maps.

```{r}
#| label: step.08-maps
#| echo: true
#| eval: false
#| fig-width: 14
#| fig-height: 7

# Map 1: Original 5-variable hardship index
p1 <- ggplot( cbg_proj ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis( option = "inferno", direction = -1,
                      name = "Hardship\nIndex" ) +
  theme_void() +
  labs(
    title    = "Original 5-Variable Hardship Index",
    subtitle = "Census-only (CBG level)",
    caption  = "Higher values = greater hardship"
  )

# Map 2: Enriched 7-variable index
p2 <- ggplot( cbg_proj ) +
  geom_sf( aes( fill = enriched_index ), color = NA ) +
  scale_fill_viridis( option = "inferno", direction = -1,
                      name = "Enriched\nIndex" ) +
  theme_void() +
  labs(
    title    = "Enriched 7-Variable Hardship Index",
    subtitle = "Census + facility proximity",
    caption  = "Includes facility count + 1-mi buffer"
  )

# Compose side by side
p1 + p2 +
  plot_annotation(
    title   = "Effect of Spatial Data Integration on the Hardship Index",
    subtitle = "Maricopa County Census Block Groups | ACS 2023",
    theme   = theme( plot.title = element_text( size = 16, face = "bold" ) )
  )
```

```{r}
#| label: step.08-difference
#| echo: true
#| eval: false

# Compute and map the difference: enriched minus original
cbg_proj$index_diff <- cbg_proj$enriched_index - cbg_proj$hardship_index

ggplot( cbg_proj ) +
  geom_sf( aes( fill = index_diff ), color = NA ) +
  scale_fill_gradient2(
    low      = "#2166ac",
    mid      = "white",
    high     = "#b2182b",
    midpoint = 0,
    name     = "Difference"
  ) +
  theme_void() +
  labs(
    title    = "Difference: Enriched Index minus Original Index",
    subtitle = "Red = spatial integration increased hardship score | Blue = decreased",
    caption  = "Maricopa County CBGs | ACS 2023"
  )
```


# Questions

### Question 1: Interpret the Comparative Maps

**Compare the original hardship index map to the enriched index map. Where do they disagree? What does spatial integration reveal that census-only analysis missed?**

> YOUR ANSWER HERE


### Question 2: Modify — Change Buffer Distance to 0.5 Miles

**Change the buffer distance from 1 mile to 0.5 miles. Recompute the enriched index and re-map. Show complete code and output.**

```{r}
#| label: q2-half-mile-buffer
#| echo: true
#| eval: false

# YOUR CODE HERE
```

```{r}
#| label: q2-half-mile-maps
#| echo: true
#| eval: false
#| fig-width: 14
#| fig-height: 7

# YOUR CODE HERE
```


### Question 3: Interpret the Modified Results (MAUP)

**How sensitive are the results to buffer distance? What does this tell you about the Modifiable Areal Unit Problem (MAUP)?**

> YOUR ANSWER HERE
