---
title: "Lab 1 Assignment | PAF 516"
subtitle: "Measurement, Indexing & Multi-Scale Mapping"
format:
  html:
    theme: readable
    highlight-style: tango
    toc: true
    toc-title: "Steps & Questions"
    self-contained: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
  fig-width: 10
---

::: {.callout-important}
## Before you begin
1. **This file contains the same pre-filled code as the tutorial.** Steps 1–8 below produce the same output you saw in the tutorial. If this file is in the same folder as your tutorial QMD, you can knit immediately (Cmd+Shift+K / Ctrl+Shift+K) and see the same results. If you placed it in a different location, run the `packages` chunk first (Cmd+Return / Ctrl+Enter), then knit.
3. **The assignment questions are at the top of this document.** Here is the workflow:
   - **Q1** asks you to interpret results from the pre-filled steps. Knit first, then review the Cronbach's alpha output and maps in Steps 3–8, then write your answers.
   - **Q2** asks you to extend the enriched index by adding SNAP benefits as an 11th variable by adapting the existing code. Fill in the scaffold chunks labeled `q2-...` at the end of this document, then re-knit.
   - **Q3** asks you to evaluate diminishing returns — does adding one more variable still shift the map meaningfully? Compare the 5→10 and 10→11 changes.
4. Submit both `Lab1_Assignment_YourLastName.qmd` and `Lab1_Assignment_YourLastName.html` to Canvas.
:::

# Assignment Questions

## Q1 — Interpret the Baseline and Enriched Results {.question}

> **Text answer — no code required.** Knit the file first, then review the outputs from Steps 3–8 and answer below.

- **Cronbach's alpha:** What is the 5-variable alpha? What is the 10-variable alpha? Did expanding the index improve internal consistency? What does this tell you about whether the new dimensions belong in the hardship construct?

[YOUR ANSWER HERE]

- **Map comparison (Step 8):** Looking at the 5-variable vs. 10-variable maps, where do they disagree most? Identify a specific area of Arizona where expanding measurement changes the apparent level of hardship.

[YOUR ANSWER HERE]

- **Difference map:** What does the red/blue pattern reveal? What types of communities — urban core, suburbs, rural, tribal areas — tend to score *higher* in the enriched index relative to the baseline? Why might that be?

[YOUR ANSWER HERE]


## Q2 — Add SNAP as an 11th Variable {.question}

> **Code + output required.**
>
> The 10-variable enriched index does not yet include SNAP (food assistance) receipt. Add **percent of households receiving SNAP benefits** as an 11th variable:
> - Numerator: `B19058_002` (households receiving SNAP)
> - Denominator: `B19058_001` (total households)
>
> Fill in the three scaffold chunks at the end of this document to:
> 1. Re-pull tract data with SNAP added
> 2. Recompute Cronbach's alpha with 11 variables and build the new index
> 3. Produce a side-by-side map (10-var vs. 11-var) **and** a difference map (11-var minus 10-var)


## Q3 — Evaluate Diminishing Returns {.question}

> **Text answer — no code required.**

- **Marginal alpha change:** How much did alpha change going from 5→10 variables? How much from 10→11 (adding SNAP)? Is there a pattern of diminishing returns?

[YOUR ANSWER HERE]

- **Marginal map change:** Compare the 5→10 difference map (Step 8) to your 10→11 difference map (Q2). Did adding SNAP produce as dramatic a spatial shift as expanding from 5 to 10 variables? Where, if anywhere, did adding SNAP still matter?

[YOUR ANSWER HERE]

- **Recommendation:** At what point should you stop adding variables to a composite index? Using evidence from your alpha comparisons and difference maps, make a recommendation for how many variables this economic hardship index should include — and justify it.

[YOUR ANSWER HERE]


---



```{r}
#| label: packages
if (!file.exists(".renv_done")) {
  if (!requireNamespace("renv", quietly = TRUE)) install.packages("renv")
  download.file(
    url      = "https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab1/renv.lock",
    destfile = "renv.lock", mode = "wb"
  )
  renv::restore(lockfile = "renv.lock", prompt = FALSE)
  file.create(".renv_done")
}

library(tidyverse)
library(tidycensus)
library(sf)
library(tigris)
library(psych)
library(corrplot)
library(viridis)
library(ggplot2)
library(patchwork)
```

```{r}
#| label: api-key
# Get a free Census API key at: https://api.census.gov/data/key_signup.html
# After receiving your key by email, run this line ONCE in the RStudio console:
# census_api_key("YOUR_KEY_HERE", install = TRUE)
# Then restart R: Session → Restart R
```


# Step 1: Retrieve Census Variables

We build an **Economic Hardship Index** from 5 computed rates at the Arizona tract level. Each rate requires a numerator/denominator pair from the ACS, so the variable list contains 10 census codes — but they collapse into 5 measures: poverty rate, unemployment rate, no high school diploma rate, no health insurance rate, and median income.

```{r}
#| label: step.01
#| results: hide

hardship_vars <- c(
  pov_below50    = "C17002_002",   # Income-to-poverty ratio under .50
  pov_50to99     = "C17002_003",   # Income-to-poverty ratio .50 to .99
  pov_den        = "C17002_001",   # Total population (poverty universe)
  median_income  = "B19013_001",   # Median household income
  pct_unemp_num  = "B23025_005",   # Unemployed population
  pct_unemp_den  = "B23025_002",   # In labor force
  pct_no_hs_num  = "B15003_002",   # Population with less than high school
  pct_no_hs_den  = "B15003_001",   # Total population 25+
  pct_no_ins_num = "B27010_017",   # Without health insurance (18–34)
  pct_no_ins_den = "B27010_001"    # Total population (insurance universe)
)

CenDF <- get_acs(
  geography    = "tract",
  variables    = hardship_vars,
  state        = "AZ",
  year         = 2023,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
```

**Variable justification:**

| Variable | Census Code | Why Included |
|----------|------------|--------------|
| Poverty rate | C17002_002 + _003 / _001 | Direct measure of economic deprivation |
| Unemployment rate | B23025_005 / B23025_002 | Labor market hardship |
| No high school diploma | B15003_002 / B15003_001 | Education predicts economic outcomes |
| No health insurance | B27010_017 / B27010_001 | Reflects economic vulnerability |
| Median household income | B19013_001 | Core income measure (reverse coded) |


# Step 2: Compute Rates and Standardize

Reshape from long to wide, compute percentage variables, and standardize to z-scores.

```{r}
#| label: step.02

CenDF_wide <- CenDF %>%
  select(-moe) %>%
  pivot_wider(names_from = variable, values_from = estimate)

CenDF_wide <- CenDF_wide %>%
  mutate(
    poverty_rate = (pov_below50 + pov_50to99) / pov_den,
    unemp_rate   = pct_unemp_num / pct_unemp_den,
    no_hs_rate   = pct_no_hs_num / pct_no_hs_den,
    no_ins_rate  = pct_no_ins_num / pct_no_ins_den
  )

# Select indicator variables and standardize
index_vars <- CenDF_wide %>%
  st_drop_geometry() %>%
  select(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income) %>%
  drop_na()

index_z <- index_vars %>%
  mutate(across(everything(), ~ as.numeric(scale(.)))) %>%
  mutate(median_income = -1 * median_income)  # reverse-code: high income = low hardship
```


# Step 3: Cronbach's Alpha

Assess whether the 5 items reliably measure the same underlying construct.

```{r}
#| label: step.03

alpha_result <- psych::alpha(index_z)
print(alpha_result)
```

```{r}
#| label: alpha-summary

alpha_value <- alpha_result$total$raw_alpha
cat("Cronbach's alpha:", round(alpha_value, 3), "\n")
```

**Interpreting the output:** The `raw_alpha` is our Cronbach's alpha. The "Reliability if an item is dropped" section shows what alpha would be without each variable — if dropping an item increases alpha, it may not fit the construct well.


# Step 4: Build the Index and Visualize

```{r}
#| label: step.04a

cor_matrix <- cor(index_z, use = "complete.obs")

corrplot(
  cor_matrix,
  method      = "color",
  type        = "upper",
  addCoef.col = "black",
  tl.col      = "black",
  tl.srt      = 45,
  title       = "Inter-Item Correlation Matrix: Economic Hardship Index",
  mar         = c(0, 0, 2, 0)
)
```

```{r}
#| label: step.04b

index_z$hardship_index <- rowMeans(index_z, na.rm = TRUE)

ggplot(index_z, aes(x = hardship_index)) +
  geom_histogram(bins = 40, fill = "#8c1d40", color = "white", alpha = 0.8) +
  geom_vline(xintercept = mean(index_z$hardship_index, na.rm = TRUE),
             linetype = "dashed", color = "#ffc627", linewidth = 1) +
  theme_minimal() +
  labs(
    title    = "Distribution of Economic Hardship Index",
    subtitle = "Arizona Census Tracts (ACS 2023)",
    x        = "Economic Hardship Index (z-score scale)",
    y        = "Number of Tracts",
    caption  = "Dashed line = mean"
  )
```

```{r}
#| label: step.04c

CenDF_complete <- CenDF_wide %>%
  drop_na(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income)
CenDF_complete$hardship_index <- index_z$hardship_index
```


# Step 5: National County-Level Map

Pull the same 5 variables at the **county level** for the entire US and map national hardship patterns.

```{r}
#| label: step.05-pull
#| results: hide

county_df <- get_acs(
  geography    = "county",
  variables    = hardship_vars,
  year         = 2023,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
```

```{r}
#| label: step.05-index

county_wide <- county_df %>%
  select(-moe) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(
    poverty_rate = (pov_below50 + pov_50to99) / pov_den,
    unemp_rate   = pct_unemp_num / pct_unemp_den,
    no_hs_rate   = pct_no_hs_num / pct_no_hs_den,
    no_ins_rate  = pct_no_ins_num / pct_no_ins_den
  )

county_index <- county_wide %>%
  st_drop_geometry() %>%
  select(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income) %>%
  drop_na() %>%
  mutate(across(everything(), ~ as.numeric(scale(.)))) %>%
  mutate(median_income = -1 * median_income)

county_index$hardship_index <- rowMeans(county_index, na.rm = TRUE)

county_complete <- county_wide %>%
  drop_na(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income)
county_complete$hardship_index <- county_index$hardship_index
```

```{r}
#| label: step.05-map

county_shifted <- shift_geometry(county_complete)

ggplot(county_shifted) +
  geom_sf(aes(fill = hardship_index), color = NA, linewidth = 0) +
  scale_fill_viridis_c(option = "inferno", name = "Hardship\nIndex", na.value = "grey80") +
  theme_void() +
  labs(
    title    = "Economic Hardship Across U.S. Counties",
    subtitle = "ACS 5-Year (2023) | Higher values = greater hardship",
    caption  = "Alaska and Hawaii shown as insets"
  )
```


# Step 6: Arizona Census Tract Map

```{r}
#| label: step.06-map

ggplot(CenDF_complete) +
  geom_sf(aes(fill = hardship_index), color = NA) +
  scale_fill_viridis(option = "inferno", direction = -1, name = "Hardship\nIndex") +
  theme_void() +
  labs(
    title    = "Economic Hardship Index by Census Tract",
    subtitle = "Arizona — ACS 5-Year (2023)",
    caption  = "Higher values = greater economic hardship"
  )
```


# Step 7: Maricopa County Block Group Map

Block groups are the smallest ACS geography, revealing fine-grained variation within tracts.

```{r}
#| label: step.07-pull
#| results: hide

cbg_df <- get_acs(
  geography    = "block group",
  variables    = hardship_vars,
  state        = "AZ",
  county       = "Maricopa",
  year         = 2023,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
```

```{r}
#| label: step.07-index

cbg_wide <- cbg_df %>%
  select(-moe) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(
    poverty_rate = (pov_below50 + pov_50to99) / pov_den,
    unemp_rate   = pct_unemp_num / pct_unemp_den,
    no_hs_rate   = pct_no_hs_num / pct_no_hs_den,
    no_ins_rate  = pct_no_ins_num / pct_no_ins_den
  )

cbg_index <- cbg_wide %>%
  st_drop_geometry() %>%
  select(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income) %>%
  drop_na() %>%
  mutate(across(everything(), ~ as.numeric(scale(.)))) %>%
  mutate(median_income = -1 * median_income)

cbg_index$hardship_index <- rowMeans(cbg_index, na.rm = TRUE)

cbg_complete <- cbg_wide %>%
  drop_na(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income)
cbg_complete$hardship_index <- cbg_index$hardship_index
```

```{r}
#| label: step.07-map

ggplot(cbg_complete) +
  geom_sf(aes(fill = hardship_index), color = NA) +
  scale_fill_viridis(option = "inferno", direction = -1, name = "Hardship\nIndex") +
  theme_void() +
  labs(
    title    = "Economic Hardship Index by Block Group",
    subtitle = "Maricopa County, AZ — ACS 5-Year (2023)",
    caption  = "Higher values = greater economic hardship"
  )
```


# Step 8: Enriched Index — Does Measurement Choice Change the Map?

The 5-variable baseline captures five well-established dimensions of economic hardship. But hardship is multi-dimensional. A broader index might also reflect:

| New Variable | Dimension | Census Code |
|---|---|---|
| Renter-occupied rate | Housing instability / wealth exclusion | B25003_003 / B25003_001 |
| No vehicle access | Transportation barrier | B08201_002 / B08201_001 |
| Single-parent household rate | Family structure stress | B11012_010 / B11001_001 |
| Public assistance income rate | Direct economic need | B19057_002 / B19057_001 |
| No internet access rate | Digital divide / economic barrier | B28002_013 / B28002_001 |

Adding these five produces a **10-variable enriched index**. The key question: does this broader measurement change which communities we identify as hardship hotspots — and how much?

```{r}
#| label: step.08-pull
#| results: hide

hardship_vars_10 <- c(
  # ── Original 5 variable pairs ──────────────────────────────────────
  pov_below50    = "C17002_002",   pov_50to99     = "C17002_003",
  pov_den        = "C17002_001",   median_income  = "B19013_001",
  pct_unemp_num  = "B23025_005",   pct_unemp_den  = "B23025_002",
  pct_no_hs_num  = "B15003_002",   pct_no_hs_den  = "B15003_001",
  pct_no_ins_num = "B27010_017",   pct_no_ins_den = "B27010_001",
  # ── 5 new dimensions ───────────────────────────────────────────────
  renter_num     = "B25003_003",   renter_den     = "B25003_001",
  noveh_num      = "B08201_002",   noveh_den      = "B08201_001",
  sinpar_num     = "B11012_010",   sinpar_den     = "B11001_001",
  pubast_num     = "B19057_002",   pubast_den     = "B19057_001",
  noinet_num     = "B28002_013",   noinet_den     = "B28002_001"
)

CenDF_10 <- get_acs(
  geography    = "tract",
  variables    = hardship_vars_10,
  state        = "AZ",
  year         = 2023,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
```

```{r}
#| label: step.08-build10

CenDF_wide_10 <- CenDF_10 %>%
  select(-moe) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(
    # Original 5
    poverty_rate = (pov_below50 + pov_50to99) / pov_den,
    unemp_rate   = pct_unemp_num  / pct_unemp_den,
    no_hs_rate   = pct_no_hs_num  / pct_no_hs_den,
    no_ins_rate  = pct_no_ins_num / pct_no_ins_den,
    # New 5
    renter_rate  = renter_num / renter_den,
    noveh_rate   = noveh_num  / noveh_den,
    sinpar_rate  = sinpar_num / sinpar_den,
    pubast_rate  = pubast_num / pubast_den,
    noinet_rate  = noinet_num / noinet_den
  )

CenDF_complete_10 <- CenDF_wide_10 %>%
  drop_na(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate,
          median_income, renter_rate, noveh_rate,
          sinpar_rate, pubast_rate, noinet_rate)

index_z_10 <- CenDF_complete_10 %>%
  st_drop_geometry() %>%
  select(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate,
         median_income, renter_rate, noveh_rate,
         sinpar_rate, pubast_rate, noinet_rate) %>%
  mutate(across(everything(), ~ as.numeric(scale(.)))) %>%
  mutate(median_income = -1 * median_income)   # reverse-code

# Compare alpha: 5-variable vs 10-variable
alpha_5  <- psych::alpha(index_z  |> select(-hardship_index))
alpha_10 <- psych::alpha(index_z_10)

cat("5-variable alpha: ", round(alpha_5$total$raw_alpha,  3), "\n")
cat("10-variable alpha:", round(alpha_10$total$raw_alpha, 3), "\n")
cat("Change:           ", round(alpha_10$total$raw_alpha - alpha_5$total$raw_alpha, 3), "\n")

# Build enriched index
index_z_10$hardship_index_10 <- rowMeans(index_z_10, na.rm = TRUE)
CenDF_complete_10$hardship_index_10 <- index_z_10$hardship_index_10
```

```{r}
#| label: step.08-compare-maps
#| fig-width: 16
#| fig-height: 6

p_5var <- ggplot(CenDF_complete) +
  geom_sf(aes(fill = hardship_index), color = NA) +
  scale_fill_viridis_c(option = "inferno", name = "Hardship",
                       na.value = "grey80") +
  theme_void() +
  labs(title = "5-Variable Index",
       subtitle = "Poverty · Unemployment · Education\nInsurance · Income")

p_10var <- ggplot(CenDF_complete_10) +
  geom_sf(aes(fill = hardship_index_10), color = NA) +
  scale_fill_viridis_c(option = "inferno", name = "Hardship",
                       na.value = "grey80") +
  theme_void() +
  labs(title = "10-Variable Enriched Index",
       subtitle = "Adds renter · no vehicle · single-parent\npublic assistance · no internet")

p_5var + p_10var +
  plot_annotation(
    title    = "Original vs. Enriched Economic Hardship Index",
    subtitle = "Does expanding measurement change the spatial story?",
    caption  = "ACS 5-Year (2023) | Arizona Census Tracts"
  )
```

```{r}
#| label: step.08-difference
#| fig-width: 9
#| fig-height: 6

compare_df <- CenDF_complete_10 %>%
  select(GEOID, geometry, hardship_index_10) %>%
  left_join(
    CenDF_complete %>% st_drop_geometry() %>% select(GEOID, hardship_index),
    by = "GEOID"
  ) %>%
  mutate(diff_10_minus_5 = hardship_index_10 - hardship_index)

ggplot(compare_df) +
  geom_sf(aes(fill = diff_10_minus_5), color = NA) +
  scale_fill_gradient2(
    low = "#4575b4", mid = "#f7f7f7", high = "#d73027",
    midpoint = 0, name = "10-var\nminus\n5-var"
  ) +
  theme_void() +
  labs(
    title    = "Where Does Measurement Choice Matter Most?",
    subtitle = "Red = tracts scoring higher with enriched index\nBlue = tracts scoring lower",
    caption  = "ACS 5-Year (2023) | Difference = 10-variable minus 5-variable index"
  )
```

**Interpreting the difference map:** Red tracts are communities where the enriched index reveals *more hardship* than the baseline captured — likely places with high renter rates, low vehicle access, or many single-parent households that census income and poverty measures undercount. Blue tracts are places where those new dimensions don't add to the hardship signal — often wealthier suburbs where some of those variables don't apply.

---

# Scaffold Chunks

```{r}
#| label: q2-pull
#| results: hide

# YOUR CODE HERE
# Hint: start from hardship_vars_10 and add:
#   snap_num = "B19058_002"   (households receiving SNAP)
#   snap_den = "B19058_001"   (total households)
# Then re-run get_acs() for AZ tracts
```

```{r}
#| label: q2-build11

# YOUR CODE HERE
# Compute snap_rate = snap_num / snap_den
# Add to index variables, z-score, rowMeans -> hardship_index_11
# Run psych::alpha() and print the 10-var vs 11-var comparison
```

```{r}
#| label: q2-maps
#| fig-width: 16
#| fig-height: 6

# YOUR CODE HERE — side-by-side: 10-variable vs 11-variable
```

```{r}
#| label: q2-difference
#| fig-width: 9
#| fig-height: 6

# YOUR CODE HERE — difference map: 11-var minus 10-var
```
