---
title: "Lab 01 — Lab Guide: Measurement, Indexing & Multi-Scale Mapping"
subtitle: "PAF 516 | M1 Assignment"
format:
  html:
    theme: readable
    highlight-style: tango
    toc: true
    self-contained: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
  fig-width: 10
---


```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set( echo = TRUE, message=F, warning=F, fig.width=10 )
options( tigris_use_cache = TRUE )
```


```{r}
#| label: packages
# ── Packages ─────────────────────────────────────────
# Load all required packages for data, mapping, and reliability analysis
library( tidyverse )
library( tidycensus )
library( tigris )      # For shift_geometry (AK/HI insets)
library( psych )       # For Cronbach's alpha
library( corrplot )    # For correlation matrix visualization
library( viridis )     # For color scales on maps
library( ggplot2 )
library( patchwork )   # For combining multi-panel figures
```

```{r}
#| label: api-key
#| eval: false
# Uncomment and add your Census API key
# census_api_key("YOUR KEY GOES HERE")
```

```{r}
#| label: fallback-data
#| eval: false

# ── FALLBACK: Load pre-built datasets if Census API is down ──────
# If the Census API is unavailable or you cannot get an API key,
# uncomment the lines below to load pre-built spatial datasets.
# These contain all census variables, computed rates, hardship index,
# and geometry — ready for mapping.
#
# Option A — from local file:
# county_complete <- readRDS("data/hardship_counties_us.rds")
# CenDF_complete  <- readRDS("data/hardship_tracts_az.rds")
# cbg_complete    <- readRDS("data/hardship_cbg_maricopa.rds")
#
# Option B — from GitHub:
# county_complete <- readRDS(url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab1/data/hardship_counties_us.rds"))
# CenDF_complete  <- readRDS(url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab1/data/hardship_tracts_az.rds"))
# cbg_complete    <- readRDS(url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab1/data/hardship_cbg_maricopa.rds"))
#
# If using fallback data, skip Steps 1-4 and go directly to Step 5.
```


# Step 1: Retrieve Census Variables

We build an **economic hardship index** using 5 census variables at the tract level for Arizona. Each variable captures a different dimension of economic disadvantage.

```{r}
#| label: browse-variables
#| eval: false

# Browse available ACS variables to find relevant indicators
vars_available <- load_variables( 2023, "acs5", cache = TRUE )
# View(vars_available)  # Use this interactively to search
```

```{r}
#| label: step.01
#| eval: false
#| results: hide

# Define our indicator variables for an "economic hardship" construct
# Each variable name is given a readable alias
hardship_vars <- c(
  pov_below50    = "C17002_002",   # Income-to-poverty ratio under .50
  pov_50to99     = "C17002_003",   # Income-to-poverty ratio .50 to .99
  pov_den        = "C17002_001",   # Total population (poverty universe)
  median_income  = "B19013_001",   # Median household income
  pct_unemp_num  = "B23025_005",   # Unemployed population
  pct_unemp_den  = "B23025_002",   # In labor force (denominator for unemployment)
  pct_no_hs_num  = "B15003_002",   # Population with less than high school
  pct_no_hs_den  = "B15003_001",   # Total population 25+ (denominator for education)
  pct_no_ins_num = "B27010_017",   # Population without health insurance (18-34)
  pct_no_ins_den = "B27010_001"    # Total population (insurance universe)
)

# Retrieve tract-level data with geometry for mapping
# Pull ALL Arizona tracts (not just Maricopa) for state-level analysis
CenDF <- get_acs(
  geography = "tract",
  variables = hardship_vars,
  state     = "AZ",
  year      = 2023,
  survey    = "acs5",
  geometry  = TRUE
)

# Fallback: load pre-built data if API unavailable
# (loads processed data — skip to Step 5 for mapping)
# Option A — from local file:
# CenDF_complete <- readRDS("data/hardship_tracts_az.rds")
# Option B — from GitHub:
# CenDF_complete <- readRDS(url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab1/data/hardship_tracts_az.rds"))
```

**Common error note:** If you get an API key error, make sure you have run `census_api_key("YOUR_KEY", install = TRUE)` at least once and restarted R.


# Step 2: Select and Justify Variables

We now reshape the data from long to wide format, compute percentages from numerator/denominator pairs, and standardize all variables to z-scores.

```{r}
#| label: step.02
#| eval: false

# Reshape from long to wide format using pivot_wider
CenDF_wide <- CenDF %>%
  select( -moe ) %>%
  pivot_wider( names_from = variable, values_from = estimate )

# Compute derived percentage variables from numerator/denominator pairs
CenDF_wide <- CenDF_wide %>%
  mutate(
    # Poverty rate: people below poverty / poverty universe
    poverty_rate = ( pov_below50 + pov_50to99 ) / pov_den,
    # Unemployment rate: unemployed / in labor force
    unemp_rate   = pct_unemp_num / pct_unemp_den,
    # Percent without high school diploma
    no_hs_rate   = pct_no_hs_num / pct_no_hs_den,
    # Percent without health insurance
    no_ins_rate  = pct_no_ins_num / pct_no_ins_den
  )

# Select just the indicator variables for our index
# Note: median_income is REVERSE CODED — high income = LOW hardship
index_vars <- CenDF_wide %>%
  as.data.frame() %>%  # drop sf geometry for psych::alpha
  select( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income ) %>%
  drop_na()  # alpha() requires complete cases

# Standardize all variables to z-scores so they are on a common scale
index_z <- index_vars %>%
  mutate( across( everything(), ~ as.numeric( scale(.) ) ) )

# Reverse-code median_income: high income should contribute LOW hardship
# Flipping the sign aligns its direction with the other items
index_z <- index_z %>%
  mutate( median_income = -1 * median_income )

# Verify: all variables should now be positively correlated
# (higher values = more hardship for all items)
```

**Variable justification:**

| Variable | Census Code | Why Included |
|----------|------------|--------------|
| Poverty rate | C17002_002 + C17002_003 / C17002_001 | Direct measure of economic deprivation |
| Unemployment rate | B23025_005 / B23025_002 | Reflects labor market hardship |
| No high school diploma | B15003_002 / B15003_001 | Education is a strong predictor of economic outcomes |
| No health insurance | B27010_017 / B27010_001 | Lack of insurance reflects economic vulnerability |
| Median household income | B19013_001 | Core income measure (reverse coded) |


# Step 3: Assess Reliability with Cronbach's Alpha

```{r}
#| label: step.03
#| eval: false

# Calculate Cronbach's alpha to assess whether our 5 items
# consistently measure the same underlying construct
alpha_result <- psych::alpha( index_z )

# Print the main results
print( alpha_result )
```

**Interpreting the output:**

- The overall `raw_alpha` is our Cronbach's alpha value
- The "Reliability if an item is dropped" section shows what alpha would be without each item
- If dropping an item *increases* alpha, that item may not fit the construct well

```{r}
#| label: alpha-summary
#| eval: false

# Extract the key alpha value for inline reporting
alpha_value <- alpha_result$total$raw_alpha
cat( "Cronbach's alpha:", round( alpha_value, 3 ), "\n" )
```


# Step 4: Build the Index and Visualize

```{r}
#| label: step.04a
#| eval: false

# Correlation matrix of indicator variables
cor_matrix <- cor( index_z, use = "complete.obs" )

# Visualize the correlation matrix
corrplot(
  cor_matrix,
  method  = "color",
  type    = "upper",
  addCoef.col = "black",
  tl.col  = "black",
  tl.srt  = 45,
  title   = "Inter-Item Correlation Matrix: Economic Hardship Index",
  mar     = c(0, 0, 2, 0)
)
```

All items should show positive correlations (after reverse-coding income), confirming they measure a common construct.

```{r}
#| label: step.04b
#| eval: false

# Create the composite index as the row mean of z-scores
# This gives equal weight to each indicator
index_z$hardship_index <- rowMeans( index_z, na.rm = TRUE )

# Histogram of the hardship index
ggplot( index_z, aes( x = hardship_index ) ) +
  geom_histogram( bins = 40, fill = "#8c1d40", color = "white", alpha = 0.8 ) +
  geom_vline( xintercept = mean( index_z$hardship_index, na.rm = TRUE ),
              linetype = "dashed", color = "#ffc627", linewidth = 1 ) +
  theme_minimal() +
  labs(
    title    = "Distribution of Economic Hardship Index",
    subtitle = "Arizona Census Tracts (ACS 2023)",
    x        = "Hardship Index (z-score scale)",
    y        = "Number of Tracts",
    caption  = "Dashed line = mean"
  )
```

```{r}
#| label: step.04c
#| eval: false

# Add the index back to the spatial data for mapping
# Match rows using the complete-case subset
CenDF_complete <- CenDF_wide %>%
  drop_na( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income )

CenDF_complete$hardship_index <- index_z$hardship_index
```


# Step 5: National County-Level Map

We now pull the same 5 hardship variables at the **county level** for the entire US, compute the index using the same method, and map national patterns of economic hardship.

```{r}
#| label: step.05-pull
#| eval: false
#| results: hide

# Pull county-level data for the entire US
# Same 5 variables, different geography
county_df <- get_acs(
  geography = "county",
  variables = hardship_vars,
  year      = 2023,
  survey    = "acs5",
  geometry  = TRUE
)

# Fallback: load pre-built data if API unavailable
# (loads processed data — skip to step.05-map for mapping)
# Option A — from local file:
# county_complete <- readRDS("data/hardship_counties_us.rds")
# Option B — from GitHub:
# county_complete <- readRDS(url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab1/data/hardship_counties_us.rds"))
```

```{r}
#| label: step.05-index
#| eval: false

# Reshape county data from long to wide
county_wide <- county_df %>%
  select( -moe ) %>%
  pivot_wider( names_from = variable, values_from = estimate )

# Compute the same rate variables as the tract level
county_wide <- county_wide %>%
  mutate(
    poverty_rate = ( pov_below50 + pov_50to99 ) / pov_den,
    unemp_rate   = pct_unemp_num / pct_unemp_den,
    no_hs_rate   = pct_no_hs_num / pct_no_hs_den,
    no_ins_rate  = pct_no_ins_num / pct_no_ins_den
  )

# Z-score standardization and reverse-code income
county_index <- county_wide %>%
  as.data.frame() %>%
  select( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income ) %>%
  drop_na() %>%
  mutate( across( everything(), ~ as.numeric( scale(.) ) ) ) %>%
  mutate( median_income = -1 * median_income )

# Composite index: row mean of z-scores
county_index$hardship_index <- rowMeans( county_index, na.rm = TRUE )

# Attach index back to spatial data
county_complete <- county_wide %>%
  drop_na( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income )
county_complete$hardship_index <- county_index$hardship_index
```

```{r}
#| label: step.05-map
#| eval: false

# Use shift_geometry to show Alaska and Hawaii as standard insets
county_shifted <- shift_geometry( county_complete )

# National county-level hardship map with AK/HI insets
map_county <- ggplot( county_shifted ) +
  geom_sf( aes( fill = hardship_index ), color = NA, linewidth = 0 ) +
  scale_fill_viridis_c( option = "inferno",
                        name = "Hardship\nIndex",
                        na.value = "grey80" ) +
  theme_void() +
  labs(
    title    = "Economic Hardship Across U.S. Counties",
    subtitle = "ACS 5-Year (2023) | Higher values = greater hardship",
    caption  = "Alaska and Hawaii shown as insets"
  )

map_county
```


# Step 6: Arizona Census Tract Map

```{r}
#| label: step.06-map
#| eval: false

# Map the hardship index across Arizona tracts
map_tract <- ggplot( CenDF_complete ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis( option = "inferno", direction = -1,
                      name = "Hardship\nIndex" ) +
  theme_void() +
  labs(
    title    = "Economic Hardship Index by Census Tract",
    subtitle = "Arizona — ACS 5-Year (2023)",
    caption  = "Higher values = greater economic hardship"
  )

map_tract
```


# Step 7: Maricopa County Block Group Map

We zoom in further to the **block group** level within Maricopa County. Block groups are the smallest geography for which ACS data are available, revealing fine-grained variation within tracts.

```{r}
#| label: step.07-pull
#| eval: false
#| results: hide

# Pull block group data for Maricopa County
# This is the finest spatial resolution available in the ACS
cbg_df <- get_acs(
  geography = "block group",
  variables = hardship_vars,
  state     = "AZ",
  county    = "Maricopa",
  year      = 2023,
  survey    = "acs5",
  geometry  = TRUE
)

# Fallback: load pre-built data if API unavailable
# (loads processed data — skip to step.07-map for mapping)
# Option A — from local file:
# cbg_complete <- readRDS("data/hardship_cbg_maricopa.rds")
# Option B — from GitHub:
# cbg_complete <- readRDS(url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab1/data/hardship_cbg_maricopa.rds"))
```

```{r}
#| label: step.07-index
#| eval: false

# Reshape and compute rates — same pipeline as tracts and counties
cbg_wide <- cbg_df %>%
  select( -moe ) %>%
  pivot_wider( names_from = variable, values_from = estimate )

cbg_wide <- cbg_wide %>%
  mutate(
    poverty_rate = ( pov_below50 + pov_50to99 ) / pov_den,
    unemp_rate   = pct_unemp_num / pct_unemp_den,
    no_hs_rate   = pct_no_hs_num / pct_no_hs_den,
    no_ins_rate  = pct_no_ins_num / pct_no_ins_den
  )

# Z-score and reverse-code income
cbg_index <- cbg_wide %>%
  as.data.frame() %>%
  select( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income ) %>%
  drop_na() %>%
  mutate( across( everything(), ~ as.numeric( scale(.) ) ) ) %>%
  mutate( median_income = -1 * median_income )

cbg_index$hardship_index <- rowMeans( cbg_index, na.rm = TRUE )

# Attach index to spatial data
cbg_complete <- cbg_wide %>%
  drop_na( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income )
cbg_complete$hardship_index <- cbg_index$hardship_index
```

```{r}
#| label: step.07-map
#| eval: false

# Block-group-level hardship map for Maricopa County
map_cbg <- ggplot( cbg_complete ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis( option = "inferno", direction = -1,
                      name = "Hardship\nIndex" ) +
  theme_void() +
  labs(
    title    = "Economic Hardship Index by Block Group",
    subtitle = "Maricopa County, AZ — ACS 5-Year (2023)",
    caption  = "Higher values = greater economic hardship"
  )

map_cbg
```


# Step 8: Three-Scale Comparison

Using `patchwork`, we place all three maps side by side to compare how spatial patterns of hardship appear at different geographic scales. This illustrates the **Modifiable Areal Unit Problem (MAUP)** — the same phenomenon looks different depending on the unit of analysis.

```{r}
#| label: step.08-patchwork
#| eval: false
#| fig-width: 16
#| fig-height: 6

# Build compact versions of each map for the 3-panel comparison
# Remove subtitles/captions to keep the panel clean
p_county <- ggplot( county_shifted ) +
  geom_sf( aes( fill = hardship_index ), color = NA, linewidth = 0 ) +
  scale_fill_viridis_c( option = "inferno", name = "Index",
                        na.value = "grey80" ) +
  theme_void() +
  labs( title = "County (US)" )

p_tract <- ggplot( CenDF_complete ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis( option = "inferno", direction = -1, name = "Index" ) +
  theme_void() +
  labs( title = "Tract (Arizona)" )

p_cbg <- ggplot( cbg_complete ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis( option = "inferno", direction = -1, name = "Index" ) +
  theme_void() +
  labs( title = "Block Group (Maricopa)" )

# patchwork: combine with + and collect shared legend
p_county + p_tract + p_cbg +
  plot_annotation(
    title   = "Economic Hardship Across Three Geographic Scales",
    caption = "ACS 5-Year (2023) | Higher values = greater hardship"
  )
```


# Save Datasets for Offline Use

The code below saves the three completed spatial datasets as `.rds` files. These files contain all census variables, computed rates, and the hardship index along with geometry. Students can load these if the Census API is unavailable.

```{r}
#| label: save-datasets
#| eval: false

# Save the three completed spatial datasets for offline use
# Each file includes: GEOID, NAME, all census variables, computed rates,
# hardship_index, and sf geometry
saveRDS( county_complete,  "hardship_counties_us.rds" )
saveRDS( CenDF_complete,   "hardship_tracts_az.rds" )
saveRDS( cbg_complete,     "hardship_cbg_maricopa.rds" )

cat( "Saved 3 datasets:\n" )
cat( "  hardship_counties_us.rds   —", nrow(county_complete), "counties\n" )
cat( "  hardship_tracts_az.rds     —", nrow(CenDF_complete), "tracts\n" )
cat( "  hardship_cbg_maricopa.rds  —", nrow(cbg_complete), "block groups\n" )
```


# Questions

### Q1 — Interpret

**What is the Cronbach's alpha? Is it acceptable (>0.7)? Looking at the 3-scale maps, where is hardship concentrated nationally? How does the spatial pattern change as you zoom from counties to tracts to CBGs?**

- **Alpha**: [YOUR ANSWER HERE]

- **National patterns**: [YOUR ANSWER HERE]

- **Scale comparison**: [YOUR ANSWER HERE]


### Q2 — Modify (FULL CODE AND OUTPUT REQUIRED)

**Add a 6th variable to the hardship index — percent receiving SNAP benefits (B22010_002 / B22010_001). Re-pull data at the tract level, recompute Cronbach's alpha with 6 variables, rebuild the index, and re-map at the tract level.**

```{r}
#| label: q2-pull
#| eval: false
#| results: hide

# YOUR CODE HERE
```

```{r}
#| label: q2-index
#| eval: false

# YOUR CODE HERE
```

```{r}
#| label: q2-map
#| eval: false

# YOUR CODE HERE
```


### Q3 — Interpret Modified

**Did adding SNAP improve or weaken Cronbach's alpha? Compare the 5-variable map to the 6-variable map. Should you keep the 6th variable? Why or why not?**

- **Alpha comparison**: [YOUR ANSWER HERE]

- **Map comparison**: [YOUR ANSWER HERE]

- **Recommendation**: [YOUR ANSWER HERE]
