---
title: "Lab 1 Assignment | PAF 516"
subtitle: "Measurement, Indexing & Multi-Scale Mapping"
format:
  html:
    theme: readable
    highlight-style: tango
    toc: true
    toc-title: "Steps & Questions"
    self-contained: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
  fig-width: 10
---

::: {.callout-important}
## Before you begin

1. **Rename this file** to `Lab1_Assignment_YourLastName.qmd` before editing.
2. **This file contains the same pre-filled code as the tutorial.** Steps 1--8 below produce the same output you saw in the tutorial. If this file is in the same folder as your tutorial QMD, you can knit immediately (Cmd+Shift+K / Ctrl+Shift+K) and see the same results. If you placed it in a different location, run the `packages` chunk first (Cmd+Return / Ctrl+Enter), then knit.
3. **The assignment questions are below.** Here is the workflow:
   - **Q1** asks you to interpret results from the pre-filled steps. Knit first, review the Cronbach's alpha output and maps in Steps 3--8, then write your answers.
   - **Q2** asks you to add a 6th variable (SNAP benefits) to the hardship index by adapting the existing code. Fill in the scaffold chunks labeled `q2-...` at the end of this document, then re-knit.
   - **Q3** asks you to evaluate whether the 6th variable improved the index by comparing the original and modified results.
4. Submit both `Lab1_Assignment_YourLastName.qmd` and `Lab1_Assignment_YourLastName.html` to Canvas.
:::


# Assignment Questions

## Q1 --- Interpret the Baseline Results {.question}

> **Text answer --- no code required.** Answer in your own words below.

- **Cronbach's alpha:** What is the value? Is it acceptable (> 0.7)? What does it tell you about the 5-variable index?

[YOUR ANSWER HERE]

- **National patterns:** Where is economic hardship most concentrated in the US county map?

[YOUR ANSWER HERE]

- **Scale comparison:** How does the spatial pattern change as you zoom from counties to tracts to block groups? What does this illustrate about the MAUP?

[YOUR ANSWER HERE]


## Q2 --- Add a 6th Variable {.question}

> **Code + output required.** Add **percent of households receiving SNAP benefits** as a 6th indicator. Census codes: numerator `B19058_002`, denominator `B19058_001`.
>
> Fill in the three scaffold chunks labeled `q2-pull`, `q2-index`, and `q2-map` at the **end of this document** (after Step 8), then re-knit to see your results.


## Q3 --- Evaluate the Addition {.question}

> **Text answer --- no code required.**

- **Alpha comparison:** Did adding SNAP improve or weaken Cronbach's alpha? By how much?

[YOUR ANSWER HERE]

- **Map comparison:** Visually compare the 5-variable and 6-variable tract maps. Do the patterns differ meaningfully?

[YOUR ANSWER HERE]

- **Recommendation:** Based on alpha and the maps, should you keep the 6th variable in the index? Why or why not?

[YOUR ANSWER HERE]


---


```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10)
options(tigris_use_cache = TRUE)
options(tigris_progress_bar = FALSE)
```

```{r}
#| label: packages
# ── Package Setup (runs automatically on first knit) ─────────────────
# Downloads the exact package versions used to build this lab.
# The if-guard ensures this only runs once — no need to comment out.
if (!file.exists(".renv_done")) {
  if (!requireNamespace("renv", quietly = TRUE)) install.packages("renv")
  download.file(
    url      = "https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab1/renv.lock",
    destfile = "renv.lock", mode = "wb"
  )
  renv::restore(lockfile = "renv.lock", prompt = FALSE)
  file.create(".renv_done")
}

library(tidyverse)
library(tidycensus)
library(sf)
library(tigris)
library(psych)
library(corrplot)
library(viridis)
library(ggplot2)
library(patchwork)
```

```{r}
#| label: api-key
# Get a free Census API key at: https://api.census.gov/data/key_signup.html
# After receiving your key by email, run this line ONCE in the RStudio console:
# census_api_key("YOUR_KEY_HERE", install = TRUE)
# Then restart R: Session → Restart R
# You only need to do this once — the key is saved to your .Renviron file.
```

```{r}
#| label: fallback-data
# ── FALLBACK: Uncomment if Census API is unavailable ──────────────────
# Option A (local — save .rds files to PAF516/Lab1/data/ first):
# county_complete <- readRDS("../data/hardship_counties_us.rds")
# CenDF_complete  <- readRDS("../data/hardship_tracts_az.rds")
# cbg_complete    <- readRDS("../data/hardship_cbg_maricopa.rds")
#
# Option B (GitHub):
# county_complete <- readRDS(url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab1/data/hardship_counties_us.rds"))
# CenDF_complete  <- readRDS(url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab1/data/hardship_tracts_az.rds"))
# cbg_complete    <- readRDS(url("https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/Lab1/data/hardship_cbg_maricopa.rds"))
# If using fallback, skip to Step 5.
```


# Step 1: Retrieve Census Variables

```{r}
#| label: step.01
#| results: hide

hardship_vars <- c(
  pov_below50    = "C17002_002",
  pov_50to99     = "C17002_003",
  pov_den        = "C17002_001",
  median_income  = "B19013_001",
  pct_unemp_num  = "B23025_005",
  pct_unemp_den  = "B23025_002",
  pct_no_hs_num  = "B15003_002",
  pct_no_hs_den  = "B15003_001",
  pct_no_ins_num = "B27010_017",
  pct_no_ins_den = "B27010_001"
)

CenDF <- get_acs(
  geography    = "tract",
  variables    = hardship_vars,
  state        = "AZ",
  year         = 2023,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
```

# Step 2: Compute Rates and Standardize

```{r}
#| label: step.02

CenDF_wide <- CenDF %>%
  select(-moe) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(
    poverty_rate = (pov_below50 + pov_50to99) / pov_den,
    unemp_rate   = pct_unemp_num / pct_unemp_den,
    no_hs_rate   = pct_no_hs_num / pct_no_hs_den,
    no_ins_rate  = pct_no_ins_num / pct_no_ins_den
  )

index_vars <- CenDF_wide %>%
  st_drop_geometry() %>%
  select(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income) %>%
  drop_na()

index_z <- index_vars %>%
  mutate(across(everything(), ~ as.numeric(scale(.)))) %>%
  mutate(median_income = -1 * median_income)
```

# Step 3: Cronbach's Alpha

```{r}
#| label: step.03

alpha_result <- psych::alpha(index_z)
print(alpha_result)
alpha_value  <- alpha_result$total$raw_alpha
cat("Cronbach's alpha:", round(alpha_value, 3), "\n")
```

# Step 4: Build Index and Visualize

```{r}
#| label: step.04a

corrplot(
  cor(index_z, use = "complete.obs"),
  method = "color", type = "upper", addCoef.col = "black",
  tl.col = "black", tl.srt = 45,
  title = "Inter-Item Correlation Matrix: Economic Hardship Index",
  mar = c(0, 0, 2, 0)
)
```

```{r}
#| label: step.04b

index_z$hardship_index <- rowMeans(index_z, na.rm = TRUE)

ggplot(index_z, aes(x = hardship_index)) +
  geom_histogram(bins = 40, fill = "#8c1d40", color = "white", alpha = 0.8) +
  geom_vline(xintercept = mean(index_z$hardship_index, na.rm = TRUE),
             linetype = "dashed", color = "#ffc627", linewidth = 1) +
  theme_minimal() +
  labs(title = "Distribution of Economic Hardship Index",
       subtitle = "Arizona Census Tracts (ACS 2023)",
       x = "Hardship Index (z-score scale)", y = "Number of Tracts")
```

```{r}
#| label: step.04c

CenDF_complete <- CenDF_wide %>%
  drop_na(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income)
CenDF_complete$hardship_index <- index_z$hardship_index
```

# Step 5: National County-Level Map

```{r}
#| label: step.05-pull
#| results: hide

county_df <- get_acs(
  geography = "county", variables = hardship_vars,
  year = 2023, survey = "acs5", geometry = TRUE, progress_bar = FALSE
)
```

```{r}
#| label: step.05-index

county_wide <- county_df %>%
  select(-moe) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(
    poverty_rate = (pov_below50 + pov_50to99) / pov_den,
    unemp_rate   = pct_unemp_num / pct_unemp_den,
    no_hs_rate   = pct_no_hs_num / pct_no_hs_den,
    no_ins_rate  = pct_no_ins_num / pct_no_ins_den
  )

county_index <- county_wide %>%
  st_drop_geometry() %>%
  select(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income) %>%
  drop_na() %>%
  mutate(across(everything(), ~ as.numeric(scale(.)))) %>%
  mutate(median_income = -1 * median_income)

county_index$hardship_index <- rowMeans(county_index, na.rm = TRUE)
county_complete <- county_wide %>%
  drop_na(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income)
county_complete$hardship_index <- county_index$hardship_index
```

```{r}
#| label: step.05-map

county_shifted <- shift_geometry(county_complete)

ggplot(county_shifted) +
  geom_sf(aes(fill = hardship_index), color = NA, linewidth = 0) +
  scale_fill_viridis_c(option = "inferno", name = "Hardship\nIndex", na.value = "grey80") +
  theme_void() +
  labs(title = "Economic Hardship Across U.S. Counties",
       subtitle = "ACS 5-Year (2023) | Higher values = greater hardship",
       caption = "Alaska and Hawaii shown as insets")
```

# Step 6: Arizona Census Tract Map

```{r}
#| label: step.06-map

ggplot(CenDF_complete) +
  geom_sf(aes(fill = hardship_index), color = NA) +
  scale_fill_viridis(option = "inferno", direction = -1, name = "Hardship\nIndex") +
  theme_void() +
  labs(title = "Economic Hardship Index by Census Tract",
       subtitle = "Arizona — ACS 5-Year (2023)")
```

# Step 7: Maricopa County Block Group Map

```{r}
#| label: step.07-pull
#| results: hide

cbg_df <- get_acs(
  geography = "block group", variables = hardship_vars,
  state = "AZ", county = "Maricopa",
  year = 2023, survey = "acs5", geometry = TRUE, progress_bar = FALSE
)
```

```{r}
#| label: step.07-index

cbg_wide <- cbg_df %>%
  select(-moe) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(
    poverty_rate = (pov_below50 + pov_50to99) / pov_den,
    unemp_rate   = pct_unemp_num / pct_unemp_den,
    no_hs_rate   = pct_no_hs_num / pct_no_hs_den,
    no_ins_rate  = pct_no_ins_num / pct_no_ins_den
  )

cbg_index <- cbg_wide %>%
  st_drop_geometry() %>%
  select(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income) %>%
  drop_na() %>%
  mutate(across(everything(), ~ as.numeric(scale(.)))) %>%
  mutate(median_income = -1 * median_income)

cbg_index$hardship_index <- rowMeans(cbg_index, na.rm = TRUE)
cbg_complete <- cbg_wide %>%
  drop_na(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income)
cbg_complete$hardship_index <- cbg_index$hardship_index
```

```{r}
#| label: step.07-map

ggplot(cbg_complete) +
  geom_sf(aes(fill = hardship_index), color = NA) +
  scale_fill_viridis(option = "inferno", direction = -1, name = "Hardship\nIndex") +
  theme_void() +
  labs(title = "Economic Hardship Index by Block Group",
       subtitle = "Maricopa County, AZ — ACS 5-Year (2023)")
```

# Step 8: Three-Scale Comparison

```{r}
#| label: step.08-patchwork
#| fig-width: 16
#| fig-height: 6

p_county <- ggplot(county_shifted) +
  geom_sf(aes(fill = hardship_index), color = NA, linewidth = 0) +
  scale_fill_viridis_c(option = "inferno", name = "Index", na.value = "grey80") +
  theme_void() + labs(title = "County (US)")

p_tract <- ggplot(CenDF_complete) +
  geom_sf(aes(fill = hardship_index), color = NA) +
  scale_fill_viridis(option = "inferno", direction = -1, name = "Index") +
  theme_void() + labs(title = "Tract (Arizona)")

p_cbg <- ggplot(cbg_complete) +
  geom_sf(aes(fill = hardship_index), color = NA) +
  scale_fill_viridis(option = "inferno", direction = -1, name = "Index") +
  theme_void() + labs(title = "Block Group (Maricopa)")

p_county + p_tract + p_cbg +
  plot_annotation(
    title   = "Economic Hardship Across Three Geographic Scales",
    caption = "ACS 5-Year (2023) | Higher values = greater hardship"
  )
```


---

# Q2 --- Your Code

> Fill in the three chunks below, then re-knit to see your results. Add **percent of households receiving SNAP benefits** (`B19058_002` / `B19058_001`) as a 6th indicator to the hardship index.

```{r}
#| label: q2-pull
#| results: hide

# YOUR CODE HERE
# Hint: add snap_num = "B19058_002" and snap_den = "B19058_001"
# to hardship_vars, then re-run get_acs() for AZ tracts
```

```{r}
#| label: q2-index

# YOUR CODE HERE
# Hint: compute snap_rate = snap_num / snap_den
# Add snap_rate to the index_vars select() call
# Standardize to z-scores (no reverse coding needed for SNAP)
# Run psych::alpha() on the 6-variable index_z
# Build hardship_index_6 as rowMeans of the 6 z-scores
```

```{r}
#| label: q2-map

# YOUR CODE HERE
# Map the 6-variable hardship index at the Arizona tract level
# Use the same ggplot / viridis style as Step 6
```
