---
title: "Lab 04 — Solutions: Spatial Autocorrelation & Hot Spots"
subtitle: "PAF 516 | M4 Assignment"
date: today
format:
  html:
    theme: readable
    highlight-style: tango
    toc: true
    self-contained: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
  fig-width: 10
---

---

> **Solutions File:** This document contains the fully executed tutorial code
> with all outputs rendered, plus completed responses for Q1--Q3. Code
> annotations explain every new function and concept.

---

# Setup & Packages

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10)

options(tigris_use_cache = TRUE)   # Cache shapefiles so re-runs are fast
```

```{r}
#| label: packages

library(tidyverse)    # Data wrangling and ggplot2
library(tidycensus)   # Pull ACS data with geometry
library(sf)           # Simple features for spatial data
library(spdep)        # Spatial dependence: weights, Moran's I, LISA
library(viridis)      # Perceptually uniform color scales
library(patchwork)    # Combine multiple ggplots side-by-side
```

::: {.callout-note}
## New package: `spdep`
The `spdep` package (spatial dependence) provides the tools for this entire
lab: building neighbor relationships, computing spatial weights, running
Moran's I tests, and classifying local clusters. Install it with
`install.packages("spdep")` if you have not already.
:::

---

# Step 1: Pull Maricopa County CBG Data & Build Hardship Index

By now you have done this four times. We pull the same five ACS hardship
variables used in Lab 1, but this time at the **Census Block Group (CBG)**
level for Maricopa County. CBGs are smaller than tracts (roughly 600--3,000
people), giving us finer spatial resolution to detect local clustering.

```{r}
#| label: pull-data
#| results: hide

# Define the ACS variables for our economic hardship index.
# Each pair is a numerator/denominator for computing a rate,
# except median_income which is a single estimate.
hardship_vars <- c(
  pov_num      = "B17001_002",   # Population below poverty level
  pov_den      = "B17001_001",   # Total population (poverty universe)
  unemp_num    = "B23025_005",   # Unemployed population
  unemp_den    = "B23025_002",   # Population in labor force
  no_hs_num    = "B15003_002",   # No high school diploma (less than 9th grade)
  no_hs_den    = "B15003_001",   # Total population 25+
  no_ins_num   = "B27010_017",   # Without health insurance
  no_ins_den   = "B27010_001",   # Total pop (insurance universe)
  median_income = "B19013_001"   # Median household income
)

# Pull at block group level for Maricopa County, AZ
cbg_raw <- get_acs(
  geography = "block group",
  variables = hardship_vars,
  state     = "AZ",
  county    = "Maricopa",
  year      = 2023,
  survey    = "acs5",
  geometry  = TRUE
)
```

```{r}
#| label: build-index

# Reshape from long to wide: one row per CBG, one column per variable
cbg_wide <- cbg_raw %>%
  select(-moe) %>%
  pivot_wider(names_from = variable, values_from = estimate)

# Compute rates from numerator/denominator pairs
cbg_wide <- cbg_wide %>%
  mutate(
    poverty_rate = pov_num / pov_den,
    unemp_rate   = unemp_num / unemp_den,
    no_hs_rate   = no_hs_num / no_hs_den,
    no_ins_rate  = no_ins_num / no_ins_den
  )

# Standardize all five indicators to z-scores, then reverse median income
# (high income = LOW hardship, so we flip the sign)
cbg_index <- cbg_wide %>%
  filter(!is.na(poverty_rate),
         !is.na(unemp_rate),
         !is.na(no_hs_rate),
         !is.na(no_ins_rate),
         !is.na(median_income)) %>%
  mutate(
    z_poverty  = as.numeric(scale(poverty_rate)),
    z_unemp   = as.numeric(scale(unemp_rate)),
    z_no_hs   = as.numeric(scale(no_hs_rate)),
    z_no_ins  = as.numeric(scale(no_ins_rate)),
    z_income  = -1 * as.numeric(scale(median_income))
  ) %>%
  mutate(
    hardship_index = (z_poverty + z_unemp + z_no_hs + z_no_ins + z_income) / 5
  )
```

```{r}
#| label: map-hardship-index
#| fig-height: 8

# Quick choropleth to see the spatial pattern before we test it statistically
ggplot(cbg_index) +
  geom_sf(aes(fill = hardship_index), color = NA) +
  scale_fill_viridis(option = "inferno", name = "Hardship\nIndex") +
  labs(
    title    = "Economic Hardship Index — Maricopa County CBGs",
    subtitle = "Higher values = greater hardship (z-score composite)",
    caption  = "Source: ACS 2019-2023 5-Year Estimates"
  ) +
  theme_void()
```

The map shows a clear visual pattern: high-hardship CBGs appear to cluster
together in certain parts of the county, particularly in south Phoenix and
parts of the west valley. But is this clustering statistically significant,
or could it be random? That is the question spatial autocorrelation answers.

---

# Step 2: Build Spatial Weights

Before we can test for spatial clustering, we need to define what "neighbor"
means. We use **queen contiguity**: two CBGs are neighbors if they share any
boundary point (including corners), similar to how a queen moves in chess.
The alternative is **rook contiguity**, which only counts shared edges (not
corners). Queen is the standard default because it captures more neighbor
relationships.

```{r}
#| label: spatial-weights

# CRITICAL: st_make_valid() fixes any geometry errors (self-intersections,
# invalid rings) in the shapefile. Without this, poly2nb() will often fail
# or produce incorrect neighbor lists.
cbg_index <- st_make_valid(cbg_index)

# poly2nb() — "polygons to neighbors"
# Identifies which CBGs share a boundary. Returns a "nb" (neighbor) object:
# a list where element i contains the indices of all neighbors of polygon i.
# queen=TRUE means corners count as shared boundaries (the default).
nb <- poly2nb(cbg_index, queen = TRUE)

# nb2listw() — "neighbors to list of weights"
# Converts the neighbor list into a spatial weights object that Moran's I needs.
# style="W" means ROW-STANDARDIZED weights: each neighbor gets weight 1/n
# where n is the number of neighbors for that polygon. This means each CBG's
# neighbors collectively contribute a weight of 1.0, regardless of how many
# neighbors it has.
# zero.policy=TRUE: some CBGs may be "islands" with no contiguous neighbors
# (e.g., disconnected polygons). Without this flag, the function would error.
# With it, islands are allowed but excluded from calculations.
lw <- nb2listw(nb, style = "W", zero.policy = TRUE)
```

::: {.callout-tip}
## Row-standardization explained
If CBG A has 3 neighbors, each gets weight 1/3. If CBG B has 6 neighbors,
each gets weight 1/6. This way, the **spatial lag** (weighted average of
neighbors) is a true average regardless of how many neighbors a polygon has.
Without row-standardization, CBGs with more neighbors would have
disproportionate influence.
:::

---

# Step 3: Global Moran's I

Global Moran's I is a single summary statistic for the **entire study area**.
It answers: "Overall, is the hardship index spatially clustered, randomly
distributed, or dispersed?"

- **I near +1** = strong positive spatial autocorrelation (clustering: similar
  values near each other)
- **I near 0** = spatial randomness (no pattern)
- **I near -1** = spatial dispersion (checkerboard: dissimilar values near
  each other)

```{r}
#| label: global-morans-i

# moran.test() computes Global Moran's I and tests significance.
# The null hypothesis is spatial randomness (CSR = complete spatial randomness).
# zero.policy=TRUE handles any island CBGs.
moran_global <- moran.test(
  cbg_index$hardship_index,
  lw,
  zero.policy = TRUE
)

moran_global
```

```{r}
#| label: interpret-global

# Extract key values for interpretation
cat("Moran's I statistic:", round(moran_global$estimate["Moran I statistic"], 4), "\n")
cat("Expected I under randomness:", round(moran_global$estimate["Expectation"], 4), "\n")
cat("p-value:", format.pval(moran_global$p.value, digits = 3), "\n")
```

**Interpretation:** The Global Moran's I is positive and the p-value is
extremely small (well below 0.05), meaning we reject the null hypothesis of
spatial randomness. Hardship values are **significantly clustered** in
Maricopa County --- CBGs with high hardship tend to be surrounded by other
high-hardship CBGs, and vice versa.

---

# Step 4: Moran Scatter Plot

The Moran scatter plot visualizes the relationship between each CBG's
hardship value (x-axis) and the average hardship of its neighbors --- the
**spatial lag** (y-axis). The slope of the fitted line equals Global Moran's I.

The four quadrants define the type of spatial association for each observation:

- **HH (upper-right):** High value surrounded by high neighbors (hot spot)
- **LL (lower-left):** Low value surrounded by low neighbors (cold spot)
- **HL (lower-right):** High value surrounded by low neighbors (spatial outlier)
- **LH (upper-left):** Low value surrounded by high neighbors (spatial outlier)

```{r}
#| label: moran-scatter-plot
#| fig-height: 7

# moran.plot() creates the scatter plot with quadrant lines at the means.
# Points in the upper-right (HH) and lower-left (LL) quadrants drive
# positive spatial autocorrelation. Points in HL and LH work against it.
moran.plot(
  cbg_index$hardship_index,
  lw,
  zero.policy = TRUE,
  xlab = "Hardship Index (standardized)",
  ylab = "Spatial Lag of Hardship Index",
  main = "Moran Scatter Plot — Economic Hardship",
  pch  = 20,
  col  = "steelblue"
)
```

Most points fall in the HH and LL quadrants, confirming positive spatial
autocorrelation. The relatively few points in the HL and LH quadrants are
spatial outliers --- CBGs that differ markedly from their neighbors.

---

# Step 5: Local Moran's I (LISA)

Global Moran's I tells us clustering exists *overall*, but not *where*.
**Local Indicators of Spatial Association (LISA)** decompose the global
statistic into a value for **each CBG**, telling us:

1. Whether that specific CBG is part of a significant cluster
2. What *type* of cluster it belongs to (HH, LL, HL, or LH)

```{r}
#| label: lisa-analysis

# localmoran() returns a matrix with one row per CBG and columns:
#   Ii      = local Moran's I for that CBG
#   E.Ii    = expected value under randomness
#   Var.Ii  = variance
#   Z.Ii    = z-score
#   Pr(z != E(Ii)) = p-value (two-sided test)
lisa <- localmoran(
  cbg_index$hardship_index,
  lw,
  zero.policy = TRUE
)

# Attach LISA results to the spatial data frame
cbg_index$lisa_i     <- lisa[, "Ii"]
cbg_index$lisa_z     <- lisa[, "Z.Ii"]
cbg_index$lisa_p     <- lisa[, "Pr(z != E(Ii))"]

# Compute the spatial lag (average of neighbors' values) for quadrant classification
cbg_index$lag_hardship <- lag.listw(lw, cbg_index$hardship_index, zero.policy = TRUE)
```

```{r}
#| label: lisa-classify

# Classify each CBG into a LISA category based on two criteria:
# 1. Statistical significance: p-value <= 0.05
# 2. Quadrant: determined by the signs of the value and its spatial lag
#
# The logic:
#   - Center the variable and its lag at their means
#   - If both are above mean: HH (hot spot)
#   - If both are below mean: LL (cold spot)
#   - If value above but lag below: HL (high outlier surrounded by low)
#   - If value below but lag above: LH (low outlier surrounded by high)
#   - If p > 0.05: Not Significant regardless of quadrant

cbg_index <- cbg_index %>%
  mutate(
    # Center at the mean
    val_centered = hardship_index - mean(hardship_index, na.rm = TRUE),
    lag_centered = lag_hardship   - mean(lag_hardship, na.rm = TRUE),

    # Assign quadrant labels
    quadrant = case_when(
      val_centered > 0 & lag_centered > 0 ~ "HH",
      val_centered < 0 & lag_centered < 0 ~ "LL",
      val_centered > 0 & lag_centered < 0 ~ "HL",
      val_centered < 0 & lag_centered > 0 ~ "LH",
      TRUE ~ "NS"
    ),

    # Apply significance filter: only keep label if p <= 0.05
    lisa_cluster = if_else(lisa_p <= 0.05, quadrant, "NS")
  )

# How many CBGs fall into each category?
table(cbg_index$lisa_cluster)
```

---

# Step 6: LISA Cluster Map

This is the payoff map. Each CBG is colored by its LISA classification:

- **Red (HH):** Hot spots --- high-hardship CBGs surrounded by high-hardship neighbors
- **Blue (LL):** Cold spots --- low-hardship CBGs surrounded by low-hardship neighbors
- **Light coral (HL):** High-hardship outliers in low-hardship areas
- **Light blue (LH):** Low-hardship outliers in high-hardship areas
- **Grey (NS):** Not statistically significant at p <= 0.05

```{r}
#| label: lisa-cluster-map
#| fig-height: 9

# Define the color scheme for LISA categories
lisa_colors <- c(
  "HH" = "red",
  "LL" = "blue",
  "HL" = "#F08080",    # light coral
  "LH" = "#ADD8E6",    # light blue
  "NS" = "grey90"
)

# Convert to factor with explicit levels so the legend is ordered correctly
cbg_index$lisa_cluster <- factor(
  cbg_index$lisa_cluster,
  levels = c("HH", "LL", "HL", "LH", "NS")
)

ggplot(cbg_index) +
  geom_sf(aes(fill = lisa_cluster), color = "white", linewidth = 0.05) +
  scale_fill_manual(
    values = lisa_colors,
    labels = c("HH (Hot Spot)",
               "LL (Cold Spot)",
               "HL (High-Low Outlier)",
               "LH (Low-High Outlier)",
               "NS (Not Significant)"),
    name   = "LISA Cluster"
  ) +
  labs(
    title    = "LISA Cluster Map — Economic Hardship Index",
    subtitle = "Maricopa County CBGs | Queen contiguity, p <= 0.05",
    caption  = "Source: ACS 2019-2023 5-Year Estimates"
  ) +
  theme_void() +
  theme(legend.position = "right")
```

```{r}
#| label: side-by-side
#| fig-height: 8

# Side-by-side comparison: raw hardship index vs. LISA clusters
p1 <- ggplot(cbg_index) +
  geom_sf(aes(fill = hardship_index), color = NA) +
  scale_fill_viridis(option = "inferno", name = "Hardship\nIndex") +
  labs(title = "Hardship Index (Raw Values)") +
  theme_void()

p2 <- ggplot(cbg_index) +
  geom_sf(aes(fill = lisa_cluster), color = "white", linewidth = 0.05) +
  scale_fill_manual(values = lisa_colors, name = "LISA\nCluster") +
  labs(title = "LISA Clusters (Significant Only)") +
  theme_void()

# patchwork: the + operator places plots side by side
p1 + p2
```

The side-by-side comparison shows the power of LISA analysis: the left map
shows the raw pattern (which we could see visually), while the right map
tells us which parts of that pattern are **statistically significant**. Many
areas that *look* like they might be clusters turn out to be not significant
at the 0.05 level.

---

# Q1 — Interpret the Global Moran's I and LISA Results

::: {.callout-important}
## Question 1

What is the Global Moran's I value and is it statistically significant?
What does this tell you about the spatial distribution of hardship in
Maricopa County? Identify 2--3 specific hot spot areas from the LISA map.
:::

**Sample Answer:**

The Global Moran's I is positive and highly statistically significant
(p < 0.001), confirming that the spatial distribution of economic hardship
in Maricopa County is **not random**. High-hardship CBGs tend to cluster
near other high-hardship CBGs, and low-hardship CBGs cluster near other
low-hardship CBGs. This means that economic disadvantage in Maricopa County
is spatially concentrated rather than evenly dispersed.

From the LISA cluster map, prominent **hot spot (HH)** concentrations appear in:

1. **South Phoenix / South Mountain area** --- a long-documented area of
   concentrated poverty and disinvestment, with high rates of poverty,
   unemployment, and low educational attainment.
2. **West-central Phoenix (Maryvale corridor)** --- another area with
   historically high hardship indicators, including large immigrant
   communities with lower insurance coverage and educational attainment.
3. **Parts of Mesa or Guadalupe** --- smaller pockets of concentrated
   hardship that emerge at the CBG level.

**Cold spots (LL)** --- areas of concentrated low hardship --- appear in
northeast Scottsdale, Paradise Valley, and parts of north Phoenix, which
are among the county's most affluent areas.

---

# Q2 — Modify: LISA Analysis on Poverty Rate Alone

::: {.callout-important}
## Question 2

Rerun the entire LISA analysis on a **single component variable** ---
poverty rate alone --- instead of the composite hardship index. Show
complete code for spatial weights, Global Moran's I, LISA classification,
and the cluster map.
:::

```{r}
#| label: q2-poverty-morans

# ── Global Moran's I for poverty rate ──
# We reuse the same spatial weights (lw) since the geography hasn't changed.
# We just swap the variable from hardship_index to poverty_rate.

moran_poverty <- moran.test(
  cbg_index$poverty_rate,
  lw,
  zero.policy = TRUE
)

moran_poverty
```

```{r}
#| label: q2-poverty-lisa

# ── Local Moran's I (LISA) for poverty rate ──
lisa_pov <- localmoran(
  cbg_index$poverty_rate,
  lw,
  zero.policy = TRUE
)

# Attach results
cbg_index$lisa_pov_i <- lisa_pov[, "Ii"]
cbg_index$lisa_pov_p <- lisa_pov[, "Pr(z != E(Ii))"]

# Compute spatial lag of poverty rate
cbg_index$lag_poverty <- lag.listw(lw, cbg_index$poverty_rate, zero.policy = TRUE)

# Classify LISA quadrants for poverty
cbg_index <- cbg_index %>%
  mutate(
    pov_centered     = poverty_rate - mean(poverty_rate, na.rm = TRUE),
    lag_pov_centered = lag_poverty  - mean(lag_poverty, na.rm = TRUE),

    pov_quadrant = case_when(
      pov_centered > 0 & lag_pov_centered > 0 ~ "HH",
      pov_centered < 0 & lag_pov_centered < 0 ~ "LL",
      pov_centered > 0 & lag_pov_centered < 0 ~ "HL",
      pov_centered < 0 & lag_pov_centered > 0 ~ "LH",
      TRUE ~ "NS"
    ),

    lisa_pov_cluster = factor(
      if_else(lisa_pov_p <= 0.05, pov_quadrant, "NS"),
      levels = c("HH", "LL", "HL", "LH", "NS")
    )
  )

table(cbg_index$lisa_pov_cluster)
```

```{r}
#| label: q2-poverty-lisa-map
#| fig-height: 9

# ── LISA Cluster Map for Poverty Rate ──
ggplot(cbg_index) +
  geom_sf(aes(fill = lisa_pov_cluster), color = "white", linewidth = 0.05) +
  scale_fill_manual(
    values = lisa_colors,
    labels = c("HH (Hot Spot)",
               "LL (Cold Spot)",
               "HL (High-Low Outlier)",
               "LH (Low-High Outlier)",
               "NS (Not Significant)"),
    name   = "LISA Cluster"
  ) +
  labs(
    title    = "LISA Cluster Map — Poverty Rate Only",
    subtitle = "Maricopa County CBGs | Queen contiguity, p <= 0.05",
    caption  = "Source: ACS 2019-2023 5-Year Estimates"
  ) +
  theme_void() +
  theme(legend.position = "right")
```

---

# Q3 — Compare Poverty Hot Spots vs. Composite Index Hot Spots

::: {.callout-important}
## Question 3

Compare the hot spot map for poverty alone vs. the composite hardship index.
Do the hot spots shift? What does this tell you about whether the composite
index captures spatial patterns differently than individual variables?
:::

```{r}
#| label: q3-comparison-maps
#| fig-height: 8

# Side-by-side LISA cluster maps for direct visual comparison
p_composite <- ggplot(cbg_index) +
  geom_sf(aes(fill = lisa_cluster), color = "white", linewidth = 0.05) +
  scale_fill_manual(values = lisa_colors, name = "LISA") +
  labs(title = "Composite Hardship Index") +
  theme_void() +
  theme(legend.position = "bottom")

p_poverty <- ggplot(cbg_index) +
  geom_sf(aes(fill = lisa_pov_cluster), color = "white", linewidth = 0.05) +
  scale_fill_manual(values = lisa_colors, name = "LISA") +
  labs(title = "Poverty Rate Only") +
  theme_void() +
  theme(legend.position = "bottom")

p_composite + p_poverty
```

```{r}
#| label: q3-cross-tabulation

# Cross-tabulate the two cluster classifications to see overlap
table(
  Composite = cbg_index$lisa_cluster,
  Poverty   = cbg_index$lisa_pov_cluster
)
```

**Sample Answer:**

The two LISA maps share a common core of hot spots but differ at the margins.
The major clusters in south Phoenix and the Maryvale corridor appear in both
maps, which makes sense because poverty is the strongest single driver of the
composite index. However, there are notable differences:

1. **The composite index identifies more hot spot CBGs** than poverty alone
   in some areas. This is because a CBG might have moderate poverty but also
   high unemployment, low education, and no insurance --- the combination
   pushes it into significant clustering in the composite but not in poverty
   alone.

2. **Some cold spots shift.** Areas that are LL (cold spots) in the composite
   may not be LL for poverty alone if they have low poverty but average-to-high
   values on other hardship indicators (e.g., a middle-income area with low
   poverty but high uninsurance rates).

3. **Spatial outliers may differ.** A CBG classified as HL (high outlier) on
   the composite might be NS on poverty alone, suggesting that its
   "outlier" status comes from non-poverty dimensions of hardship.

This comparison demonstrates the value of composite indices: they capture
**multidimensional disadvantage** that no single variable can represent. A
neighborhood may not be poor in income terms but may still experience
concentrated hardship through unemployment, low education, and lack of
insurance. The composite index reveals these overlapping vulnerabilities
that a single-variable analysis would miss.

---

# Common Errors & Troubleshooting

::: {.callout-warning}
## Frequently encountered issues

**1. `poly2nb()` fails with topology errors**
Always run `st_make_valid()` before building neighbor lists. Invalid
geometries (self-intersections, ring errors) are common in Census
shapefiles and will cause `poly2nb()` to crash or produce wrong results.

**2. `Error: empty neighbour sets found`**
Some CBGs may be "islands" (no shared boundary with any other CBG). Use
`zero.policy=TRUE` in `nb2listw()`, `moran.test()`, and `localmoran()` to
handle these. Islands are excluded from spatial calculations rather than
causing errors.

**3. All LISA clusters show as NS (Not Significant)**
Check that you are using the correct variable. If the variable has very
little spatial variation (e.g., nearly all CBGs have the same value),
Moran's I will be near zero and no local clusters will be significant.

**4. Colors do not match the legend**
Make sure the `lisa_cluster` column is a factor with levels in the same
order as your color vector: `c("HH", "LL", "HL", "LH", "NS")`.

**5. `lag.listw()` returns NA values**
This happens for island CBGs when `zero.policy=TRUE`. These CBGs have no
neighbors, so no spatial lag can be computed. They will be classified as
NS in the LISA analysis.
:::
