---
title: "Spatial Data Integration"
subtitle: "Module 3 | PAF 516 — Community Analytics"
author: "Arizona State University"
format:
  beamer:
    theme: Madrid
    colortheme: beaver
    fonttheme: professionalfonts
    slide-level: 2
    aspectratio: 169
    toc: false
    include-in-header:
      text: |
        \definecolor{asuMaroon}{HTML}{8C1D40}
        \definecolor{asuGold}{HTML}{FFC627}
        \setbeamercolor{frametitle}{bg=asuMaroon,fg=white}
        \setbeamercolor{title}{fg=white}
        \setbeamercolor{block title}{bg=asuMaroon,fg=white}
        \setbeamercolor{alerted text}{fg=asuMaroon}
        \setbeamertemplate{footline}[frame number]
        \setbeamerfont{frametitle}{size=\large,series=\bfseries}
        \usepackage{booktabs}
execute:
  echo: false
  warning: false
  message: false
---

# Part 1: CRS, Spatial Joins & Buffer Analysis

## Module Overview

> "Your economic hardship index measures *who lives* in a place. This module adds *what is near* a place."

In Labs 1–2 you built the index entirely from census variables. This module teaches you to combine census data with **external point data** — EPA facilities, grocery stores, transit stops — using spatial operations to test environmental justice hypotheses.

**What you will learn:**

| Topic | Key Tool |
|-------|----------|
| Coordinate Reference Systems | `st_crs()`, `st_transform()` |
| Point-in-polygon spatial joins | `st_join()` |
| Buffer analysis | `st_buffer()` |
| Buffer sensitivity (UGCoP) | Kwan (2012) |
| Environmental justice framework | Maantay, Pastor et al. |

## Coordinate Reference Systems

Before combining spatial datasets, they must share the same **CRS**.

**Geographic CRS (EPSG:4326 — WGS84)**

- Latitude/longitude in degrees
- Global — good for web maps and data exchange
- Distances distorted (1 degree ≠ constant length)
- Default for most downloaded data

**Projected CRS (EPSG:2868 — Arizona Central)**

- Flat surface with linear units (US survey feet)
- Accurate distances and areas in Maricopa County
- Required for `st_buffer()` to produce meaningful distances

::: {.callout-warning}
`st_buffer(data, dist = 1000)` on EPSG:4326 creates a buffer of 1000 **degrees** — roughly 100,000 km. Always project first.
:::

## CRS in R

```{r}
#| eval: false
#| echo: true
library(sf)

# Check CRS
st_crs(cbg_data)
st_crs(facilities_sf)

# Project to Arizona Central (US survey feet) before spatial ops
cbg_proj        <- st_transform(cbg_data, crs = 2868)
facilities_proj <- st_transform(facilities_sf, crs = 2868)

# Fix minor topology errors before joins
cbg_proj <- st_make_valid(cbg_proj)

# Convert back to WGS84 for web maps
cbg_wgs84 <- st_transform(cbg_proj, crs = 4326)
```

**Rule:** Project → spatial operations → convert back for visualization.

## Spatial Joins: st_join()

A spatial join attaches attributes from one layer to another based on **geographic relationship** — not a shared column.

**Point-in-polygon (most common in community analytics):**

```{r}
#| eval: false
#| echo: true
# How many EPA facilities fall within each block group?
cbg_with_fac <- st_join(cbg_proj, facilities_proj,
                        join = st_intersects)

# Count facilities per CBG
facility_counts <- cbg_with_fac |>
  st_drop_geometry() |>
  group_by(GEOID) |>
  summarise(facility_count = sum(!is.na(name)))

cbg_proj <- cbg_proj |>
  left_join(facility_counts, by = "GEOID") |>
  mutate(facility_count = replace_na(facility_count, 0))
```

## Buffer Analysis: st_buffer()

Buffers create a **proximity zone** around a spatial feature.

```{r}
#| eval: false
#| echo: true
# 1 mile = 5,280 feet (EPSG:2868 uses US survey feet)
facility_buffers <- st_buffer(facilities_proj, dist = 5280)

# Dissolve overlapping buffers into one combined zone
buffer_union <- st_union(facility_buffers)

# Which CBGs intersect the buffer? (binary: 0/1)
cbg_proj$near_facility <- as.integer(
  lengths(st_intersects(cbg_proj, buffer_union)) > 0
)
```

**Common uses:**

- Block groups within 1 mile of a hazardous facility
- Grocery stores within 0.5 miles of each neighborhood centroid
- Hospital service area coverage

---

# Part 2: The UGCoP

## What Is UGCoP?

> *Kwan, M.P. (2012). The uncertain geographic context problem. Annals of the AAG, 102(5), 958–968.*
> doi:10.1080/00045608.2012.687349

**Core argument:**

> "The actual geographic context that influences people's behavior and health is **uncertain** and cannot be directly observed."

In practice:

- The researcher must choose a buffer size: 0.25 mi? 0.5 mi? 1 mi? 2 mi?
- There is rarely a principled theoretical basis for this choice
- Different buffer sizes change which units are classified as "exposed"
- Different classifications produce different statistical results

::: {.callout-important}
Buffer distance choice is a methodological decision — not a neutral technical step.
:::

## UGCoP vs. MAUP

| | **MAUP** | **UGCoP** |
|--|----------|-----------|
| **What varies** | Polygon boundaries | Buffer radius |
| **Mechanism** | Aggregation changes statistics | Exposure zone changes who is exposed |
| **Classic example** | County vs. tract vs. block group | 0.5 mi vs. 1 mi vs. 2 mi |
| **Named by** | Openshaw & Taylor (1979) | Kwan (2012) |
| **Applies to** | Polygon analysis | Proximity analysis |

Both MAUP and UGCoP share one insight: **geographic context is a methodological choice**, and that choice shapes results.

## Buffer Sensitivity in Practice

Three block groups at different distances from an EPA facility:

\begin{center}
\small
\begin{tabular}{lccc}
\toprule
\textbf{Buffer} & \textbf{A (0.3 mi)} & \textbf{B (0.7 mi)} & \textbf{C (1.1 mi)} \\
\midrule
0.5 mi & \textcolor{red}{Exposed} & --- & --- \\
1.0 mi & \textcolor{red}{Exposed} & \textcolor{red}{Exposed} & --- \\
2.0 mi & \textcolor{red}{Exposed} & \textcolor{red}{Exposed} & \textcolor{red}{Exposed} \\
\bottomrule
\end{tabular}
\end{center}

Same block groups, **three different classifications** — this is UGCoP.

**Lab 3, Q2:** Change distance bands and compare. This tests whether conclusions are threshold-sensitive.

## Best Practices for Buffer Analysis

How to address UGCoP:

1. **Use a theoretically grounded distance** — air dispersion models, walkability studies, and physical hazard decay rates each suggest specific distances
2. **Report results at multiple distances** — if conclusions hold at 0.5, 1.0, and 1.5 miles, the result is robust; if only one distance produces findings, flag the sensitivity
3. **Match distances from comparable studies** — Maantay (2002) and EPA EJScreen use facility-specific distances; consistency aids comparability
4. **Be explicit in reporting** — state the buffer distance, justify the choice, acknowledge sensitivity

::: {.callout-tip}
A result that holds across multiple buffer distances is far more credible than one that only appears at a single threshold.
:::

---

# Part 3: Environmental Justice

## The Environmental Justice Framework

EJ research asks: are environmental burdens disproportionately concentrated in low-income and minority communities?

**The three-part analysis:**

| Component | Data Source | Tool |
|-----------|-------------|------|
| **Demographics** — who lives here | ACS census variables | `tidycensus` |
| **Hazards** — what is near here | EPA TRI, Superfund listings | Point data |
| **Spatial integration** — do they overlap? | Combined layers | `st_join()`, `st_buffer()` |

Lab 3 operationalizes this by combining 5 census variables measuring socioeconomic hardship with spatial proximity measures to environmental burden locations.

## Pastor, Sadd & Hipp (2001)

> *Which came first? Toxic facilities, minority move-in, and environmental justice. Journal of Urban Affairs, 23(1), 1–21.*

**The central EJ research question:**

- Were hazardous facilities **sited** in minority neighborhoods? *(disproportionate siting)*
- Or did minorities **move in** after facilities were built? *(market dynamics)*

**Key findings:**

- Used longitudinal data for LA County census tracts (1970–1990)
- Applied buffer-based proximity measures to track siting over time
- Minority and low-income communities had disproportionate exposure **before** move-in
- **Supports disproportionate siting hypothesis** over residential sorting
- Uses point-in-polygon joins + buffers — the same tools in Lab 3

## UGCoP in EJ Research

Maantay (2002) and Mohai & Saha (2015) both document how distance threshold choice affects EJ findings:

- **Results can flip across distances** — a community appears "exposed" at 1 mile but not at 0.5 miles, or vice versa
- **No single distance is universally correct** — air, water, and noise hazards have different physical decay rates
- **Smaller buffers concentrate findings** — a 0.25-mile buffer identifies only immediately adjacent communities
- **Policy consequences** — if a 1-mile buffer determines EJ remediation funding eligibility, the threshold choice has direct resource allocation implications

::: {.callout-warning}
**Lab 3, Q3** asks you to connect buffer sensitivity to UGCoP and cite Maantay (2002) and Mohai & Saha (2015). This is the theoretical payoff of Q2's empirical test.
:::

## Module 3 Summary

**Key takeaways:**

- **CRS before everything.** Mismatched coordinate systems produce wrong results. Always check and transform before spatial operations.
- **Spatial joins count by location.** `st_join()` attaches attributes using geographic relationship — no shared column needed.
- **Buffers define exposure.** `st_buffer()` creates proximity zones. The distance threshold is consequential.
- **UGCoP: buffers have a problem too.** Kwan (2012): different distances produce different classifications. Test multiple distances and report sensitivity.
- **EJ connects burden to hardship.** Spatial integration reveals whether environmental hazards compound census-measured disadvantage in the same communities.

**Coming up in Module 4:** Spatial Autocorrelation — Moran's I, LISA hot spots, and whether hardship clusters spatially across Maricopa County.
