---
title: "Lab 01 — Solutions: Measurement, Indexing & Multi-Scale Mapping"
subtitle: "PAF 516 | M1 Assignment"
date: today
format:
  html:
    theme: readable
    highlight-style: tango
    toc: true
    self-contained: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
  fig-width: 10
---


```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set( echo = TRUE, message=F, warning=F, fig.width=10 )
options( tigris_use_cache = TRUE )
```


```{r}
#| label: packages
# ── Packages ─────────────────────────────────────────
# Load all required packages for data, mapping, and reliability analysis
library( tidyverse )
library( tidycensus )
library( psych )       # For Cronbach's alpha
library( corrplot )    # For correlation matrix visualization
library( viridis )     # For color scales on maps
library( ggplot2 )
library( patchwork )   # For combining multi-panel figures
```

```{r}
#| label: api-key
# Uncomment and add your Census API key
# census_api_key("YOUR KEY GOES HERE")
```


# ── Step 1: Retrieve Census Variables ──────────────────

We build an **economic hardship index** using 5 census variables at the tract level for Maricopa County, AZ. Each variable captures a different dimension of economic disadvantage.

```{r}
#| label: browse-variables
#| eval: false

# Browse available ACS variables to find relevant indicators
vars_available <- load_variables( 2023, "acs5", cache = TRUE )
# View(vars_available)  # Use this interactively to search
```

```{r}
#| label: step.01
#| results: hide

# Define our indicator variables for an "economic hardship" construct
# Each variable name is given a readable alias
hardship_vars <- c(
  pct_poverty    = "B17001_002",   # Population below poverty level
  pop_total_pov  = "B17001_001",   # Total population (poverty universe) - for denominator
  median_income  = "B19013_001",   # Median household income
  pct_unemp_num  = "B23025_005",   # Unemployed population
  pct_unemp_den  = "B23025_002",   # In labor force (denominator for unemployment)
  pct_no_hs_num  = "B15003_002",   # Population with less than high school
  pct_no_hs_den  = "B15003_001",   # Total population 25+ (denominator for education)
  pct_no_ins_num = "B27010_017",   # Population without health insurance (18-34)
  pct_no_ins_den = "B27010_001"    # Total population (insurance universe)
)

# Retrieve tract-level data with geometry for mapping
CenDF <- get_acs(
  geography = "tract",
  variables = hardship_vars,
  state     = "AZ",
  county    = "Maricopa",
  year      = 2023,
  survey    = "acs5",
  geometry  = TRUE
)
```

**Common error note:** If you get an API key error, make sure you have run `census_api_key("YOUR_KEY", install = TRUE)` at least once and restarted R.


# ── Step 2: Select and Justify Variables ───────────────

We now reshape the data from long to wide format, compute percentages from numerator/denominator pairs, and standardize all variables to z-scores.

```{r}
#| label: step.02

# Reshape from long to wide format using pivot_wider
CenDF_wide <- CenDF %>%
  select( -moe ) %>%
  pivot_wider( names_from = variable, values_from = estimate )

# Compute derived percentage variables from numerator/denominator pairs
CenDF_wide <- CenDF_wide %>%
  mutate(
    # Poverty rate: people below poverty / poverty universe
    poverty_rate = pct_poverty / pop_total_pov,
    # Unemployment rate: unemployed / in labor force
    unemp_rate   = pct_unemp_num / pct_unemp_den,
    # Percent without high school diploma
    no_hs_rate   = pct_no_hs_num / pct_no_hs_den,
    # Percent without health insurance
    no_ins_rate  = pct_no_ins_num / pct_no_ins_den
  )

# Select just the indicator variables for our index
# Note: median_income is REVERSE CODED — high income = LOW hardship
index_vars <- CenDF_wide %>%
  as.data.frame() %>%  # drop sf geometry for psych::alpha
  select( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income ) %>%
  drop_na()  # alpha() requires complete cases

# Standardize all variables to z-scores so they are on a common scale
index_z <- index_vars %>%
  mutate( across( everything(), ~ as.numeric( scale(.) ) ) )

# Reverse-code median_income: high income should contribute LOW hardship
# Flipping the sign aligns its direction with the other items
index_z <- index_z %>%
  mutate( median_income = -1 * median_income )

# Verify: all variables should now be positively correlated
# (higher values = more hardship for all items)
```

**Variable justification:**

| Variable | Census Code | Why Included |
|----------|------------|--------------|
| Poverty rate | B17001_002 / B17001_001 | Direct measure of economic deprivation |
| Unemployment rate | B23025_005 / B23025_002 | Reflects labor market hardship |
| No high school diploma | B15003_002 / B15003_001 | Education is a strong predictor of economic outcomes |
| No health insurance | B27010_017 / B27010_001 | Lack of insurance reflects economic vulnerability |
| Median household income | B19013_001 | Core income measure (reverse coded) |


# ── Step 3: Assess Reliability with Cronbach's Alpha ──

```{r}
#| label: step.03

# Calculate Cronbach's alpha to assess whether our 5 items
# consistently measure the same underlying construct
alpha_result <- psych::alpha( index_z )

# Print the main results
print( alpha_result )
```

**Interpreting the output:**

- The overall `raw_alpha` is our Cronbach's alpha value
- The "Reliability if an item is dropped" section shows what alpha would be without each item
- If dropping an item *increases* alpha, that item may not fit the construct well

```{r}
#| label: alpha-summary

# Extract the key alpha value for inline reporting
alpha_value <- alpha_result$total$raw_alpha
cat( "Cronbach's alpha:", round( alpha_value, 3 ), "\n" )
```


# ── Step 4: Build the Index and Visualize ──────────────

```{r}
#| label: step.04a

# Correlation matrix of indicator variables
cor_matrix <- cor( index_z, use = "complete.obs" )

# Visualize the correlation matrix
corrplot(
  cor_matrix,
  method  = "color",
  type    = "upper",
  addCoef.col = "black",
  tl.col  = "black",
  tl.srt  = 45,
  title   = "Inter-Item Correlation Matrix: Economic Hardship Index",
  mar     = c(0, 0, 2, 0)
)
```

All items should show positive correlations (after reverse-coding income), confirming they measure a common construct.

```{r}
#| label: step.04b

# Create the composite index as the row mean of z-scores
# This gives equal weight to each indicator
index_z$hardship_index <- rowMeans( index_z, na.rm = TRUE )

# Histogram of the hardship index
ggplot( index_z, aes( x = hardship_index ) ) +
  geom_histogram( bins = 40, fill = "#8c1d40", color = "white", alpha = 0.8 ) +
  geom_vline( xintercept = mean( index_z$hardship_index, na.rm = TRUE ),
              linetype = "dashed", color = "#ffc627", linewidth = 1 ) +
  theme_minimal() +
  labs(
    title    = "Distribution of Economic Hardship Index",
    subtitle = "Maricopa County Census Tracts (ACS 2023)",
    x        = "Hardship Index (z-score scale)",
    y        = "Number of Tracts",
    caption  = "Dashed line = mean"
  )
```

```{r}
#| label: step.04c

# Add the index back to the spatial data for mapping
# Match rows using the complete-case subset
CenDF_complete <- CenDF_wide %>%
  drop_na( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income )

CenDF_complete$hardship_index <- index_z$hardship_index

# Map the hardship index across tracts
map_tract <- ggplot( CenDF_complete ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis( option = "inferno", direction = -1,
                      name = "Hardship\nIndex" ) +
  theme_void() +
  labs(
    title    = "Economic Hardship Index by Census Tract",
    subtitle = "Maricopa County, AZ — ACS 5-Year (2023)",
    caption  = "Higher values = greater economic hardship"
  )

map_tract
```


# ── Step 4b: National County-Level Map ─────────────────

We now pull the same 5 hardship variables at the **county level** for the entire US, compute the index using the same method, and map national patterns of economic hardship.

```{r}
#| label: step.04b-pull
#| results: hide

# Pull county-level data for the entire US
# Same 5 variables, different geography
county_df <- get_acs(
  geography = "county",
  variables = hardship_vars,
  year      = 2023,
  survey    = "acs5",
  geometry  = TRUE
)
```

```{r}
#| label: step.04b-index

# Reshape county data from long to wide
county_wide <- county_df %>%
  select( -moe ) %>%
  pivot_wider( names_from = variable, values_from = estimate )

# Compute the same rate variables as the tract level
county_wide <- county_wide %>%
  mutate(
    poverty_rate = pct_poverty / pop_total_pov,
    unemp_rate   = pct_unemp_num / pct_unemp_den,
    no_hs_rate   = pct_no_hs_num / pct_no_hs_den,
    no_ins_rate  = pct_no_ins_num / pct_no_ins_den
  )

# Z-score standardization and reverse-code income
county_index <- county_wide %>%
  as.data.frame() %>%
  select( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income ) %>%
  drop_na() %>%
  mutate( across( everything(), ~ as.numeric( scale(.) ) ) ) %>%
  mutate( median_income = -1 * median_income )

# Composite index: row mean of z-scores
county_index$hardship_index <- rowMeans( county_index, na.rm = TRUE )

# Attach index back to spatial data
county_complete <- county_wide %>%
  drop_na( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income )
county_complete$hardship_index <- county_index$hardship_index
```

```{r}
#| label: step.04b-map

# National county-level hardship map
# shift_geometry not used here; simple contiguous-US view
map_county <- ggplot( county_complete ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis( option = "inferno", direction = -1,
                      name = "Hardship\nIndex" ) +
  theme_void() +
  labs(
    title   = "Economic Hardship Index by US County",
    subtitle = "ACS 5-Year (2023)",
    caption = "Higher values = greater economic hardship"
  )

map_county
```


# ── Step 4c: Maricopa CBG-Level Map ───────────────────

We zoom in further to the **block group** level within Maricopa County. Block groups are the smallest geography for which ACS data are available, revealing fine-grained variation within tracts.

```{r}
#| label: step.04c-pull
#| results: hide

# Pull block group data for Maricopa County
# This is the finest spatial resolution available in the ACS
cbg_df <- get_acs(
  geography = "block group",
  variables = hardship_vars,
  state     = "AZ",
  county    = "Maricopa",
  year      = 2023,
  survey    = "acs5",
  geometry  = TRUE
)
```

```{r}
#| label: step.04c-index

# Reshape and compute rates — same pipeline as tracts and counties
cbg_wide <- cbg_df %>%
  select( -moe ) %>%
  pivot_wider( names_from = variable, values_from = estimate )

cbg_wide <- cbg_wide %>%
  mutate(
    poverty_rate = pct_poverty / pop_total_pov,
    unemp_rate   = pct_unemp_num / pct_unemp_den,
    no_hs_rate   = pct_no_hs_num / pct_no_hs_den,
    no_ins_rate  = pct_no_ins_num / pct_no_ins_den
  )

# Z-score and reverse-code income
cbg_index <- cbg_wide %>%
  as.data.frame() %>%
  select( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income ) %>%
  drop_na() %>%
  mutate( across( everything(), ~ as.numeric( scale(.) ) ) ) %>%
  mutate( median_income = -1 * median_income )

cbg_index$hardship_index <- rowMeans( cbg_index, na.rm = TRUE )

# Attach index to spatial data
cbg_complete <- cbg_wide %>%
  drop_na( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income )
cbg_complete$hardship_index <- cbg_index$hardship_index
```

```{r}
#| label: step.04c-map

# Block-group-level hardship map for Maricopa County
map_cbg <- ggplot( cbg_complete ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis( option = "inferno", direction = -1,
                      name = "Hardship\nIndex" ) +
  theme_void() +
  labs(
    title    = "Economic Hardship Index by Block Group",
    subtitle = "Maricopa County, AZ — ACS 5-Year (2023)",
    caption  = "Higher values = greater economic hardship"
  )

map_cbg
```


# ── Step 4d: Three-Scale Comparison ───────────────────

Using `patchwork`, we place all three maps side by side to compare how spatial patterns of hardship appear at different geographic scales. This illustrates the **Modifiable Areal Unit Problem (MAUP)** — the same phenomenon looks different depending on the unit of analysis.

```{r}
#| label: step.04d-patchwork
#| fig-width: 16
#| fig-height: 6

# Build compact versions of each map for the 3-panel comparison
# Remove subtitles/captions to keep the panel clean
p_county <- ggplot( county_complete ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis( option = "inferno", direction = -1, name = "Index" ) +
  theme_void() +
  labs( title = "County (US)" )

p_tract <- ggplot( CenDF_complete ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis( option = "inferno", direction = -1, name = "Index" ) +
  theme_void() +
  labs( title = "Tract (Maricopa)" )

p_cbg <- ggplot( cbg_complete ) +
  geom_sf( aes( fill = hardship_index ), color = NA ) +
  scale_fill_viridis( option = "inferno", direction = -1, name = "Index" ) +
  theme_void() +
  labs( title = "Block Group (Maricopa)" )

# patchwork: combine with + and collect shared legend
p_county + p_tract + p_cbg +
  plot_annotation(
    title   = "Economic Hardship Across Three Geographic Scales",
    caption = "ACS 5-Year (2023) | Higher values = greater hardship"
  )
```


# ── Questions ──────────────────────────────────────────

### Q1 — Interpret

**What is the Cronbach's alpha? Is it acceptable (>0.7)? Looking at the 3-scale maps, where is hardship concentrated nationally? How does the spatial pattern change as you zoom from counties to tracts to CBGs?**

- **Alpha**: The Cronbach's alpha for our 5-item economic hardship index is `r round( alpha_value, 2 )`. This is `r ifelse( alpha_value > 0.7, "above the 0.7 threshold, indicating acceptable internal consistency", "below the 0.7 threshold, suggesting the items may not consistently measure the same construct" )`.

- **National patterns**: At the county level, high-hardship areas are concentrated in the rural South (Mississippi Delta, Appalachia, South Texas border counties) and parts of tribal nations in the West. The county map shows broad regional clusters but smooths over within-county variation.

- **Scale comparison**: As we zoom from counties to tracts to block groups, the spatial pattern becomes more granular. At the tract level within Maricopa County, we see clear hardship corridors — particularly in south Phoenix and parts of central Mesa. At the block-group level, even more fine-grained variation emerges: adjacent block groups within the same tract can differ substantially, revealing pockets of hardship (or prosperity) that tract-level data masks. This is a practical illustration of the **Modifiable Areal Unit Problem (MAUP)** — the choice of geographic unit shapes what patterns we observe.


### Q2 — Modify (FULL CODE AND OUTPUT REQUIRED)

**Add a 6th variable to the hardship index — percent receiving SNAP benefits (B22010_002 / B22010_001). Re-pull data at the tract level, recompute Cronbach's alpha with 6 variables, rebuild the index, and re-map at the tract level.**

```{r}
#| label: q2-pull
#| results: hide

# Define expanded variable list with SNAP benefits added
# B22010_002 = households receiving SNAP; B22010_001 = total households
hardship_vars_6 <- c(
  hardship_vars,
  snap_num = "B22010_002",   # Households receiving SNAP/food stamps
  snap_den = "B22010_001"    # Total households (SNAP universe)
)

# Re-pull tract data with the expanded variable set
CenDF_6 <- get_acs(
  geography = "tract",
  variables = hardship_vars_6,
  state     = "AZ",
  county    = "Maricopa",
  year      = 2023,
  survey    = "acs5",
  geometry  = TRUE
)
```

```{r}
#| label: q2-index

# Reshape and compute all 6 rate variables
CenDF_6_wide <- CenDF_6 %>%
  select( -moe ) %>%
  pivot_wider( names_from = variable, values_from = estimate )

CenDF_6_wide <- CenDF_6_wide %>%
  mutate(
    poverty_rate = pct_poverty / pop_total_pov,
    unemp_rate   = pct_unemp_num / pct_unemp_den,
    no_hs_rate   = pct_no_hs_num / pct_no_hs_den,
    no_ins_rate  = pct_no_ins_num / pct_no_ins_den,
    snap_rate    = snap_num / snap_den       # NEW: percent receiving SNAP
  )

# Select all 6 indicators, z-score, and reverse-code income
index_vars_6 <- CenDF_6_wide %>%
  as.data.frame() %>%
  select( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income, snap_rate ) %>%
  drop_na()

index_z_6 <- index_vars_6 %>%
  mutate( across( everything(), ~ as.numeric( scale(.) ) ) ) %>%
  mutate( median_income = -1 * median_income )

# Cronbach's alpha with 6 items
alpha_result_6 <- psych::alpha( index_z_6 )
alpha_value_6  <- alpha_result_6$total$raw_alpha

cat( "6-variable Cronbach's alpha:", round( alpha_value_6, 3 ), "\n" )
print( alpha_result_6 )
```

```{r}
#| label: q2-map

# Build the 6-variable composite index
index_z_6$hardship_index_6 <- rowMeans( index_z_6, na.rm = TRUE )

# Attach to spatial data
CenDF_6_complete <- CenDF_6_wide %>%
  drop_na( poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, median_income, snap_rate )
CenDF_6_complete$hardship_index_6 <- index_z_6$hardship_index_6

# Map the 6-variable hardship index
ggplot( CenDF_6_complete ) +
  geom_sf( aes( fill = hardship_index_6 ), color = NA ) +
  scale_fill_viridis( option = "inferno", direction = -1,
                      name = "Hardship\nIndex (6-var)" ) +
  theme_void() +
  labs(
    title    = "Economic Hardship Index (6 Variables) by Census Tract",
    subtitle = "Maricopa County, AZ — ACS 5-Year (2023)",
    caption  = "Includes SNAP benefits as 6th indicator"
  )
```


### Q3 — Interpret Modified

**Did adding SNAP improve or weaken Cronbach's alpha? Compare the 5-variable map to the 6-variable map. Should you keep the 6th variable? Why or why not?**

- **Alpha comparison**: The 5-variable alpha was `r round( alpha_value, 3 )` and the 6-variable alpha is `r round( alpha_value_6, 3 )`. `r ifelse( alpha_value_6 >= alpha_value, "Adding SNAP increased (or maintained) alpha, indicating that SNAP receipt is consistent with the other hardship indicators.", "Adding SNAP slightly decreased alpha, suggesting it introduces some noise, though the change may be minor." )`

- **Map comparison**: The spatial patterns in the 5-variable and 6-variable maps are very similar — high-hardship tracts in south Phoenix and surrounding areas remain high-hardship under both specifications. SNAP receipt is strongly correlated with poverty, so it reinforces rather than changes the geographic pattern.

- **Recommendation**: `r ifelse( alpha_value_6 >= alpha_value, "SNAP improves reliability and adds conceptual coverage (program participation as a hardship dimension), so it is worth keeping.", "Although alpha decreased slightly, SNAP adds a conceptually distinct dimension — actual program uptake rather than demographic characteristics. If the decrease is small (within ~0.01-0.02), retaining it is reasonable for broader construct validity." )` The decision depends on whether you prioritize parsimony (5 variables) or construct breadth (6 variables).
