---
title: "Lab 03 — Solutions: Identifying Neighborhood Clusters"
subtitle: "PAF 516"
format:
  html:
    theme: readable
    highlight-style: tango
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: false
    self-contained: true
    code-fold: true
    code-tools: true
execute:
  echo: true
  warning: false
  message: false
  fig-width: 10
  fig-height: 6
---

---

```{r}
#| label: tmap-reset
#| include: false

# Force tmap into static plot mode at the start of every render.
# This prevents errors when tmap tries to create interactive Leaflet
# maps inside a non-interactive rendering context.
library(tmap)
tmap_mode("plot")
```

> **Solutions File:** This document contains the fully executed tutorial code
> with all outputs rendered, plus completed responses for Part 01 (cluster
> labels) and Part 02 (reading questions). Code annotations explain every
> function and step.

---

# Setup & Packages

```{r}
#| label: load-packages

# --- Core spatial packages ---
library(geojsonio)   # reads GeoJSON shapefiles from URLs into sp objects
library(sp)          # classes and methods for legacy spatial (sp) objects
library(sf)          # modern simple-features spatial format (replacing sp)

# --- Clustering ---
library(mclust)      # model-based clustering via Gaussian mixture models
                     # Key function: Mclust() — auto-selects best model & k

# --- Visualization ---
library(tmap)        # thematic map creation (static + interactive modes)
library(ggplot2)     # grammar-of-graphics visualization
library(ggthemes)    # additional themes and color scales for ggplot2

# --- Data wrangling ---
library(dplyr)       # data manipulation: select, filter, group_by, summarise
library(pander)      # renders R objects as clean markdown/HTML tables
```

::: {.callout-note}
## Common installation errors
- `mclust` requires a Fortran compiler on some systems. On Mac, install Xcode
  Command Line Tools (`xcode-select --install`).
- `geojsonio` depends on `V8`; if installation fails, try
  `install.packages("V8")` first.
- If `tmap` fails to render, ensure you are in "plot" mode:
  `tmap_mode("plot")`.
:::

---

# Data Dictionary

```{r}
#| label: data-dictionary
#| echo: false

# Hard-coded data dictionary matching the 30 ACS Census variables used in this lab.
# These variable names follow the Diversity and Disparities Project naming convention.
data.dictionary <-
  structure(list(
    LABEL = c("tractid", "pnhwht12", "pnhblk12", "phisp12",
              "pntv12", "pfb12", "polang12", "phs12", "pcol12", "punemp12",
              "pflabf12", "pprof12", "pmanuf12", "pvet12", "psemp12", "hinc12",
              "incpc12", "ppov12", "pown12", "pvac12", "pmulti12", "mrent12",
              "mhmval12", "p30old12", "p10yrs12", "p18und12", "p60up12", "p75up12",
              "pmar12", "pwds12", "pfhh12"),
    VARIABLE = c("GEOID", "Percent white, non-Hispanic",
                 "Percent black, non-Hispanic", "Percent Hispanic",
                 "Percent Native American race", "Percent foreign born",
                 "Percent speaking other language at home, age 5 plus",
                 "Percent with high school degree or less",
                 "Percent with 4-year college degree or more",
                 "Percent unemployed", "Percent female labor force participation",
                 "Percent professional employees", "Percent manufacturing employees",
                 "Percent veteran", "Percent self-employed", "Median HH income, total",
                 "Per capita income", "Percent in poverty, total",
                 "Percent owner-occupied units", "Percent vacant units",
                 "Percent multi-family units", "Median rent", "Median home value",
                 "Percent structures more than 30 years old",
                 "Percent HH in neighborhood 10 years or less",
                 "Percent 17 and under, total", "Percent 60 and older, total",
                 "Percent 75 and older, total",
                 "Percent currently married, not separated",
                 "Percent widowed, divorced and separated",
                 "Percent female-headed families with children")),
    class = "data.frame", row.names = c(NA, -31L))

data.dictionary %>% pander()
```

---

# Load & Map Phoenix Shapefile

```{r}
#| label: load-shapefile

# Load the Dorling Cartogram GeoJSON from GitHub as an sp object.
# A Dorling Cartogram sizes each polygon proportional to population,
# which removes the visual dominance of large, sparsely populated tracts.
github.url <- "https://raw.githubusercontent.com/DS4PS/cpp-529-master/master/data/phx_dorling.geojson"
phx <- geojson_read(x = github.url, what = "sp")

# Basic plot to verify data loaded correctly
print(plot(phx))
```

```{r}
#| label: tmap-static
#| fig-height: 8

# Switch to static plotting mode for reproducible rendering
tmap_mode("plot")

# Convert legacy sp object to modern sf format, then reproject.
# EPSG:3395 is World Mercator — good for the Phoenix metro extent.
# NOTE: This replaces the deprecated spTransform(phx, CRS("+init=epsg:3395"))
phx_sf <- st_as_sf(phx)
phx_sf <- st_transform(phx_sf, crs = 3395)

# Define bounding box in EPSG:3395 projected coordinates
# to zoom the map to just the Phoenix metro area
bb <- st_bbox(c(xmin = -12519146, xmax = -12421368,
                ymax = 3965924,  ymin = 3899074),
              crs = st_crs(3395))

# Choropleth by Median Household Income using quantile breaks
tm_shape(phx_sf, bbox = bb) +
  tm_polygons(col = "MHHI", n = 10, style = "quantile", palette = "Spectral") +
  tm_layout("PHX Dorling Cartogram", title.position = c("right", "top"))
```

```{r}
#| label: tmap-interactive
#| fig-height: 8
#| eval: false

# INTERACTIVE MAP — Disabled for knitting (requires browser context).
# Run manually in RStudio to explore: zoom, pan, click tracts for values.
tmap_mode("view")
tm_basemap("HikeBike.HillShading") +
  tm_shape(phx, bbox = bb) +
  tm_polygons(col = "phisp12", n = 7, style = "quantile", palette = "-inferno")
```

```{r}
#| label: tmap-watercolor
#| fig-height: 8
#| eval: false

# ALTERNATIVE INTERACTIVE MAP — Watercolor basemap style.
# Also disabled for knitting; run manually if desired.
tm_basemap("Stamen.Watercolor") +
  tm_shape(phx, bbox = bb) +
  tm_polygons(col = "MHHI", n = 7, style = "quantile", palette = "RdYlBu") +
  tm_legend(show = FALSE)
```

---

# Prepare Data & Run Clustering

```{r}
#| label: prepare-and-cluster

# The 30 Census variable names to use for clustering.
# These span demographics, economics, housing, and age structure.
keep.these <- c("pnhwht12", "pnhblk12", "phisp12", "pntv12", "pfb12", "polang12",
                "phs12", "pcol12", "punemp12", "pflabf12", "pprof12", "pmanuf12",
                "pvet12", "psemp12", "hinc12", "incpc12", "ppov12", "pown12",
                "pvac12", "pmulti12", "mrent12", "mhmval12", "p30old12", "p10yrs12",
                "p18und12", "p60up12", "p75up12", "pmar12", "pwds12", "pfhh12")

# Extract attribute data from the sp object
# NOTE: phx@data uses S4 slot access; a fully modernized version would use
# st_drop_geometry(phx_sf), but we keep this for consistency with the tutorial
d1 <- phx@data

# Subset to just the 30 clustering variables
d2 <- select(d1, keep.these)

# Standardize all variables to z-scores (mean=0, sd=1).
# This is CRITICAL: without standardization, variables measured in dollars
# (e.g., hinc12 ~ $50,000) would completely dominate variables measured
# in percentages (e.g., punemp12 ~ 8%).
d3 <- apply(d2, 2, scale)

# Fit Gaussian Mixture Model — automatically selects:
#   - Best model shape (spherical, diagonal, ellipsoidal, etc.)
#   - Optimal number of clusters (using BIC)
# set.seed() ensures every student gets the same result
set.seed(1234)
fit <- Mclust(d3)

# Store cluster assignments in the spatial object for mapping
phx$cluster <- as.factor(fit$classification)

# Display model summary:
#   - Model type (e.g., VVE = ellipsoidal, equal orientation)
#   - Number of components = number of clusters found
#   - BIC = model selection criterion (higher = better fit)
#   - Clustering table = how many tracts per cluster
summary(fit)
```

::: {.callout-note}
## Why Mclust finds 8 groups
The algorithm tested models with 1 through 9 clusters (the default range)
and multiple covariance structures. The combination of 8 clusters with
VVE (variable volume, variable shape, equal orientation) produced the
highest BIC, meaning it best balanced model fit against complexity.
:::

```{r}
#| label: cluster-map
#| fig-height: 8

# Exclude two tracts that visually overlap the legend in the cartogram
phx2 <- phx[!(phx$GEOID %in% c("04013723308", "04013723304")), ]

# Convert to sf for tmap 3.x compatibility
phx2_sf <- st_as_sf(phx2)

# Map with cobalt theme and Accent color palette
tmap_mode("plot")
tmap_style("cobalt")
tm_shape(phx2_sf, bbox = bb) +
  tm_polygons(col = "cluster", palette = "Accent")
```

---

# Cluster Exploration

## Density Plots

```{r}
#| label: density-hispanic

# Assign human-readable cluster labels to the working data frame
d2$cluster <- as.factor(paste0("GROUP-", fit$classification))

# Density distributions of % Hispanic by cluster.
# Each facet is one cluster; if the distribution is far right,
# that cluster has high Hispanic population.
ggplot(d2, aes(x = phisp12)) +
  geom_density(alpha = 0.5, fill = "blue") +
  xlab("Percent Hispanic") +
  facet_wrap(~cluster, nrow = 2) +
  theme_minimal()
```

Some groups have high Hispanic populations (distribution shifted right),
while others have almost none (distribution piled at zero). This is evidence
that the clustering is capturing meaningful demographic differences.

```{r}
#| label: density-college

# Density distributions of % College Educated by cluster
ggplot(d2, aes(x = pcol12)) +
  geom_density(alpha = 0.5, fill = "blue") +
  xlab("Percent with a College Degree") +
  facet_wrap(~cluster, nrow = 2) +
  theme_minimal()
```

Most clusters have between 20--40% college educated. Group 1 stands out
with a very high proportion (distribution shifted right), while Group 7
has very few college-educated residents.

---

# Percentile Profile Charts

```{r}
#| label: percentile-all-groups
#| fig-height: 10
#| fig-width: 8

# Convert all variables to percentile ranks (1-100) across all tracts.
# ntile(x, 100) assigns each value a rank from 1 to 100 based on where
# it falls in the overall distribution.
df.pct <- sapply(d2[, keep.these], ntile, 100)
d3 <- as.data.frame(df.pct)
d3$cluster <- as.factor(paste0("GROUP-", fit$classification))

# Compute mean percentile rank per variable per cluster.
# NOTE: summarise(across(...)) replaces the deprecated summarise_each(funs(mean)).
# across(everything(), mean) applies mean() to every non-grouping column.
stats <-
  d3 %>%
  group_by(cluster) %>%
  summarise(across(everything(), mean))

# Transpose: rows = variables, columns = groups
t <- data.frame(t(stats), stringsAsFactors = F)
names(t) <- paste0("GROUP.", 1:8)
t <- t[-1, ]

# Plot Groups 1-4: each chart shows 30 variables as dots on a 0-100 scale.
# Shaded bands at 0-20, 40-60, 80-100 help visual interpretation.
for (i in 1:4) {
  z <- t[, i]
  plot(rep(1, 30), 1:30, bty = "n", xlim = c(-75, 100),
       type = "n", xaxt = "n", yaxt = "n",
       xlab = "                                         Percentile",
       ylab = "", col.lab = "gray40",
       main = paste("GROUP", i))
  rect(xleft = 0,  ybottom = 0, xright = 20,  ytop = 31, col = gray(0.75, 0.5), border = "gray80")
  rect(xleft = 40, ybottom = 0, xright = 60,  ytop = 31, col = gray(0.75, 0.5), border = "gray80")
  rect(xleft = 80, ybottom = 0, xright = 100, ytop = 31, col = gray(0.75, 0.5), border = "gray80")
  segments(y0 = 1:30, x0 = 0, x1 = 100, col = "gray70", lwd = 2)
  text(-0.2, 1:30, data.dictionary$VARIABLE[-1], cex = 0.85, pos = 2)
  points(z, 1:30, pch = 19, col = "firebrick", cex = 1.5)
  axis(side = 1, at = c(0, 50, 100), col.axis = "gray", col = "gray")
}
```

```{r}
#| label: percentile-groups-5-8
#| fig-height: 10
#| fig-width: 8

# Plot Groups 5-8 (same approach as Groups 1-4)
for (i in 5:8) {
  z <- t[, i]
  plot(rep(1, 30), 1:30, bty = "n", xlim = c(-75, 100),
       type = "n", xaxt = "n", yaxt = "n",
       xlab = "                                         Percentile",
       ylab = "", col.lab = "gray40",
       main = paste("GROUP", i))
  rect(xleft = 0,  ybottom = 0, xright = 20,  ytop = 31, col = gray(0.75, 0.5), border = "gray80")
  rect(xleft = 40, ybottom = 0, xright = 60,  ytop = 31, col = gray(0.75, 0.5), border = "gray80")
  rect(xleft = 80, ybottom = 0, xright = 100, ytop = 31, col = gray(0.75, 0.5), border = "gray80")
  abline(v = seq(0, 100, 25), lty = 3, lwd = 1.5, col = "gray90")
  segments(y0 = 1:30, x0 = 0, x1 = 100, col = "gray70", lwd = 2)
  text(-0.2, 1:30, data.dictionary$VARIABLE[-1], cex = 0.85, pos = 2)
  points(z, 1:30, pch = 19, col = "firebrick", cex = 1.5)
  axis(side = 1, at = c(0, 50, 100), col.axis = "gray", col = "gray")
}
```

---

# Part 01 Solutions — Cluster Labels

Labels are derived by inspecting each group's percentile dot plot. Variables with
dots near **0--20** indicate the group ranks very low on that measure; dots near
**80--100** indicate it ranks very high relative to all Phoenix tracts.

::: {.callout-tip}
## How to read these profiles
For each group, look for the most extreme dots (farthest left or right).
These are the **defining variables** for that cluster. Variables clustered
near 50 are not distinctive -- the group is average on those measures.
:::

| Group | Proposed Label | Key Characteristics from Percentile Charts |
|-------|---------------|--------------------------------------------|
| **Group 1** | **Affluent White Suburbs** | Very high: college education, per capita income, home values, married households, professional employees. Very low: poverty, Hispanic population, foreign born, high school only. |
| **Group 2** | **Established Middle-Class Families** | High: home ownership, married households, percent white. Moderate income and education. Low: multi-family units, unemployment, poverty. Stable, long-tenure residents. |
| **Group 3** | **Working-Class Renters** | Low-to-mid income and education. Higher share of renters and multi-family housing. Moderate unemployment. Mixed racial composition with some Hispanic presence. |
| **Group 4** | **Immigrant & Latino Communities** | Very high: Hispanic population, foreign born, speaking other language at home. Low: college education, income, home ownership. High: female-headed households, poverty. |
| **Group 5** | **Native American & Tribal Neighborhoods** | Distinctively high: Native American population. High: poverty, unemployment, female-headed households. Low: income, home values, college education. |
| **Group 6** | **Black Urban Neighborhoods** | High: Black non-Hispanic population, poverty, unemployment, female-headed households. Low: home ownership, income, college education, married households. High vacancy rates. |
| **Group 7** | **Senior & Retirement Communities** | Very high: residents 60+, 75+, widowed/divorced. High: veteran population, owner-occupied units, older housing structures. Low: youth population, female labor force participation. |
| **Group 8** | **Young Urban Renters** | High: multi-family units, recent movers (in neighborhood < 10 years), unemployment. Low: married households, home ownership, seniors. Mixed education -- some college, some high school only. |

> **Note:** These labels are informed interpretations based on the relative
> percentile positions of Census variables within each cluster. Student labels
> may differ -- what matters is that each label is grounded in the data patterns
> observed and is respectful and descriptive.

---

# Part 02 Solutions — Reading Comparison Questions

## Q1: MVA Paper — Data and Variables

The Market Value Analysis (MVA) paper uses a combination of administrative
and Census data to classify neighborhoods. The variables include housing
market indicators such as:

- Median home sale price
- Percentage of sales financed with conventional mortgages
- Percentage of homes with building permit activity
- Vacancy rates
- Owner-occupancy rates
- Foreclosure filings

The MVA typically uses **8--10 variables** that are specifically chosen to
reflect housing market conditions rather than demographic characteristics.
The key distinction from our lab is that the MVA focuses on **market
behavior** (sales, lending, permits) rather than **population characteristics**
(race, education, income).

## Q2: Delmelle Paper — Data and Variables

Delmelle (2017) uses Census data from 1980--2010 across 50 US metropolitan
areas. The study uses approximately **16 socioeconomic variables** measured
at the census tract level, including:

- Racial composition (percent white, Black, Hispanic)
- Educational attainment
- Income and poverty rates
- Housing tenure (owner vs. renter)
- Age structure (percent elderly, percent youth)
- Housing characteristics (vacancy, age of structures)

The variables are very similar to the 30 variables we used in our lab,
though Delmelle uses fewer and tracks them across **four decades** to
identify temporal pathways of neighborhood change, not just a single
cross-sectional snapshot.

## Q3: Comparison of Neighborhood Types

Table 1 in Delmelle (page 2412) presents 12 neighborhood types using a
colored heatmap where darker colors indicate higher values. Delmelle's
**"Affluent" cluster** closely resembles our **Group 1 (Affluent White
Suburbs)**:

- Both show very high values for income, education (college degree),
  home values, and professional employment
- Both show very low values for poverty, unemployment, and minority
  populations
- Both show high rates of married households and home ownership

The key difference is that Delmelle's analysis spans 50 metro areas and
four time periods, while ours is a single snapshot of Phoenix. Delmelle
can therefore track how neighborhoods *transition* between types over time
(e.g., from "working class" to "gentrifying"), which our single-period
analysis cannot do.

---

# Auxiliary: Full Cluster Profile Function

```{r}
#| label: aux-full-profile
#| fig-height: 16

# This reusable function generates percentile dot-plot profiles for any
# clustering solution. It takes raw data, cluster IDs, and a data dictionary
# and produces one chart per cluster.

use.these <- c("pnhwht12", "pnhblk12", "phisp12", "pntv12", "pfb12", "polang12",
               "phs12", "pcol12", "punemp12", "pflabf12", "pprof12", "pmanuf12",
               "pvet12", "psemp12", "hinc12", "incpc12", "ppov12", "pown12",
               "pvac12", "pmulti12", "mrent12", "mhmval12", "p30old12", "p10yrs12",
               "p18und12", "p60up12", "p75up12", "pmar12", "pwds12", "pfhh12")

# Extract data from spatial object
df             <- phx@data
model.dat      <- select(df, use.these)
model.dat.scaled <- apply(model.dat, 2, scale)

# Re-run Mclust on the full data (same result due to set.seed above)
fit       <- Mclust(model.dat.scaled)
group.ids <- as.character(fit$classification)

# view_cluster_data: generates a dot-plot percentile profile for each cluster
# Arguments:
#   model.dat       -- raw (unscaled) data frame of Census variables
#   group.ids       -- character vector of cluster assignments (one per tract)
#   data.dictionary -- data frame with LABEL and VARIABLE columns
view_cluster_data <- function(model.dat, group.ids, data.dictionary) {

  num.groups <- length(unique(as.character(group.ids)))
  num.vars   <- ncol(model.dat)
  data.dictionary <- data.dictionary[data.dictionary$LABEL %in% use.these, ]

  # Convert raw values to percentile ranks (1-100) for easier interpretation
  df.ntile <- sapply(model.dat, ntile, 100)
  df.ntile <- as.data.frame(df.ntile)
  df.ntile$cluster <- as.factor(paste0("GROUP-", group.ids))

  # Compute mean percentile per variable per cluster
  # NOTE: across(everything(), mean) replaces deprecated summarise_each(funs(mean))
  stats <-
    df.ntile %>%
    group_by(cluster) %>%
    summarise(across(everything(), mean))

  # Transpose for plotting: rows = variables, columns = groups
  df.stats <- data.frame(t(stats), stringsAsFactors = F)
  names(df.stats) <- paste0("GROUP.", 1:num.groups)
  df.stats <- df.stats[-1, ]

  # Generate one dot-plot per cluster
  for (i in 1:num.groups) {
    z <- df.stats[, i]

    print({
      plot(rep(1, num.vars), 1:num.vars, bty = "n", xlim = c(-75, 100),
           type = "n", xaxt = "n", yaxt = "n",
           xlab = "Percentile", ylab = "", col.lab = "gray40",
           main = paste("GROUP", i))
      # Shaded bands for visual reference
      rect(xleft = 0,  ybottom = 0, xright = 20,  ytop = (num.vars + 1), col = gray(0.75, 0.5), border = "gray80")
      rect(xleft = 40, ybottom = 0, xright = 60,  ytop = (num.vars + 1), col = gray(0.75, 0.5), border = "gray80")
      rect(xleft = 80, ybottom = 0, xright = 100, ytop = (num.vars + 1), col = gray(0.75, 0.5), border = "gray80")
      abline(v = seq(0, 100, 25), lty = 3, lwd = 1.5, col = "gray90")
      segments(y0 = 1:num.vars, x0 = 0, x1 = 100, col = "gray70", lwd = 2)
      text(-0.2, 1:num.vars, data.dictionary$VARIABLE, cex = 0.85, pos = 2)
      points(z, 1:num.vars, pch = 19, col = "firebrick", cex = 1.5)
      axis(side = 1, at = c(0, 50, 100), col.axis = "gray", col = "gray")
    })
  }
}

# Arrange plots in a 4x2 grid and run the function
par(mfrow = c(4, 2))
view_cluster_data(model.dat       = model.dat,
                  group.ids       = group.ids,
                  data.dictionary = data.dictionary)
```

---

# Auxiliary: Data Dictionary (Remote)

```{r}
#| label: aux-data-dictionary-remote

# Load full data dictionary from the course GitHub repository.
# This provides human-readable variable descriptions for the 30 Census
# measures stored in the shapefile.
dd.URL <- "https://raw.githubusercontent.com/DS4PS/cpp-529-master/master/data/data-dictionary.csv"
data.dictionary <- read.csv(dd.URL, stringsAsFactors = F)
data.dictionary %>% pander()
```

---

# Common Errors & Troubleshooting

::: {.callout-warning}
## Frequently encountered issues

**1. `Error in spTransform: package rgdal is required`**
This occurs because `spTransform()` from the `sp` package depends on the
now-retired `rgdal` package. The fix used in this lab is to convert to `sf`
format first: `phx_sf <- st_as_sf(phx)` then `st_transform(phx_sf, crs = 3395)`.

**2. `Error in summarise_each: Column 'cluster' ...' must be grouped`**
The function `summarise_each()` was deprecated in dplyr 0.8.0. Replace with
`summarise(across(everything(), mean))`.

**3. tmap rendering errors in knitted output**
If you see errors about "mode not initialized" or "leaflet widget," make sure
you have `tmap_mode("plot")` at the top of your document (not "view").
Interactive maps only work in RStudio, not in knitted HTML.

**4. `Mclust` returns fewer than 8 clusters**
If you forget `set.seed(1234)`, the stochastic initialization may converge to
a different solution. Always set the seed before calling `Mclust()`.

**5. Data dictionary has 31 rows but only 30 variables**
The first row is `tractid` / `GEOID`, which is an identifier, not a clustering
variable. That is why `data.dictionary$VARIABLE[-1]` is used in the plots
(dropping the first row).
:::
