---
title: "Lab 02 — Solutions: Cartographic Design & Multi-Scale Visual Analytics"
subtitle: "PAF 516 | M2 Assignment"
date: today
format:
  html:
    theme: readable
    highlight-style: tango
    toc: true
    self-contained: true
    number-sections: false
execute:
  echo: true
  warning: false
  message: false
  fig-width: 10
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10)

# Cache Census TIGER/Line geometries so they don't re-download every render
options(tigris_use_cache = TRUE)
```

# Overview

In Lab 1 you built a 5-variable **economic hardship index** and mapped it
across Maricopa County census tracts. That map used a single color ramp with
default classification breaks --- sensible defaults, but defaults nonetheless.

This lab asks a sharper question: **how do cartographic choices shape the
analytical story?** The same data can look dramatically different depending on
how you classify values, what you pair them with, and what geographic scale you
use. By the end of this lab you will be able to:

1. Apply four classification methods (quantile, Jenks, equal interval, standard
   deviation) and explain how each changes the visual narrative.
2. Create a **bivariate choropleth** that maps *two* variables simultaneously.
3. Build **small-multiple** maps that compare patterns across geographic scales.
4. Produce an **interactive Leaflet map** with popups, layer controls, and
   basemap options.

**New packages introduced**: `classInt`, `biscale`, `leaflet`, `ggspatial`,
`patchwork`.

---

# Required Packages

```{r}
#| label: packages

# ── Load packages ─────────────────────────────────────────────
# Core data & spatial
library(tidyverse)
library(tidycensus)
library(sf)

# Cartographic tools (new in Lab 2)
library(classInt)   # Classification algorithms (Jenks, quantile, etc.)
library(biscale)    # Bivariate choropleth palette + legend
library(leaflet)    # Interactive web maps
library(ggspatial)  # Scale bars and north arrows for ggplot maps
library(patchwork)  # Combine multiple ggplots into one figure
library(viridis)    # Perceptually uniform color scales
```

```{r}
#| label: api-key
#| eval: false

# Uncomment once and restart R — your key is stored permanently
# census_api_key("YOUR KEY GOES HERE", install = TRUE)
```

---

# ── Step 1: Pull Data & Build the Hardship Index ──────────────

We reuse the exact same index construction from Lab 1: five ACS variables
measuring economic hardship, z-score standardized, with median income reversed.
This time we also pull **two new demographic variables** --- percent minority
(non-white) and percent renter-occupied --- that we will use for bivariate
mapping later.

## 1a. Define variables

```{r}
#| label: step.01a

# The 5 hardship index variables (identical to Lab 1)
hardship_vars <- c(
  pov_num      = "B17001_002",   # Population below poverty
  pov_den      = "B17001_001",   # Poverty universe total
  med_income   = "B19013_001",   # Median household income
  unemp_num    = "B23025_005",   # Unemployed
  unemp_den    = "B23025_002",   # In labor force
  nohs_num     = "B15003_002",   # No high school diploma (< 9th grade)
  nohs_den     = "B15003_001",   # Population 25+
  noins_num    = "B27010_017",   # No health insurance (18-34)
  noins_den    = "B27010_001"    # Insurance universe
)

# New demographic variables for bivariate mapping
demo_vars <- c(
  race_total   = "B03002_001",   # Total population (Hispanic/Latino origin by race)
  race_wnh     = "B03002_003",   # White alone, not Hispanic or Latino
  tenure_total = "B25003_001",   # Total occupied housing units
  tenure_rent  = "B25003_003"    # Renter-occupied units
)

# Combine into one request
all_vars <- c(hardship_vars, demo_vars)
```

## 1b. Pull Maricopa County block group data

```{r}
#| label: step.01b
#| results: hide

# Pull ACS 5-year estimates at the block group level for Maricopa County.
# geometry = TRUE attaches spatial boundaries for mapping.
cbg_raw <- get_acs(
  geography = "block group",
  variables = all_vars,
  state     = "AZ",
  county    = "Maricopa",
  year      = 2023,
  survey    = "acs5",
  geometry  = TRUE
)
```

## 1c. Reshape and compute rates

```{r}
#| label: step.01c

# Pivot from long to wide: one row per CBG, one column per variable.
# We drop the margin of error column since we're working with estimates only.
cbg <- cbg_raw %>%
  select(-moe) %>%
  pivot_wider(names_from = variable, values_from = estimate)

# Compute derived rates from numerator/denominator pairs
cbg <- cbg %>%
  mutate(
    poverty_rate = pov_num / pov_den,
    unemp_rate   = unemp_num / unemp_den,
    no_hs_rate   = nohs_num / nohs_den,
    no_ins_rate  = noins_num / noins_den,
    pct_minority = (race_total - race_wnh) / race_total,
    pct_renter   = tenure_rent / tenure_total
  )

# Z-score standardize the 5 hardship variables, reverse median income,
# and compute the composite index as the row mean of z-scores.
cbg <- cbg %>%
  mutate(
    across(
      c(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, med_income),
      ~ as.numeric(scale(.)),
      .names = "z_{.col}"
    ),
    # Reverse median income so higher = more hardship
    z_med_income = -1 * z_med_income,
    # Composite hardship index: mean of the 5 z-scores
    hardship_index = rowMeans(
      pick(z_poverty_rate, z_unemp_rate, z_no_hs_rate, z_no_ins_rate, z_med_income),
      na.rm = TRUE
    )
  )

# Quick summary — how many CBGs, what's the index range?
cat("CBGs:", nrow(cbg), "\n")
cat("Hardship index range:",
    round(min(cbg$hardship_index, na.rm = TRUE), 2), "to",
    round(max(cbg$hardship_index, na.rm = TRUE), 2), "\n")
```

---

# ── Step 2: Classification Comparison ─────────────────────────

A choropleth map's story depends heavily on how you **classify** continuous
values into discrete color bins. Here we apply four common methods to the
hardship index and compare them side-by-side.

| Method | Logic | Effect |
|--------|-------|--------|
| **Quantile** | Equal number of observations per bin | Maximizes color variation; can exaggerate small differences |
| **Jenks (natural breaks)** | Minimizes within-class variance | Finds "natural" groupings; data-driven |
| **Equal interval** | Equal-width bins across the range | Intuitive scale; can leave most observations in one or two bins |
| **Standard deviation** | Bins centered on the mean, each 1 SD wide | Highlights outliers; assumes roughly normal distribution |

```{r}
#| label: step.02a

# ── Compute classification breaks using classInt ──────────────
# classInt::classIntervals() takes a numeric vector and a method name,
# and returns an object whose $brks element contains the break points.
# We request 5 classes (n = 5) for each method.

# Drop NAs for classification
hi_values <- cbg$hardship_index[!is.na(cbg$hardship_index)]

brk_quantile <- classIntervals(hi_values, n = 5, style = "quantile")
brk_jenks    <- classIntervals(hi_values, n = 5, style = "jenks")
brk_equal    <- classIntervals(hi_values, n = 5, style = "equal")
brk_sd       <- classIntervals(hi_values, n = 5, style = "sd")
```

```{r}
#| label: step.02b

# ── Helper function to cut the index and produce a map ────────
# This avoids repeating the same ggplot code four times.
# We use cut() with the break points from classIntervals to assign
# each CBG to a class, then map the classes as a factor fill.

make_class_map <- function(data, breaks_obj, method_label) {
  # cut() assigns each value to a bin defined by breaks_obj$brks
  data <- data %>%
    mutate(
      hi_class = cut(
        hardship_index,
        breaks  = breaks_obj$brks,
        include.lowest = TRUE,
        dig.lab = 2
      )
    )

  ggplot(data) +
    geom_sf(aes(fill = hi_class), color = NA) +
    scale_fill_viridis_d(option = "inferno", direction = -1, na.value = "grey80") +
    coord_sf(datum = NA) +
    theme_void(base_size = 11) +
    labs(
      title = method_label,
      fill  = "Hardship\nClass"
    )
}
```

```{r}
#| label: step.02c
#| fig-height: 10

# ── Build 4 maps and arrange in a 2x2 grid ───────────────────
# patchwork lets us combine ggplots with + and / operators.
# The + operator places plots side-by-side; / stacks them vertically.

p_quant <- make_class_map(cbg, brk_quantile, "Quantile")
p_jenks <- make_class_map(cbg, brk_jenks,    "Jenks Natural Breaks")
p_equal <- make_class_map(cbg, brk_equal,    "Equal Interval")
p_sd    <- make_class_map(cbg, brk_sd,       "Standard Deviation")

(p_quant + p_jenks) / (p_equal + p_sd) +
  plot_annotation(
    title    = "How Classification Method Changes the Story",
    subtitle = "Economic Hardship Index — Maricopa County CBGs (ACS 2019-2023)",
    caption  = "Source: ACS 5-Year Estimates, 2019-2023"
  )
```

**Key observation:** The *same data* produces four visually different maps. The
quantile method spreads color evenly across CBGs, making hardship appear
widespread. The equal interval method concentrates most CBGs in a narrow range,
making only extreme outliers visible. Jenks finds natural clusters in the data.
Standard deviation highlights how far each area deviates from the mean.

---

# ── Step 3: Bivariate Choropleth — Hardship x % Minority ──────

A standard choropleth maps **one** variable. A bivariate choropleth maps
**two** variables simultaneously by crossing their classifications into a 3x3
(or 2x2) color grid. This lets us see where *both* hardship and minority
concentration are high --- areas of potential "double disadvantage."

```{r}
#| label: step.03a

# ── Create the bivariate classes using the biscale package ────
# bi_class() takes a data frame and two variable names, and adds a
# new column (bi_class) that encodes the 3x3 combination.
# dim = 3 means each variable is split into terciles (low/mid/high).

cbg_bi <- cbg %>%
  filter(!is.na(hardship_index), !is.na(pct_minority)) %>%
  bi_class(
    x   = hardship_index,  # x-axis variable
    y   = pct_minority,    # y-axis variable
    dim = 3,               # 3x3 grid = 9 color categories
    style = "quantile"     # use quantile breaks for both variables
  )
```

```{r}
#| label: step.03b
#| fig-height: 8

# ── Build the bivariate map ──────────────────────────────────
# bi_scale_fill() maps the bi_class column to a bivariate color palette.
# "DkViolet" is a purple-to-teal palette that works well for diverging pairs.

map_bi <- ggplot(cbg_bi) +
  geom_sf(aes(fill = bi_class), color = NA, show.legend = FALSE) +
  bi_scale_fill(pal = "DkViolet", dim = 3) +
  coord_sf(datum = NA) +
  theme_void(base_size = 12) +
  labs(
    title    = "Economic Hardship x Percent Minority",
    subtitle = "Maricopa County Block Groups (ACS 2019-2023)",
    caption  = "Source: ACS 5-Year Estimates, 2019-2023"
  )

# ── Build the bivariate legend ────────────────────────────────
# bi_legend() creates the 3x3 color key that tells readers how to
# interpret the two dimensions. We place it in the lower-left.

legend_bi <- bi_legend(
  pal = "DkViolet",
  dim = 3,
  xlab = "Higher Hardship →",
  ylab = "Higher % Minority →",
  size = 9
)

# ── Combine map and legend using patchwork ────────────────────
# inset_element() overlays the legend on the map at specified coordinates.
map_bi +
  inset_element(
    legend_bi,
    left   = 0.0,
    bottom = 0.0,
    right  = 0.30,
    top    = 0.30
  )
```

**Interpreting the bivariate map:** The darkest purple areas (upper-right of the
legend) have *both* high hardship and high minority concentration --- these are
the "double disadvantage" neighborhoods. Light areas (lower-left) have low
hardship and low minority share. The off-diagonal colors show areas where the
two variables diverge: high hardship but low minority share, or vice versa.

---

# ── Step 4: Small Multiples Across Scales ─────────────────────

One of the most important lessons in spatial analysis is that **patterns change
with scale**. A national county map reveals broad regional divides; a tract map
shows neighborhood variation; a CBG map exposes block-level inequity. Here we
build the hardship index at all three scales and compare.

## 4a. Pull and compute at county and tract levels

```{r}
#| label: step.04a
#| results: hide

# ── Pull US county-level data ─────────────────────────────────
# This gives us ~3,200 counties across the entire US.
county_raw <- get_acs(
  geography = "county",
  variables = hardship_vars,
  year      = 2023,
  survey    = "acs5",
  geometry  = TRUE
)

# ── Pull AZ tract-level data ─────────────────────────────────
# Census tracts within Arizona (~1,800 tracts).
tract_raw <- get_acs(
  geography = "tract",
  variables = hardship_vars,
  state     = "AZ",
  year      = 2023,
  survey    = "acs5",
  geometry  = TRUE
)
```

```{r}
#| label: step.04b

# ── Reshape and compute hardship index at each scale ──────────
# We apply the same pipeline used for CBGs: pivot wide, compute rates,
# z-score standardize, reverse income, take row mean.

build_index <- function(df) {
  df %>%
    select(-moe) %>%
    pivot_wider(names_from = variable, values_from = estimate) %>%
    mutate(
      poverty_rate = pov_num / pov_den,
      unemp_rate   = unemp_num / unemp_den,
      no_hs_rate   = nohs_num / nohs_den,
      no_ins_rate  = noins_num / noins_den
    ) %>%
    mutate(
      across(
        c(poverty_rate, unemp_rate, no_hs_rate, no_ins_rate, med_income),
        ~ as.numeric(scale(.)),
        .names = "z_{.col}"
      ),
      z_med_income = -1 * z_med_income,
      hardship_index = rowMeans(
        pick(z_poverty_rate, z_unemp_rate, z_no_hs_rate, z_no_ins_rate, z_med_income),
        na.rm = TRUE
      )
    )
}

county_df <- build_index(county_raw)
tract_df  <- build_index(tract_raw)
# CBG already built above — we reuse 'cbg'
```

## 4b. Create the small-multiple comparison

```{r}
#| label: step.04c
#| fig-height: 5

# ── National county map ───────────────────────────────────────
# shift_geometry() moves Alaska and Hawaii for a clean CONUS layout.
# We filter to lower 48 + DC for simplicity.

county_lower48 <- county_df %>%
  filter(!str_sub(GEOID, 1, 2) %in% c("02", "15", "72", "78", "66", "69", "60"))

p_county <- ggplot(county_lower48) +
  geom_sf(aes(fill = hardship_index), color = NA) +
  scale_fill_viridis(option = "inferno", direction = -1, name = "Hardship") +
  coord_sf(datum = NA) +
  theme_void(base_size = 10) +
  labs(title = "US Counties")
```

```{r}
#| label: step.04d
#| fig-height: 5

# ── Arizona tract map ────────────────────────────────────────
p_tract <- ggplot(tract_df) +
  geom_sf(aes(fill = hardship_index), color = NA) +
  scale_fill_viridis(option = "inferno", direction = -1, name = "Hardship") +
  coord_sf(datum = NA) +
  theme_void(base_size = 10) +
  labs(title = "Arizona Tracts")
```

```{r}
#| label: step.04e
#| fig-height: 5

# ── Maricopa CBG map ─────────────────────────────────────────
p_cbg <- ggplot(cbg) +
  geom_sf(aes(fill = hardship_index), color = NA) +
  scale_fill_viridis(option = "inferno", direction = -1, name = "Hardship") +
  coord_sf(datum = NA) +
  theme_void(base_size = 10) +
  labs(title = "Maricopa CBGs")
```

```{r}
#| label: step.04f
#| fig-height: 6

# ── Combine the three scales into one figure ──────────────────
# patchwork layout: national on top spanning full width, state and
# local side-by-side below.

p_county / (p_tract + p_cbg) +
  plot_annotation(
    title    = "Economic Hardship Across Three Geographic Scales",
    subtitle = "ACS 5-Year Estimates, 2019-2023",
    caption  = "Source: ACS 5-Year Estimates, 2019-2023"
  )
```

**Scale matters.** The county map shows broad regional patterns (e.g., the
Mississippi Delta, Appalachia, the Southwest border). The tract map reveals
within-Arizona variation that is invisible at the county level. The CBG map
exposes block-level heterogeneity *within* tracts --- some tracts that appear
moderate at the tract level contain both very-high and very-low hardship CBGs.

---

# ── Step 5: Interactive Leaflet Map ───────────────────────────

Static maps are great for print; interactive maps let users explore. Here we
build a Leaflet map of Maricopa County CBGs with popups that show each CBG's
hardship index and component values.

```{r}
#| label: step.05a

# ── Prepare data for Leaflet ─────────────────────────────────
# Leaflet requires WGS84 (EPSG:4326) coordinate reference system.
# st_transform() reprojects our data from whatever CRS tidycensus
# returned to the standard lat/lon CRS that Leaflet expects.

cbg_leaf <- cbg %>%
  filter(!is.na(hardship_index)) %>%
  st_transform(crs = 4326)

# ── Build a color palette ────────────────────────────────────
# colorNumeric() from leaflet maps a continuous variable to a color
# ramp. We use the "inferno" palette reversed so dark = high hardship.

pal <- colorNumeric(
  palette = viridis(256, option = "inferno", direction = -1),
  domain  = cbg_leaf$hardship_index
)
```

```{r}
#| label: step.05b

# ── Build the popup HTML string ──────────────────────────────
# Each popup shows the CBG's GEOID, hardship index, and the two
# demographic variables so users can explore patterns interactively.

popup_text <- paste0(
  "<strong>CBG: </strong>",   cbg_leaf$GEOID, "<br>",
  "<strong>Hardship Index: </strong>",
    round(cbg_leaf$hardship_index, 2), "<br>",
  "<strong>Poverty Rate: </strong>",
    round(cbg_leaf$poverty_rate * 100, 1), "%<br>",
  "<strong>% Minority: </strong>",
    round(cbg_leaf$pct_minority * 100, 1), "%<br>",
  "<strong>% Renter: </strong>",
    round(cbg_leaf$pct_renter * 100, 1), "%"
)
```

```{r}
#| label: step.05c

# ── Assemble the Leaflet map ─────────────────────────────────
# leaflet() initializes the map; addProviderTiles() adds a basemap;
# addPolygons() draws the CBG boundaries with our color palette;
# addLayersControl() lets users toggle between basemaps.

leaflet(cbg_leaf) %>%
  # Two basemap options — user can toggle between them

  addProviderTiles(providers$CartoDB.Positron, group = "Light") %>%
  addProviderTiles(providers$Esri.WorldImagery,  group = "Satellite") %>%
  # CBG polygons colored by hardship index
  addPolygons(
    fillColor   = ~pal(hardship_index),
    fillOpacity = 0.7,
    color       = "white",       # border color
    weight      = 0.3,           # border width
    popup       = popup_text,
    highlightOptions = highlightOptions(
      weight    = 2,
      color     = "#ffc627",     # ASU gold highlight on hover
      bringToFront = TRUE
    )
  ) %>%
  # Color legend
  addLegend(
    pal      = pal,
    values   = ~hardship_index,
    title    = "Hardship Index",
    position = "bottomright"
  ) %>%
  # Layer control lets users switch basemaps
  addLayersControl(
    baseGroups = c("Light", "Satellite"),
    options    = layersControlOptions(collapsed = FALSE)
  )
```

---

# ── Step 6: Export-Quality Figure with ggspatial ──────────────

For reports and presentations you often need a polished static map with a scale
bar and north arrow. The `ggspatial` package adds these cartographic elements to
any `ggplot2` + `geom_sf()` map.

```{r}
#| label: step.06
#| fig-height: 8

# ── Publication-ready map with scale bar and north arrow ──────
# annotation_scale() adds a scale bar (default = km).
# annotation_north_arrow() adds a north arrow with configurable style.

ggplot(cbg %>% filter(!is.na(hardship_index))) +
  geom_sf(aes(fill = hardship_index), color = NA) +
  scale_fill_viridis(
    option    = "inferno",
    direction = -1,
    name      = "Hardship\nIndex"
  ) +
  # ggspatial scale bar — placed at bottom-left

  annotation_scale(
    location = "bl",
    width_hint = 0.3,
    style = "ticks"
  ) +
  # ggspatial north arrow — placed at top-right
  annotation_north_arrow(
    location = "tr",
    style = north_arrow_fancy_orienteering()
  ) +
  coord_sf(datum = NA) +
  theme_void(base_size = 12) +
  labs(
    title    = "Economic Hardship Index",
    subtitle = "Maricopa County Block Groups (ACS 2019-2023)",
    caption  = "Source: ACS 5-Year Estimates, 2019-2023",
    fill     = "Hardship\nIndex"
  )
```

---

# Questions

## Question 1 — Interpret: Classification and the Spatial Story

Look at the 2x2 classification comparison from Step 2.

**1a.** Which classification method makes economic hardship appear *most
widespread* across Maricopa County? Which makes it appear *most concentrated*?

> **Answer:** The **quantile** method makes hardship appear most widespread
> because it forces an equal number of CBGs into each class --- so the top class
> always contains 20% of all CBGs regardless of how extreme their values are.
> The **equal interval** method makes hardship appear most concentrated because
> it divides the full numeric range into equal-width bins; since most CBGs
> cluster near the middle of the distribution, the highest bin captures only the
> few extreme outliers. The result is a map that appears mostly uniform with a
> handful of dark patches.

**1b.** In a policy memo arguing for increased funding to "high hardship" areas,
which classification method would you choose and why? What about a memo arguing
that hardship is *not* a widespread problem?

> **Answer:** To argue hardship is widespread, use **quantile** breaks --- the
> map will show dark colors spread across many CBGs. To argue hardship is
> limited, use **equal interval** --- the map will appear mostly light, with
> only a few dark outliers. This demonstrates why classification is not a
> neutral technical decision; it is an analytical and rhetorical choice. The
> **Jenks** method is often the most defensible middle ground because it lets
> the data determine the groupings rather than imposing an arbitrary scheme.

---

## Question 2 — Modify: Bivariate Map with Hardship x % Renter-Occupied

Create a **second bivariate choropleth** using `hardship_index` and
`pct_renter` (percent renter-occupied housing) instead of `pct_minority`.
Show the complete code and output.

```{r}
#| label: q2.a

# ── Bivariate classification: hardship x % renter ────────────
# Same approach as Step 3 but substituting pct_renter for pct_minority.
# This lets us see whether areas of high hardship are also areas of
# high renter concentration (potential housing instability).

cbg_bi_renter <- cbg %>%
  filter(!is.na(hardship_index), !is.na(pct_renter)) %>%
  bi_class(
    x     = hardship_index,
    y     = pct_renter,
    dim   = 3,
    style = "quantile"
  )
```

```{r}
#| label: q2.b
#| fig-height: 8

# ── Build the bivariate map ──────────────────────────────────
map_bi_renter <- ggplot(cbg_bi_renter) +
  geom_sf(aes(fill = bi_class), color = NA, show.legend = FALSE) +
  bi_scale_fill(pal = "DkViolet", dim = 3) +
  coord_sf(datum = NA) +
  theme_void(base_size = 12) +
  labs(
    title    = "Economic Hardship x Percent Renter-Occupied",
    subtitle = "Maricopa County Block Groups (ACS 2019-2023)",
    caption  = "Source: ACS 5-Year Estimates, 2019-2023"
  )

# ── Build the bivariate legend ────────────────────────────────
legend_bi_renter <- bi_legend(
  pal  = "DkViolet",
  dim  = 3,
  xlab = "Higher Hardship →",
  ylab = "Higher % Renter →",
  size = 9
)

# ── Combine map and legend ────────────────────────────────────
map_bi_renter +
  inset_element(
    legend_bi_renter,
    left   = 0.0,
    bottom = 0.0,
    right  = 0.30,
    top    = 0.30
  )
```

---

## Question 3 — Interpret: Comparing the Two Bivariate Maps

Look at the bivariate map from Step 3 (hardship x % minority) and your new map
from Question 2 (hardship x % renter).

**3a.** Do the same areas show "double disadvantage" in both maps? Or do the
spatial patterns differ?

> **Answer:** There is substantial overlap but not a perfect match. Many CBGs in
> south Phoenix and parts of central Mesa show dark purple in *both* maps ---
> these are areas with simultaneously high hardship, high minority share, and
> high renter rates. However, some areas near university campuses (e.g., near
> ASU Tempe) may show high renter concentration without high minority share or
> extreme hardship, appearing in the upper-left corner of the renter map's
> legend (high renter, low hardship). Conversely, some predominantly minority
> CBGs in outlying areas may have moderate-to-high homeownership rates, creating
> divergence between the two maps.

**3b.** What does this comparison tell you about the relationship between
economic hardship, race/ethnicity, and housing tenure in Maricopa County?

> **Answer:** The comparison reveals that hardship, minority concentration, and
> renter status are **correlated but not identical** spatial phenomena. Areas
> where all three overlap represent the deepest structural disadvantage ---
> communities facing economic deprivation, racial segregation, and housing
> instability simultaneously. The areas where the maps diverge are analytically
> important: they challenge the assumption that these dimensions always co-occur
> and point to distinct policy levers. For example, high-renter/low-hardship
> areas near universities need different interventions than high-minority/
> high-hardship areas with stable homeownership. Bivariate mapping makes these
> distinctions visible in a way that single-variable maps cannot.
