<!-- REVISED: 2026-03-08 | See logs/module-2-log.md for change rationale -->
<!-- DIFF: Compare against original ../CourseMaterials/Labs.1-6/Lab 2/Lab2_Solutions/Lab2_Solutions.qmd -->
---
title: "Lab 02 — Solutions: Intro to Census Data"
subtitle: "PAF 516"
format:
  html:
    theme: readable
    highlight-style: tango
    self-contained: true
execute:
  echo: true
  warning: false
  message: false
  fig-width: 10
---


---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set( echo = TRUE, message=F, warning=F, fig.width=10 )
```

```{r}
#| label: packages
# Core packages for this lab
library( tidycensus )
library( tidyverse )   # includes ggplot2, dplyr, tidyr
library( viridis )     # color scales for maps
```

```{r}
#| label: api-key
# Uncomment and add your Census API key
# census_api_key("YOUR KEY GOES HERE")
```

# Step 1: Getting the Data

```{r}
#| label: var-search
#| echo: true
#| eval: true

# Search for variable names in the ACS 5-year data
VarSearch <- load_variables( 2017, "acs5", cache=TRUE )

# Convert all labels to upper case for easier searching
VarSearch$label <- toupper( VarSearch$label )

# Search for median house value variables
medianvalue <- VarSearch %>%
  mutate( contains.medianvalue = grepl( "MEDIAN VALUE", label ) ) %>%
  filter( contains.medianvalue )

# Search for median household income variables
medianincome <- VarSearch %>%
  mutate( contains.medianincome = grepl( "MEDIAN HOUSEHOLD INCOME", label ) ) %>%
  filter( contains.medianincome )
```

Now that these objects are created, we can take a look at what the variable search rendered:

```{r}
#| label: view-vars
#| echo: true
#| eval: false

# View interactively to identify the correct variable codes
View( medianvalue )
View( medianincome )
```

After looking through the objects, we can see that `B19049_001` corresponds to the total median household income and `B25077_001` corresponds to the median house value.

```{r}
#| label: get-acs-2017
#| echo: true
#| eval: true

# Define variables with readable names using a named vector
vars.to.use <- c( HHIncome   = "B19049_001",
                  HouseValue = "B25077_001" )

# Retrieve county-level data with spatial geometry for mapping
CenDF <- get_acs( geography = "county",
                  year      = 2017,
                  survey    = "acs5",
                  variables = vars.to.use,
                  geometry  = TRUE )
```


# Step 2: Transforming the Data

```{r}
#| label: transform-2017
#| echo: true
#| eval: true

# Transform from long to wide format and compute the ratio
# NOTE: Using pivot_wider() instead of deprecated spread()
CenDF <- CenDF %>%
  select( -moe ) %>%
  pivot_wider( names_from = variable, values_from = estimate ) %>%
  mutate( HHInc_HousePrice_Ratio = HouseValue / HHIncome )
```

# Step 3: Data Exploration

```{r}
#| label: explore-2017
#| echo: true
#| eval: true

# Sort by ratio: lowest to highest (most affordable counties)
CenDF.lh <- CenDF %>%
  select( NAME, HHInc_HousePrice_Ratio ) %>%
  arrange( HHInc_HousePrice_Ratio )
head( CenDF.lh, 10 )

# Sort by ratio: highest to lowest (least affordable counties)
CenDF.hl <- CenDF %>%
  select( NAME, HHInc_HousePrice_Ratio ) %>%
  arrange( desc( HHInc_HousePrice_Ratio ) )
head( CenDF.hl, 10 )
```

# Step 4: Map the Data

```{r}
#| label: map-2017
#| echo: true
#| eval: true

# Create a choropleth map of the house-price-to-income ratio
ggplot( CenDF ) +
  geom_sf( aes( fill = HHInc_HousePrice_Ratio ), color = NA ) +
  coord_sf( datum = NA ) +
  labs( title    = "House-price-to-income ratio by county",
        subtitle = "Alaska and Hawaii are shifted and not to scale",
        caption  = "Source: ACS 5-year, 2017",
        fill     = "House-price-\nto-income ratio" ) +
  scale_fill_viridis( direction = -1 )
```


# Questions

### Question 1: Exploration.

1a. In step 3 above, what is the county with the highest (and lowest) house-price-to-income ratio? What is the average house-price-to-income ratio across all US counties over the time period?

   - Answer: The county with the highest house-price-to-income ratio is `r CenDF.hl$NAME[1]` and the county with the lowest is `r CenDF.lh$NAME[1]`. The average ratio across all US counties is `r round( mean( CenDF$HHInc_HousePrice_Ratio, na.rm = TRUE ), 2 )`.

1b. Where is Maricopa County on the list? How many years of median income will it take to buy a home in Maricopa County?

   - Answer: When looking from highest house-price-to-income to lowest, Maricopa County is `r which( CenDF.hl$NAME == "Maricopa County, Arizona" )` on the list and it will take `r round( CenDF$HHInc_HousePrice_Ratio[CenDF$NAME == "Maricopa County, Arizona"], 2 )` years of median income to buy a house.

1c. Where is Los Angeles on the list and how does its ranking compare to the list reported in the Citylab report?

   - Answer: Los Angeles is `r which( CenDF.hl$NAME == "Los Angeles County, California" )` on the list from high to low. This departs from the article which lists LA as first.


### Question 2: Temporal analysis.

2a. Go back to Step 1 above, and enter in 2012 as the year value. Repeat Steps 2-4.

```{r}
#| label: get-acs-2012
#| echo: true
#| eval: true

# Retrieve 2008-2012 ACS data for temporal comparison
CenDF.12 <- get_acs( geography = "county",
                     year      = 2012,
                     survey    = "acs5",
                     variables = vars.to.use,
                     geometry  = TRUE )

# Transform and compute ratio (using pivot_wider, not spread)
CenDF.12 <- CenDF.12 %>%
  select( -moe ) %>%
  pivot_wider( names_from = variable, values_from = estimate ) %>%
  mutate( HHInc_HousePrice_Ratio = HouseValue / HHIncome )

# Sort for exploration
CenDF.12.lh <- CenDF.12 %>%
  select( NAME, HHInc_HousePrice_Ratio ) %>%
  arrange( HHInc_HousePrice_Ratio )

CenDF.12.hl <- CenDF.12 %>%
  select( NAME, HHInc_HousePrice_Ratio ) %>%
  arrange( desc( HHInc_HousePrice_Ratio ) )

# Map the 2012 data
ggplot( CenDF.12 ) +
  geom_sf( aes( fill = HHInc_HousePrice_Ratio ), color = NA ) +
  coord_sf( datum = NA ) +
  labs( title    = "House-price-to-income ratio by county",
        subtitle = "Alaska and Hawaii are shifted and not to scale",
        caption  = "Source: ACS 5-year, 2012",
        fill     = "House-price-\nto-income ratio" ) +
  scale_fill_viridis( direction = -1 )
```

   - Answer: The county with the highest house-price-to-income ratio in 2012 is `r CenDF.12.hl$NAME[1]` and the county with the lowest is `r CenDF.12.lh$NAME[1]`.


2b. Compare and contrast the findings over the different time periods.

```{r}
#| label: compare-years
#| echo: true
#| eval: true

# Prepare 2017 data for joining (drop geometry to join as data frames)
df.17 <- CenDF %>%
  sf::st_drop_geometry() %>%
  select( GEOID, HHInc_HousePrice_Ratio, NAME ) %>%
  rename( HHInc_HousePrice_Ratio.17 = HHInc_HousePrice_Ratio )

# Prepare 2012 data for joining
df.12 <- CenDF.12 %>%
  sf::st_drop_geometry() %>%
  select( GEOID, HHInc_HousePrice_Ratio ) %>%
  rename( HHInc_HousePrice_Ratio.12 = HHInc_HousePrice_Ratio )

# Join the two time periods
both.year <- full_join( df.17, df.12, by = "GEOID" )

# Compute the change score (2017 minus 2012)
both.year$HHInc_HousePrice_Ratio.d <-
  both.year$HHInc_HousePrice_Ratio.17 - both.year$HHInc_HousePrice_Ratio.12

# Summary statistics of the change
summary( both.year$HHInc_HousePrice_Ratio.d )

# Histogram of the change in ratio
ggplot( both.year, aes( x = HHInc_HousePrice_Ratio.d ) ) +
  geom_histogram( color = "black", fill = "lightblue", bins = 40 ) +
  theme_minimal() +
  labs( title = "Change in House-Price-to-Income Ratio by County\n(ACS 2017 vs. ACS 2012)",
        x     = "Change in Ratio (2017 - 2012)",
        y     = "Count" )
```

   - Answer: The overall change is close to zero, although the average decreased slightly. For Maricopa County, the change was `r round( both.year$HHInc_HousePrice_Ratio.d[both.year$NAME == "Maricopa County, Arizona"][1], 2 )`, showing an increase from 2012 to 2017. As for Los Angeles county, there was an increase of `r round( both.year$HHInc_HousePrice_Ratio.d[both.year$NAME == "Los Angeles County, California"][1], 2 )`.

### Question 3: High-Resolution Analysis.

3a. Change geography from `county` to `tract` for Maricopa County.

```{r}
#| label: get-acs-az
#| echo: true
#| eval: true
#| results: hide

# Retrieve tract-level data for Maricopa County only
CenDF.az <- get_acs( geography = "tract",
                     year      = 2017,
                     state     = "AZ",
                     county    = "Maricopa County",
                     survey    = "acs5",
                     variables = vars.to.use,
                     geometry  = TRUE )

# Transform and compute ratio (using pivot_wider, not spread)
CenDF.az <- CenDF.az %>%
  select( -moe ) %>%
  pivot_wider( names_from = variable, values_from = estimate ) %>%
  mutate( HHInc_HousePrice_Ratio = HouseValue / HHIncome )

# Summary statistics
summary( CenDF.az$HHInc_HousePrice_Ratio )
```

   - Answer: Here are the summary statistics:
     the min is `r round( min( CenDF.az$HHInc_HousePrice_Ratio, na.rm = TRUE ), 2 )`, the max is `r round( max( CenDF.az$HHInc_HousePrice_Ratio, na.rm = TRUE ), 2 )`, the median is `r round( median( CenDF.az$HHInc_HousePrice_Ratio, na.rm = TRUE ), 2 )`, the mean is `r round( mean( CenDF.az$HHInc_HousePrice_Ratio, na.rm = TRUE ), 2 )`, and the sd is `r round( sd( CenDF.az$HHInc_HousePrice_Ratio, na.rm = TRUE ), 2 )`.


3b. Compare and contrast the findings for the different levels of geography.

```{r}
#| label: compare-geography

# Build a comparison table: county-level vs tract-level statistics
compare.table <-
  rbind(
    # County-level info for Maricopa
    c( summary(
      CenDF$HHInc_HousePrice_Ratio[CenDF$NAME == "Maricopa County, Arizona"]
    ), 0 ),
    # Tract-level info for Maricopa
    summary( CenDF.az$HHInc_HousePrice_Ratio )
  )

rownames( compare.table ) <- c( "county", "tract" )
colnames( compare.table )[7] <- "NAs"

round( compare.table, 2 )
```

   - Answer: We can see that the measures of central tendency are similar. However, a major difference is with the ends of the distribution. The variation in the distributions is considerable when we compare the min/max for each level — tract-level data reveals much more heterogeneity within Maricopa County than the single county-level value suggests.
