---
title: "Economic Hardship Dashboard — Maricopa County Example"
subtitle: "Instructor Reference | Completed Analysis"
format:
  dashboard:
    theme:
      - darkly
      - ../../assets/dashboard-theme.scss
    orientation: columns
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
options(tigris_use_cache = TRUE)
options(tigris_progress_bar = FALSE)
set.seed(42)
```

```{r}
#| label: packages
#| include: false
if (!file.exists("~/.paf516_ready")) {
  if (!requireNamespace("renv", quietly = TRUE)) install.packages("renv")
  tmp <- tempfile(fileext = ".lock")
  download.file(
    url      = "https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/renv.lock",
    destfile = tmp, mode = "wb"
  )
  renv::restore(lockfile = tmp, prompt = FALSE)
  file.create("~/.paf516_ready")
}
library(tidyverse); library(tidycensus); library(sf); library(spdep)
library(viridis); library(patchwork); library(biscale); library(mapgl); library(DT)
library(scales)
```

```{r}
#| label: county-config
# ── STUDENT: change these lines + the subtitle in the YAML header above ─
# Examples of other metros:
#   Los Angeles:      TARGET_STATE="CA", TARGET_COUNTY="Los Angeles",  TARGET_LABEL="Los Angeles County, CA"
#   Cook (Chicago):   TARGET_STATE="IL", TARGET_COUNTY="Cook",         TARGET_LABEL="Cook County, IL"
#   Miami-Dade:       TARGET_STATE="FL", TARGET_COUNTY="Miami-Dade",   TARGET_LABEL="Miami-Dade County, FL"
#   King (Seattle):   TARGET_STATE="WA", TARGET_COUNTY="King",         TARGET_LABEL="King County, WA"
#   Harris (Houston): TARGET_STATE="TX", TARGET_COUNTY="Harris",       TARGET_LABEL="Harris County, TX"
TARGET_STATE  <- "AZ"
TARGET_COUNTY <- "Maricopa"
TARGET_LABEL  <- "Maricopa County, AZ"
# ─────────────────────────────────────────────────────────────────────────
```

```{r}
#| label: index-config
hardship_vars <- c(
  pov_below50   = "C17002_002",
  pov_50to99    = "C17002_003",
  pov_den       = "C17002_001",
  unemp_num     = "B23025_005",
  unemp_den     = "B23025_002",
  median_income = "B19013_001"
)
renter_vars <- c(
  renter_num = "B25003_003",
  renter_den = "B25003_001"
)

build_hardship_index <- function(raw_df) {
  raw_df %>%
    select(-moe) %>%
    pivot_wider(names_from = variable, values_from = estimate) %>%
    mutate(
      poverty_rate = (pov_below50 + pov_50to99) / pov_den,
      unemp_rate   = unemp_num / unemp_den
    ) %>%
    filter(!is.na(poverty_rate), !is.na(unemp_rate), !is.na(median_income)) %>%
    (function(df) {
      geom <- if (inherits(df, "sf")) st_geometry(df) else NULL
      tbl  <- st_drop_geometry(df)
      z_pov  <- as.numeric(scale(tbl$poverty_rate))
      z_unemp <- as.numeric(scale(tbl$unemp_rate))
      z_inc  <- -1 * as.numeric(scale(tbl$median_income))
      tbl$hardship_index <- rowMeans(cbind(z_pov, z_unemp, z_inc), na.rm = TRUE)
      if (!is.null(geom)) st_sf(tbl, geometry = geom) else tbl
    })()
}
```

```{r}
#| label: pull-national
#| cache: true
county_raw <- get_acs(
  geography    = "county",
  variables    = hardship_vars,
  year         = 2023,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
county_2023 <- build_hardship_index(county_raw)
```

```{r}
#| label: pull-local-current
#| cache: true
local_combined_raw <- get_acs(
  geography    = "tract",
  variables    = c(hardship_vars, renter_vars),
  state        = TARGET_STATE,
  county       = TARGET_COUNTY,
  year         = 2023,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
# Split: hardship index
local_2023 <- build_hardship_index(
  local_combined_raw %>% filter(variable %in% names(hardship_vars))
)
# Renter rate join
renter_tbl <- local_combined_raw %>%
  filter(variable %in% names(renter_vars)) %>%
  st_drop_geometry() %>% select(-moe) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(renter_rate = renter_num / renter_den) %>%
  select(GEOID, renter_rate)
local_2023 <- left_join(local_2023, renter_tbl, by = "GEOID")
```

```{r}
#| label: pull-temporal
#| cache: true
local_2013_raw <- get_acs(
  geography = "tract", variables = hardship_vars,
  state = TARGET_STATE, county = TARGET_COUNTY,
  year = 2013, survey = "acs5", geometry = TRUE, progress_bar = FALSE
)
local_2019_raw <- get_acs(
  geography = "tract", variables = hardship_vars,
  state = TARGET_STATE, county = TARGET_COUNTY,
  year = 2019, survey = "acs5", geometry = TRUE, progress_bar = FALSE
)
local_2016_raw <- get_acs(
  geography = "tract", variables = hardship_vars,
  state = TARGET_STATE, county = TARGET_COUNTY,
  year = 2016, survey = "acs5", geometry = FALSE, progress_bar = FALSE
)
```

```{r}
#| label: pooled-std
widen_and_rate <- function(df) {
  df %>%
    select(-moe) %>%
    pivot_wider(names_from = variable, values_from = estimate) %>%
    mutate(
      poverty_rate = (pov_below50 + pov_50to99) / pov_den,
      unemp_rate   = unemp_num / unemp_den
    ) %>%
    filter(!is.na(poverty_rate), !is.na(unemp_rate), !is.na(median_income))
}

local_2013_wide <- widen_and_rate(local_2013_raw)
local_2019_wide <- widen_and_rate(local_2019_raw)

r2013 <- local_2013_wide %>% st_drop_geometry() %>%
  select(GEOID, poverty_rate, unemp_rate, median_income)
r2019 <- local_2019_wide %>% st_drop_geometry() %>%
  select(GEOID, poverty_rate, unemp_rate, median_income)

vars_pool <- c("poverty_rate", "unemp_rate", "median_income")
pooled_stats <- map_dfr(vars_pool, function(v) {
  vals <- c(r2013[[v]], r2019[[v]])
  tibble(variable = v, mean = mean(vals, na.rm = TRUE), sd = sd(vals, na.rm = TRUE))
})

apply_pooled_z <- function(df, stats) {
  for (v in stats$variable) {
    df[[paste0("z_", v)]] <- (df[[v]] - stats$mean[stats$variable==v]) /
                              stats$sd[stats$variable==v]
  }
  df$z_median_income <- -1 * df$z_median_income
  df$hardship_index  <- rowMeans(
    df[, c("z_poverty_rate","z_unemp_rate","z_median_income")], na.rm = TRUE)
  df
}

r2013_z <- apply_pooled_z(r2013, pooled_stats)
r2019_z <- apply_pooled_z(r2019, pooled_stats)

change_df <- local_2019_wide %>%
  select(GEOID, NAME) %>%
  left_join(r2019_z %>% select(GEOID, hardship_2019 = hardship_index), by = "GEOID") %>%
  inner_join(r2013_z %>% select(GEOID, hardship_2013 = hardship_index), by = "GEOID") %>%
  mutate(hardship_change = hardship_2019 - hardship_2013)

lo <- quantile(change_df$hardship_change, 0.02, na.rm = TRUE)
hi <- quantile(change_df$hardship_change, 0.98, na.rm = TRUE)
max_abs <- max(abs(lo), abs(hi))
change_df <- change_df %>%
  mutate(change_plot = pmax(pmin(hardship_change, max_abs), -max_abs))
```

```{r}
#| label: lisa-analysis
nb    <- poly2nb(change_df, queen = TRUE)
w     <- nb2listw(nb, style = "W", zero.policy = TRUE)
lisa  <- localmoran(change_df$hardship_2019, w, zero.policy = TRUE)
z19   <- scale(change_df$hardship_2019)[,1]
wz19  <- lag.listw(w, z19, zero.policy = TRUE)
p19   <- lisa[, "Pr(z != E(Ii))"]

change_df <- change_df %>%
  mutate(lisa_2019 = case_when(
    z19 > 0 & wz19 > 0 & p19 < 0.05 ~ "HH",
    z19 < 0 & wz19 < 0 & p19 < 0.05 ~ "LL",
    z19 > 0 & wz19 < 0 & p19 < 0.05 ~ "HL",
    z19 < 0 & wz19 > 0 & p19 < 0.05 ~ "LH",
    TRUE ~ "NS"
  ))
```

```{r}
#| label: trajectories
lisa_2013  <- localmoran(change_df$hardship_2013, w, zero.policy = TRUE)
z13  <- scale(change_df$hardship_2013)[,1]
wz13 <- lag.listw(w, z13, zero.policy = TRUE)
p13  <- lisa_2013[, "Pr(z != E(Ii))"]

local_2016_wide <- widen_and_rate(local_2016_raw %>% mutate(geometry = NULL))
r2016_z <- apply_pooled_z(
  local_2016_wide %>% select(GEOID, poverty_rate, unemp_rate, median_income),
  pooled_stats
)

change_df <- change_df %>%
  mutate(
    lisa_2013 = case_when(
      z13 > 0 & wz13 > 0 & p13 < 0.05 ~ "HH",
      z13 < 0 & wz13 < 0 & p13 < 0.05 ~ "LL",
      z13 > 0 & wz13 < 0 & p13 < 0.05 ~ "HL",
      z13 < 0 & wz13 > 0 & p13 < 0.05 ~ "LH",
      TRUE ~ "NS"
    ),
    trajectory = case_when(
      lisa_2013 == "HH" & lisa_2019 == "HH" ~ "Persistent HH",
      lisa_2013 != "HH" & lisa_2019 == "HH" ~ "Emerging HH",
      lisa_2013 == "HH" & lisa_2019 != "HH" ~ "Dissolving HH",
      lisa_2013 == "LL" & lisa_2019 == "LL" ~ "Persistent LL",
      lisa_2013 != "LL" & lisa_2019 == "LL" ~ "Emerging LL",
      lisa_2013 == "HL" | lisa_2019 == "HL" ~ "HL Outlier",
      TRUE                                   ~ "Stable NS"
    )
  ) %>%
  left_join(r2016_z %>% select(GEOID, hardship_2016 = hardship_index), by = "GEOID") %>%
  mutate(trend_type = case_when(
    hardship_2013 > hardship_2016 & hardship_2016 > hardship_2019 ~ "Consistently improving",
    hardship_2013 < hardship_2016 & hardship_2016 < hardship_2019 ~ "Consistently worsening",
    hardship_2013 > hardship_2016 & hardship_2016 < hardship_2019 ~ "V-shaped (improved mid)",
    hardship_2013 < hardship_2016 & hardship_2016 > hardship_2019 ~ "Inverted-V (worsened mid)",
    TRUE ~ "Flat / non-monotone"
  ))
```

```{r}
#| label: summary-stats
moran_global <- moran.test(change_df$hardship_2019, w, zero.policy = TRUE)

n_hh <- sum(change_df$lisa_2019 == "HH", na.rm = TRUE)
n_ll <- sum(change_df$lisa_2019 == "LL", na.rm = TRUE)
pct_improved    <- round(100 * mean(change_df$hardship_change < -0.05, na.rm=TRUE), 1)
pct_worsened    <- round(100 * mean(change_df$hardship_change >  0.05, na.rm=TRUE), 1)
pct_persist_hh  <- round(100 * mean(change_df$trajectory == "Persistent HH", na.rm=TRUE), 1)
pct_emerg_hh    <- round(100 * mean(change_df$trajectory == "Emerging HH",   na.rm=TRUE), 1)

county_hi <- county_2023 %>% st_drop_geometry() %>%
  filter(!is.na(hardship_index)) %>%
  slice_max(hardship_index, n = 1, with_ties = FALSE) %>%
  mutate(label = paste0(str_remove(NAME, ",.*"), " (", round(hardship_index,2), ")")) %>%
  pull(label)
n_counties <- sum(!is.na(county_2023$hardship_index))

traj_colors <- c(
  "Persistent HH"="#D7191C", "Emerging HH"="#FDAE61", "Dissolving HH"="#FEE08B",
  "Persistent LL"="#2C7BB6", "Emerging LL"="#ABD9E9", "HL Outlier"="#9E5BB4",
  "Stable NS"="#CCCCCC"
)
lisa_colors <- c("HH"="#D7191C","LL"="#2C7BB6","HL"="#FDAE61","LH"="#ABD9E9","NS"="#CCCCCC")
```

# National Context

## Column {width=62%}

```{r}
#| label: map-national
county_4326 <- county_2023 %>%
  st_transform(4326) %>%
  filter(!is.na(hardship_index)) %>%
  st_simplify(preserveTopology = TRUE, dTolerance = 5000)
maplibre(
  bounds = st_bbox(county_4326),
  style  = carto_style("dark-matter")
) %>%
  add_fill_layer(
    id     = "counties",
    source = list(type = "geojson", data = county_4326),
    fill_color = interpolate(
      column   = "hardship_index",
      values   = as.numeric(quantile(county_4326$hardship_index, c(0, 0.25, 0.5, 0.75, 1))),
      stops    = c("#313695","#74ADD1","#FFFFBF","#F46D43","#A50026"),
      na_color = "#333333"
    ),
    fill_opacity = 0.8,
    tooltip = c("NAME", "hardship_index")
  ) %>%
  add_legend(
    "Economic Hardship Index",
    values   = round(quantile(county_4326$hardship_index, c(0,0.25,0.5,0.75,1)), 2),
    colors   = c("#313695","#74ADD1","#FFFFBF","#F46D43","#A50026"),
    position = "bottom-left"
  )
```

## Column {width=38%}

```{r}
#| label: vb-counties
#| content: valuebox
list(title="Counties Mapped", value=comma(n_counties), icon="map")
```

```{r}
#| label: vb-highest
#| content: valuebox
list(title="Highest Hardship County", value=county_hi, icon="exclamation-triangle-fill", color="danger")
```

```{r}
#| label: vb-rank
#| content: valuebox
list(
  title = paste0(TARGET_LABEL, " National Rank"),
  value = {
    rank_val <- rank(-county_2023$hardship_index, na.last="keep", ties.method="min")
    target_geoid <- county_2023 %>%
      st_drop_geometry() %>%
      filter(str_detect(NAME, fixed(TARGET_COUNTY)),
             str_detect(NAME, state.name[match(TARGET_STATE, state.abb)])) %>%
      slice(1)
    if(nrow(target_geoid) > 0) {
      r <- rank_val[which(county_2023$GEOID == target_geoid$GEOID[1])]
      paste0("#", r, " of ", n_counties)
    } else "—"
  },
  icon = "bar-chart-fill"
)
```

```{r}
#| label: table-national
county_2023 %>%
  st_drop_geometry() %>%
  filter(!is.na(hardship_index)) %>%
  arrange(desc(hardship_index)) %>%
  mutate(
    County   = str_remove(NAME, ", .+"),
    State    = str_extract(NAME, "(?<=, )[A-Z].+"),
    Hardship = round(hardship_index, 3)
  ) %>%
  select(County, State, Hardship) %>%
  head(20) %>%
  DT::datatable(options=list(pageLength=8, dom="tp"), rownames=FALSE)
```

# Local Hardship

## Column {width=50%}

```{r}
#| label: map-local
#| fig-height: 7
ggplot(local_2023) +
  geom_sf(aes(fill = hardship_index), color = NA) +
  scale_fill_viridis_c(option="inferno", direction=-1,
                       name="Hardship\nIndex", na.value="grey40") +
  theme_void(base_size=11) +
  labs(title    = paste0("Economic Hardship: ", TARGET_LABEL),
       subtitle = "Census tracts | ACS 2019–2023 5-Year",
       caption  = "3-variable index: poverty rate, unemployment rate, median income (reversed)")
```

## Column {width=50%}

```{r}
#| label: map-bivariate
#| fig-height: 7
bi_data <- local_2023 %>%
  filter(!is.na(hardship_index), !is.na(renter_rate)) %>%
  bi_class(x=hardship_index, y=renter_rate, style="quantile", dim=3)

p_bi <- ggplot(bi_data) +
  geom_sf(aes(fill=bi_class), color=NA, show.legend=FALSE) +
  bi_scale_fill(pal="DkBlue", dim=3) +
  theme_void(base_size=11) +
  labs(title    = "Hardship × Renter Rate",
       subtitle = paste0(TARGET_LABEL, " | ACS 2019–2023"),
       caption  = "Dark blue = high hardship AND high renter rate")

p_legend <- bi_legend(pal="DkBlue", dim=3,
                       xlab="Hardship →", ylab="% Renter →", size=9)

p_bi + inset_element(p_legend, left=0.72, bottom=0.05, right=1, top=0.35)
```

# Hot Spots

## Column {width=60%}

```{r}
#| label: map-lisa
#| fig-height: 8
ggplot(change_df) +
  geom_sf(aes(fill=lisa_2019), color=NA) +
  scale_fill_manual(values=lisa_colors,
    labels=c("HH"="Hot Spot (HH)","LL"="Cold Spot (LL)",
             "HL"="High-Low Outlier","LH"="Low-High Outlier","NS"="Not Significant"),
    name="LISA Cluster\n(2019, p<0.05)") +
  theme_void(base_size=11) +
  labs(title    = paste0("Spatial Hardship Clusters: ", TARGET_LABEL),
       subtitle = "LISA analysis on 2015–2019 ACS economic hardship index",
       caption  = paste0("Global Moran's I = ", round(moran_global$estimate[1],3),
                         " (p < 0.001) — significant positive spatial autocorrelation"))
```

## Column {width=40%}

```{r}
#| label: vb-hh
#| content: valuebox
list(title="Hot Spot Tracts (HH)", value=n_hh, icon="fire", color="danger")
```

```{r}
#| label: vb-ll
#| content: valuebox
list(title="Cold Spot Tracts (LL)", value=n_ll, icon="snow", color="primary")
```

```{r}
#| label: vb-moran
#| content: valuebox
list(title="Global Moran's I", value=round(unname(moran_global$estimate[1]),3),
     icon="diagram-3", color="secondary")
```

```{r}
#| label: table-hotspots
change_df %>%
  st_drop_geometry() %>%
  filter(lisa_2019 %in% c("HH","LL")) %>%
  arrange(desc(hardship_2019)) %>%
  mutate(
    Tract    = str_extract(NAME, "Census Tract [\\d\\.]+"),
    Cluster  = lisa_2019,
    Hardship = round(hardship_2019, 3),
    Trajectory = trajectory
  ) %>%
  select(Tract, Cluster, Hardship, Trajectory) %>%
  DT::datatable(options=list(pageLength=10, dom="tp"), rownames=FALSE)
```

# Change & Trajectories

## Column {width=50%}

```{r}
#| label: map-trajectory
#| fig-height: 8
ggplot(change_df) +
  geom_sf(aes(fill=trajectory), color=NA) +
  scale_fill_manual(values=traj_colors,
    labels=c("Persistent HH"="Persistent Hot Spot",
             "Emerging HH"  ="Emerging Hot Spot",
             "Dissolving HH"="Dissolving Hot Spot",
             "Persistent LL"="Persistent Cold Spot",
             "Emerging LL"  ="Emerging Cold Spot",
             "HL Outlier"   ="Spatial Outlier",
             "Stable NS"    ="No Dominant Pattern"),
    name="Trajectory\n(2013→2019)") +
  theme_void(base_size=11) +
  labs(title    = paste0("Neighborhood Hardship Trajectories: ", TARGET_LABEL),
       subtitle = "LISA cluster transition, 2013–2019 | Pooled z-score standardization",
       caption  = "Pooled mean/SD across both years — comparable change scores")
```

## Column {width=50%}

```{r}
#| label: vb-improved
#| content: valuebox
list(title="Tracts Improved (2013→2019)", value=paste0(pct_improved,"%"),
     icon="arrow-down-circle", color="success")
```

```{r}
#| label: vb-worsened
#| content: valuebox
list(title="Tracts Worsened (2013→2019)", value=paste0(pct_worsened,"%"),
     icon="arrow-up-circle", color="warning")
```

```{r}
#| label: vb-persistent
#| content: valuebox
list(title="Persistently High Hardship", value=paste0(pct_persist_hh,"%"),
     icon="exclamation-circle", color="danger")
```

```{r}
#| label: vb-emerging
#| content: valuebox
list(title="Emerging Hot Spots", value=paste0(pct_emerg_hh,"%"),
     icon="arrow-up-right-circle", color="warning")
```

```{r}
#| label: map-change
#| fig-height: 5.5
ggplot(change_df) +
  geom_sf(aes(fill=change_plot), color=NA) +
  scale_fill_gradient2(low="#2C7BB6", mid="white", high="#D7191C",
    midpoint=0, limits=c(-max_abs, max_abs), name="Hardship\nChange") +
  theme_void(base_size=10) +
  labs(title    = paste0("Hardship Change (2013→2019): ", TARGET_LABEL),
       subtitle = "Blue = improved | Red = worsened | Winsorized at 2nd/98th pct")
```

# Policy

## Column {width=55%}

```{r}
#| label: table-policy
change_df %>%
  st_drop_geometry() %>%
  filter(trajectory %in% c("Persistent HH","Emerging HH")) %>%
  arrange(desc(hardship_2019)) %>%
  mutate(
    Tract        = str_extract(NAME, "Census Tract [\\d\\.]+"),
    Trajectory   = trajectory,
    `Hardship 2019` = round(hardship_2019, 3),
    `Change 2013–19` = round(hardship_change, 3)
  ) %>%
  select(Tract, Trajectory, `Hardship 2019`, `Change 2013–19`) %>%
  DT::datatable(
    filter  = "top",
    options = list(pageLength=12, dom="ftp"),
    rownames = FALSE
  )
```

## Column {width=45%}

::: {.card}
## Recommendation 1 — Place-Based Workforce Investment in South Phoenix Persistent Hot Spots
The 175 High-High cluster tracts concentrated in the south Phoenix corridor — spanning the South Mountain Village, Laveen, and Maryvale planning areas — show persistent economic hardship across both the 2013 and 2019 analysis windows, with 16.6% of all county tracts classified as persistently high hardship. The Global Moran's I of 0.624 confirms that this clustering is not random but reflects deeply structural spatial inequality in employment access, income distribution, and poverty concentration. Maricopa County Workforce Development, in partnership with the City of Phoenix Community and Economic Development Department, should establish three satellite workforce training centers within the persistent HH cluster, targeting the approximately 150 tracts where hardship indices exceed 1.0 — prioritizing sector-based training in healthcare and logistics, the two fastest-growing industries within a 30-minute commute radius of the cluster. This place-based approach addresses the spatial mismatch between where hardship concentrates and where employment opportunities exist, a dynamic that county-wide job training programs systematically fail to address.
:::

::: {.card}
## Recommendation 2 — Early Intervention Monitoring for Emerging Hot Spots in West Mesa and South Glendale
Approximately 2.8% of Maricopa County tracts are classified as Emerging HH — neighborhoods that were not significant hardship clusters in 2013 but became statistically significant hot spots by 2019. These newly forming clusters, located primarily in west Mesa south of US-60 and in south Glendale near the 51st Avenue corridor, represent an early warning signal that hardship is spatially expanding beyond the historically disadvantaged south Phoenix core. The Maricopa Association of Governments (MAG) should integrate this LISA trajectory framework into its annual Regional Analysis Letter, triggering automatic referral to the City of Mesa Office of Economic Development and City of Glendale Community Services when any tract transitions from NS to HH status across consecutive ACS releases. Early intervention in these emerging clusters — including expedited Section 8 voucher allocation and targeted small-business micro-lending — is substantially more cost-effective than remediation after hardship becomes entrenched, as demonstrated by the persistence of the south Phoenix cluster across an entire decade.
:::

::: {.card}
## Recommendation 3 — Countywide Hardship Surveillance Dashboard for Data-Driven Resource Allocation
The trajectory analysis reveals that 85.7% of Maricopa County tracts improved between 2013 and 2019, yet 16.6% remained persistently high-hardship and 2.8% newly emerged — indicating that countywide improvement masks deepening inequality in specific neighborhoods. The strong spatial autocorrelation (Moran's I = 0.624) means that hardship is not distributed randomly but clusters in ways that compound disadvantage through shared infrastructure deficits, limited service access, and concentrated poverty spillovers. Arizona State University's Morrison Institute for Public Policy, in partnership with the Maricopa County Board of Supervisors, should deploy a publicly accessible ACS-updated surveillance dashboard — refreshed annually as new 5-year estimates are released — that tracks the three-point trajectory classification (consistently improving, consistently worsening, V-shaped, inverted-V) for every tract. This monitoring infrastructure would enable the County's Community Services Department to shift from reactive crisis response to proactive, spatially targeted resource allocation, directing federal Community Development Block Grant (CDBG) funds to tracts showing the V-shaped pattern (temporary improvement followed by reversal) before they re-enter persistent hot-spot status.
:::
