---
title: "Economic Hardship in Arizona"
subtitle: "A Community Analytics Dashboard"
format:
  dashboard:
    theme:
      - flatly
      - ../../assets/dashboard-theme.scss
    orientation: columns
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
options(tigris_use_cache = TRUE)
options(tigris_progress_bar = FALSE)
set.seed(42)
```

```{r}
#| label: packages
#| include: false
if (!file.exists("~/.paf516_ready")) {
  if (!requireNamespace("renv", quietly = TRUE)) install.packages("renv")
  tmp <- tempfile(fileext = ".lock")
  download.file(
    url      = "https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/renv.lock",
    destfile = tmp, mode = "wb"
  )
  renv::restore(lockfile = tmp, prompt = FALSE)
  file.create("~/.paf516_ready")
}
library(tidyverse); library(tidycensus); library(sf); library(spdep)
library(viridis); library(patchwork); library(tigris); library(DT)
library(mapgl); library(glue); library(plotly); library(scales)
```

```{r}
#| label: county-config
# ── STUDENT: change these 3 lines + update the YAML subtitle above ──────
# Examples:
#   Los Angeles:      TARGET_STATE="CA", TARGET_COUNTY="Los Angeles",  TARGET_LABEL="Los Angeles County, CA"
#   Cook (Chicago):   TARGET_STATE="IL", TARGET_COUNTY="Cook",         TARGET_LABEL="Cook County, IL"
#   Miami-Dade:       TARGET_STATE="FL", TARGET_COUNTY="Miami-Dade",   TARGET_LABEL="Miami-Dade County, FL"
#   King (Seattle):   TARGET_STATE="WA", TARGET_COUNTY="King",         TARGET_LABEL="King County, WA"
#   Harris (Houston): TARGET_STATE="TX", TARGET_COUNTY="Harris",       TARGET_LABEL="Harris County, TX"
TARGET_STATE  <- "AZ"
TARGET_COUNTY <- "Maricopa"
TARGET_LABEL  <- "Maricopa County, AZ"
# ─────────────────────────────────────────────────────────────────────────
```

```{r}
#| label: index-config
hardship_vars <- c(
  pov_below50   = "C17002_002",
  pov_50to99    = "C17002_003",
  pov_den       = "C17002_001",
  unemp_num     = "B23025_005",
  unemp_den     = "B23025_002",
  median_income = "B19013_001"
)
renter_vars <- c(
  renter_num = "B25003_003",
  renter_den = "B25003_001"
)

build_hardship_index <- function(raw_df) {
  raw_df %>%
    select(-moe) %>%
    pivot_wider(names_from = variable, values_from = estimate) %>%
    mutate(
      poverty_rate = (pov_below50 + pov_50to99) / pov_den,
      unemp_rate   = unemp_num / unemp_den
    ) %>%
    filter(!is.na(poverty_rate), !is.na(unemp_rate), !is.na(median_income)) %>%
    (function(df) {
      geom <- if (inherits(df, "sf")) st_geometry(df) else NULL
      tbl  <- st_drop_geometry(df)
      z_pov   <- as.numeric(scale(tbl$poverty_rate))
      z_unemp <- as.numeric(scale(tbl$unemp_rate))
      z_inc   <- -1 * as.numeric(scale(tbl$median_income))
      tbl$hardship_index <- rowMeans(cbind(z_pov, z_unemp, z_inc), na.rm = TRUE)
      if (!is.null(geom)) st_sf(tbl, geometry = geom) else tbl
    })()
}
```

```{r}
#| label: pull-national
#| cache: true
county_raw <- get_acs(
  geography    = "county",
  variables    = hardship_vars,
  year         = 2023,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
county_2023 <- build_hardship_index(county_raw) %>%
  filter(!str_sub(GEOID, 1, 2) %in% c("72","78","66","60","69"))
```

```{r}
#| label: pull-az-tracts
#| cache: true
# All Arizona census tracts — for the state-level interactive map
az_tract_raw <- get_acs(
  geography    = "tract",
  variables    = hardship_vars,
  state        = TARGET_STATE,
  year         = 2023,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
az_tract_2023 <- build_hardship_index(az_tract_raw)

# Maricopa County boundary — derived from tract union for yellow outline
maricopa_boundary <- az_tract_2023 %>%
  filter(str_detect(NAME, TARGET_COUNTY)) %>%
  st_union() %>%
  st_sf() %>%
  st_transform(4326)
```

```{r}
#| label: pull-local-current
#| cache: true
local_combined_raw <- get_acs(
  geography    = "tract",
  variables    = c(hardship_vars, renter_vars),
  state        = TARGET_STATE,
  county       = TARGET_COUNTY,
  year         = 2023,
  survey       = "acs5",
  geometry     = TRUE,
  progress_bar = FALSE
)
local_2023 <- build_hardship_index(
  local_combined_raw %>% filter(variable %in% names(hardship_vars))
)
renter_tbl <- local_combined_raw %>%
  filter(variable %in% names(renter_vars)) %>%
  st_drop_geometry() %>% select(-moe) %>%
  pivot_wider(names_from = variable, values_from = estimate) %>%
  mutate(renter_rate = renter_num / renter_den) %>%
  select(GEOID, renter_rate)
local_2023 <- left_join(local_2023, renter_tbl, by = "GEOID")
```

```{r}
#| label: pull-temporal
#| cache: true
local_2013_raw <- get_acs(
  geography = "tract", variables = hardship_vars,
  state = TARGET_STATE, county = TARGET_COUNTY,
  year = 2013, survey = "acs5", geometry = TRUE, progress_bar = FALSE
)
local_2019_raw <- get_acs(
  geography = "tract", variables = hardship_vars,
  state = TARGET_STATE, county = TARGET_COUNTY,
  year = 2019, survey = "acs5", geometry = TRUE, progress_bar = FALSE
)
local_2016_raw <- get_acs(
  geography = "tract", variables = hardship_vars,
  state = TARGET_STATE, county = TARGET_COUNTY,
  year = 2016, survey = "acs5", geometry = FALSE, progress_bar = FALSE
)
```

```{r}
#| label: pooled-std
widen_and_rate <- function(df) {
  df %>%
    select(-moe) %>%
    pivot_wider(names_from = variable, values_from = estimate) %>%
    mutate(
      poverty_rate = (pov_below50 + pov_50to99) / pov_den,
      unemp_rate   = unemp_num / unemp_den
    ) %>%
    filter(!is.na(poverty_rate), !is.na(unemp_rate), !is.na(median_income))
}

local_2013_wide <- widen_and_rate(local_2013_raw)
local_2019_wide <- widen_and_rate(local_2019_raw)

r2013 <- local_2013_wide %>% st_drop_geometry() %>%
  select(GEOID, poverty_rate, unemp_rate, median_income)
r2019 <- local_2019_wide %>% st_drop_geometry() %>%
  select(GEOID, poverty_rate, unemp_rate, median_income)

vars_pool <- c("poverty_rate", "unemp_rate", "median_income")
pooled_stats <- map_dfr(vars_pool, function(v) {
  vals <- c(r2013[[v]], r2019[[v]])
  tibble(variable = v, mean = mean(vals, na.rm = TRUE), sd = sd(vals, na.rm = TRUE))
})

apply_pooled_z <- function(df, stats) {
  for (v in stats$variable) {
    df[[paste0("z_", v)]] <- (df[[v]] - stats$mean[stats$variable==v]) /
                              stats$sd[stats$variable==v]
  }
  df$z_median_income <- -1 * df$z_median_income
  df$hardship_index  <- rowMeans(
    df[, c("z_poverty_rate","z_unemp_rate","z_median_income")], na.rm = TRUE)
  df
}

r2013_z <- apply_pooled_z(r2013, pooled_stats)
r2019_z <- apply_pooled_z(r2019, pooled_stats)

change_df <- local_2019_wide %>%
  select(GEOID, NAME) %>%
  left_join(r2019_z %>% select(GEOID, hardship_2019 = hardship_index), by = "GEOID") %>%
  inner_join(r2013_z %>% select(GEOID, hardship_2013 = hardship_index), by = "GEOID") %>%
  mutate(hardship_change = hardship_2019 - hardship_2013)

lo      <- quantile(change_df$hardship_change, 0.02, na.rm = TRUE)
hi      <- quantile(change_df$hardship_change, 0.98, na.rm = TRUE)
max_abs <- max(abs(lo), abs(hi))
change_df <- change_df %>%
  mutate(change_plot = pmax(pmin(hardship_change, max_abs), -max_abs))
```

```{r}
#| label: lisa-analysis
nb   <- poly2nb(change_df, queen = TRUE)
w    <- nb2listw(nb, style = "W", zero.policy = TRUE)
lisa <- localmoran(change_df$hardship_2019, w, zero.policy = TRUE)
z19  <- scale(change_df$hardship_2019)[,1]
wz19 <- lag.listw(w, z19, zero.policy = TRUE)
p19  <- lisa[, "Pr(z != E(Ii))"]

change_df <- change_df %>%
  mutate(lisa_2019 = case_when(
    z19 > 0 & wz19 > 0 & p19 < 0.05 ~ "HH",
    z19 < 0 & wz19 < 0 & p19 < 0.05 ~ "LL",
    z19 > 0 & wz19 < 0 & p19 < 0.05 ~ "HL",
    z19 < 0 & wz19 > 0 & p19 < 0.05 ~ "LH",
    TRUE ~ "NS"
  ))
```

```{r}
#| label: trajectories
lisa_2013 <- localmoran(change_df$hardship_2013, w, zero.policy = TRUE)
z13  <- scale(change_df$hardship_2013)[,1]
wz13 <- lag.listw(w, z13, zero.policy = TRUE)
p13  <- lisa_2013[, "Pr(z != E(Ii))"]

local_2016_wide <- widen_and_rate(local_2016_raw %>% mutate(geometry = NULL))
r2016_z <- apply_pooled_z(
  local_2016_wide %>% select(GEOID, poverty_rate, unemp_rate, median_income),
  pooled_stats
)

change_df <- change_df %>%
  mutate(
    lisa_2013 = case_when(
      z13 > 0 & wz13 > 0 & p13 < 0.05 ~ "HH",
      z13 < 0 & wz13 < 0 & p13 < 0.05 ~ "LL",
      z13 > 0 & wz13 < 0 & p13 < 0.05 ~ "HL",
      z13 < 0 & wz13 > 0 & p13 < 0.05 ~ "LH",
      TRUE ~ "NS"
    ),
    trajectory = case_when(
      lisa_2013 == "HH" & lisa_2019 == "HH" ~ "Persistent HH",
      lisa_2013 != "HH" & lisa_2019 == "HH" ~ "Emerging HH",
      lisa_2013 == "HH" & lisa_2019 != "HH" ~ "Dissolving HH",
      lisa_2013 == "LL" & lisa_2019 == "LL" ~ "Persistent LL",
      lisa_2013 != "LL" & lisa_2019 == "LL" ~ "Emerging LL",
      lisa_2013 == "HL" | lisa_2019 == "HL" ~ "HL Outlier",
      TRUE                                   ~ "Stable NS"
    )
  ) %>%
  left_join(r2016_z %>% select(GEOID, hardship_2016 = hardship_index), by = "GEOID") %>%
  mutate(trend_type = case_when(
    hardship_2013 > hardship_2016 & hardship_2016 > hardship_2019 ~ "Consistently improving",
    hardship_2013 < hardship_2016 & hardship_2016 < hardship_2019 ~ "Consistently worsening",
    hardship_2013 > hardship_2016 & hardship_2016 < hardship_2019 ~ "V-shaped (improved mid)",
    hardship_2013 < hardship_2016 & hardship_2016 > hardship_2019 ~ "Inverted-V (worsened mid)",
    TRUE ~ "Flat / non-monotone"
  ))
```

```{r}
#| label: summary-stats
moran_global <- moran.test(change_df$hardship_2019, w, zero.policy = TRUE)

n_hh           <- sum(change_df$lisa_2019 == "HH", na.rm = TRUE)
n_ll           <- sum(change_df$lisa_2019 == "LL", na.rm = TRUE)
n_tracts       <- nrow(change_df)
pct_improved   <- round(100 * mean(change_df$hardship_change < -0.05, na.rm=TRUE), 1)
pct_worsened   <- round(100 * mean(change_df$hardship_change >  0.05, na.rm=TRUE), 1)
pct_persist_hh <- round(100 * mean(change_df$trajectory == "Persistent HH", na.rm=TRUE), 1)
pct_emerg_hh   <- round(100 * mean(change_df$trajectory == "Emerging HH",   na.rm=TRUE), 1)
n_persist_hh   <- sum(change_df$trajectory == "Persistent HH", na.rm = TRUE)
n_emerg_hh     <- sum(change_df$trajectory == "Emerging HH",   na.rm = TRUE)

n_counties <- sum(!is.na(county_2023$hardship_index))

traj_colors <- c(
  "Persistent HH" = "#db2b27", "Emerging HH"   = "#fdbf11",
  "Dissolving HH" = "#fdd870", "Persistent LL" = "#1696d2",
  "Emerging LL"   = "#73bfe2", "HL Outlier"    = "#ec008b",
  "Stable NS"     = "#d2d2d2"
)
lisa_colors <- c(
  "HH" = "#db2b27", "LL" = "#1696d2",
  "HL" = "#fdbf11", "LH" = "#73bfe2", "NS" = "#d2d2d2"
)

theme_urbn_dash <- function(base_size = 11) {
  theme_void(base_size = base_size) +
  theme(
    plot.background   = element_rect(fill = "#ffffff", color = NA),
    panel.background  = element_rect(fill = "#ffffff", color = NA),
    plot.title        = element_text(color = "#333333", size = base_size + 1,
                                     face = "bold", margin = margin(b = 3)),
    plot.subtitle     = element_text(color = "#767676", size = base_size - 1,
                                     margin = margin(b = 6)),
    plot.caption      = element_text(color = "#767676", size = base_size - 2,
                                     hjust = 0, margin = margin(t = 4)),
    legend.background = element_rect(fill = "#ffffff", color = NA),
    legend.key        = element_rect(fill = "#ffffff", color = NA),
    legend.text       = element_text(color = "#333333", size = base_size - 1),
    legend.title      = element_text(color = "#767676", size = base_size - 1),
    plot.margin       = margin(8, 8, 8, 8)
  )
}

# National context rankings
us_ranks <- county_2023 %>%
  st_drop_geometry() %>%
  filter(!is.na(hardship_index)) %>%
  mutate(us_rank = rank(-hardship_index, ties.method = "min")) %>%
  select(GEOID, us_rank)

az_counties <- county_2023 %>%
  st_drop_geometry() %>%
  filter(str_detect(NAME, "Arizona")) %>%
  left_join(us_ranks, by = "GEOID") %>%
  arrange(desc(hardship_index)) %>%
  mutate(
    az_rank     = row_number(),
    n_az        = n(),
    county_name = str_remove(NAME, " County, Arizona"),
    County = case_when(
      az_rank == 1    ~ paste0(county_name, " (Worst in AZ)"),
      az_rank == n_az ~ paste0(county_name, " (Best in AZ)"),
      county_name == TARGET_COUNTY ~ paste0(county_name, " ← your county"),
      TRUE ~ county_name
    ),
    `Economic Hardship Index` = round(hardship_index, 3)
  ) %>%
  select(`AZ Rank` = az_rank, County, `Economic Hardship Index`,
         `US Rank` = us_rank)
```

# National Context

## Column {width=55%}

```{r}
#| label: map-national
#| title: "Economic Hardship Across U.S. Counties"
county_map <- county_2023 %>%
  filter(!is.na(hardship_index)) %>%
  st_simplify(preserveTopology = TRUE, dTolerance = 5000) %>%
  st_transform(4326)

az_outline <- states(cb = TRUE) %>%
  filter(STUSPS == TARGET_STATE) %>%
  st_transform(4326)

maplibre(
  style  = carto_style("dark-matter"),
  bounds = c(-125.0, 24.0, -66.0, 50.0)
) |>
  add_fill_layer(
    id = "hardship-fill",
    source = county_map,
    fill_color = interpolate(
      column = "hardship_index",
      values = c(-2, -1, 0, 1, 2),
      stops  = c("#1696d2", "#73bfe2", "#f5f5f5", "#ec6a5a", "#db2b27")
    ),
    fill_opacity = 0.8,
    tooltip = "NAME"
  ) |>
  add_line_layer(
    id = "county-borders",
    source = county_map,
    line_color = "#333333",
    line_width = 0.2
  ) |>
  add_line_layer(
    id = "az-highlight",
    source = az_outline,
    line_color = "#fdbf11",
    line_width = 2.5
  ) |>
  add_legend(
    "Economic Hardship Index",
    values   = c("Worse-off", "", "Average", "", "Better-off"),
    colors   = c("#db2b27", "#ec6a5a", "#f5f5f5", "#73bfe2", "#1696d2"),
    position = "bottom-left",
    width    = 220
  )
```

## Column {width=45%}

### Row {height=35%}

```{r}
#| label: table-az-counties
#| title: "Arizona Counties — Economic Hardship Rankings"
az_counties %>%
  DT::datatable(
    options = list(pageLength = 8, dom = "tp"),
    rownames = FALSE
  ) %>%
  DT::formatStyle(
    columns = "County",
    target  = "row",
    backgroundColor = DT::styleEqual(
      paste0(TARGET_COUNTY, " \u2190 your county"),
      "#fffde6"
    )
  )
```

### Row {height=65%}

```{r}
#| label: card-dot-compare
#| title: "All Arizona Counties vs. U.S. Extremes"
#| fig-height: 6
top5_us <- county_2023 %>%
  st_drop_geometry() %>%
  left_join(us_ranks, by = "GEOID") %>%
  filter(!is.na(hardship_index)) %>%
  slice_min(us_rank, n = 5, with_ties = FALSE) %>%
  mutate(
    state_name   = str_extract(NAME, "(?<=, ).+"),
    state_abb    = state.abb[match(state_name, state.name)],
    county_short = str_remove(NAME, " County.*"),
    group = "Top 5 US (Most Economic Hardship)",
    label = paste0(county_short, ", ", coalesce(state_abb, state_name))
  )

bot5_us <- county_2023 %>%
  st_drop_geometry() %>%
  left_join(us_ranks, by = "GEOID") %>%
  filter(!is.na(hardship_index)) %>%
  slice_max(us_rank, n = 5, with_ties = FALSE) %>%
  mutate(
    state_name   = str_extract(NAME, "(?<=, ).+"),
    state_abb    = state.abb[match(state_name, state.name)],
    county_short = str_remove(NAME, " County.*"),
    group = "Bottom 5 US (Least Economic Hardship)",
    label = paste0(county_short, ", ", coalesce(state_abb, state_name))
  )

az_all <- county_2023 %>%
  st_drop_geometry() %>%
  left_join(us_ranks, by = "GEOID") %>%
  filter(str_detect(NAME, "Arizona"), !is.na(hardship_index)) %>%
  mutate(
    group = "Arizona Counties",
    label = paste0(str_remove(NAME, " County, Arizona"), ", AZ")
  )

compare_df <- bind_rows(top5_us, az_all, bot5_us) %>%
  mutate(
    color_group = case_when(
      group == "Arizona Counties"                    ~ "#fdbf11",
      group == "Top 5 US (Most Economic Hardship)"   ~ "#db2b27",
      TRUE                                           ~ "#1696d2"
    ),
    label = fct_reorder(label, hardship_index)
  )

az_mean    <- mean(compare_df$hardship_index[compare_df$group == "Arizona Counties"],
                   na.rm = TRUE)
target_val <- compare_df$hardship_index[str_detect(compare_df$label, TARGET_COUNTY)][1]
mid_label  <- levels(compare_df$label)[ceiling(nlevels(compare_df$label) / 2)]

p <- ggplot(compare_df,
       aes(x = hardship_index, y = label, color = color_group,
           text = paste0(label, "\nIndex: ", round(hardship_index, 3)))) +
  geom_point(size = 2.8) +
  scale_color_identity() +
  theme_urbn_dash(base_size = 10) +
  theme(
    axis.text.y        = element_text(size = 8, color = "#333333"),
    axis.text.x        = element_text(size = 9, color = "#767676"),
    panel.grid.major.x = element_line(color = "#eeeeee", linewidth = 0.4)
  ) +
  labs(x = "Economic Hardship Index", y = NULL)

ggplotly(p, tooltip = "text") %>%
  layout(
    title  = list(
      text = paste0("All Arizona Counties vs. U.S. Extremes",
                    "<br><sup><span style=\"color:#fdbf11\">&#9679;</span> AZ Counties &nbsp;",
                    "<span style=\"color:#db2b27\">&#9679;</span> 5 Highest US &nbsp;",
                    "<span style=\"color:#1696d2\">&#9679;</span> 5 Lowest US &nbsp;",
                    "<span style=\"color:#9d9d9d\">&#9474;</span> AZ Mean</sup>"),
      font = list(size = 12, color = "#333333")
    ),
    margin = list(t = 50, b = 30),
    xaxis  = list(title = list(text = "Economic Hardship Index",
                               font = list(size = 11, color = "#555555"))),
    yaxis  = list(automargin = TRUE),
    shapes = list(list(
      type = "line",
      x0 = az_mean, x1 = az_mean, y0 = 0, y1 = 1, yref = "paper",
      line = list(color = "#9d9d9d", width = 1.5, dash = "dot")
    ))
  ) %>%
  config(displayModeBar = FALSE)
```

# Arizona in Focus

## Column {width=60%}

```{r}
#| label: map-az-tracts
#| title: "Economic Hardship: Arizona Census Tracts (2023)"
az_map <- az_tract_2023 %>%
  filter(!is.na(hardship_index)) %>%
  st_simplify(preserveTopology = TRUE, dTolerance = 500) %>%
  st_transform(4326)

maplibre(
  style  = carto_style("dark-matter"),
  bounds = c(-115.0, 31.0, -108.5, 37.5)
) |>
  add_fill_layer(
    id           = "az-tracts",
    source       = az_map,
    fill_color   = interpolate(
      column = "hardship_index",
      values = c(-2, -1, 0, 1, 2),
      stops  = c("#1696d2", "#73bfe2", "#f5f5f5", "#ec6a5a", "#db2b27")
    ),
    fill_opacity = 0.8,
    tooltip      = "NAME"
  ) |>
  add_line_layer(
    id           = "maricopa-border",
    source       = maricopa_boundary,
    line_color   = "#fdbf11",
    line_width   = 2.5
  ) |>
  add_legend(
    "Economic Hardship Index",
    values   = c("Worse-off", "", "Average", "", "Better-off"),
    colors   = c("#db2b27", "#ec6a5a", "#f5f5f5", "#73bfe2", "#1696d2"),
    position = "bottom-left",
    width    = 220
  )
```

## Column {width=40%}

### Row {height=72%}

```{r}
#| label: map-lisa
#| title: "LISA Clusters — Statistically Significant Hardship Patterns"
#| fig-height: 8
ggplot(change_df) +
  geom_sf(aes(fill = lisa_2019), color = NA) +
  scale_fill_manual(
    values = lisa_colors,
    labels = c(
      "HH" = "Hot Spot — High Hardship Cluster",
      "LL" = "Cold Spot — Low Hardship Cluster",
      "HL" = "High-Low Outlier",
      "LH" = "Low-High Outlier",
      "NS" = "Not Statistically Significant"
    ),
    name = "LISA Cluster\n(2019, p < 0.05)"
  ) +
  theme_urbn_dash(base_size = 10) +
  labs(
    title    = paste0("Hardship Clusters: ", TARGET_LABEL),
    subtitle = "Census tracts where hardship is statistically concentrated among neighbors",
    caption  = paste0("Global Moran's I = ",
                      round(moran_global$estimate[1], 3),
                      " (p < 0.001) — strong positive spatial autocorrelation")
  )
```

### Row {height=28%}

```{r}
#| label: vb-hh
#| content: valuebox
list(title = "Hot Spot Tracts (HH)",
     value = n_hh, icon = "fire", color = "danger")
```

```{r}
#| label: vb-ll
#| content: valuebox
list(title = "Cold Spot Tracts (LL)",
     value = n_ll, icon = "snow", color = "primary")
```

```{r}
#| label: vb-moran
#| content: valuebox
list(title = "Global Moran's I",
     value = round(unname(moran_global$estimate[1]), 3),
     icon = "diagram-3", color = "secondary")
```

# Hardship Trajectories: 2013–2019

## Column {width=55%}

```{r}
#| label: map-trajectory
#| title: "Neighborhood Hardship Trajectories (2013–2019)"
#| fig-height: 9
ggplot(change_df) +
  geom_sf(aes(fill = trajectory), color = NA) +
  scale_fill_manual(
    values = traj_colors,
    labels = c(
      "Persistent HH" = "Persistent Hot Spot — entrenched hardship",
      "Emerging HH"   = "Emerging Hot Spot — newly forming cluster",
      "Dissolving HH" = "Dissolving Hot Spot — improving cluster",
      "Persistent LL" = "Persistent Cold Spot — stable low hardship",
      "Emerging LL"   = "Emerging Cold Spot",
      "HL Outlier"    = "Spatial Outlier",
      "Stable NS"     = "No Dominant Pattern"
    ),
    name = "Trajectory\n(2013 → 2019)"
  ) +
  theme_urbn_dash(base_size = 10) +
  labs(
    title    = paste0("Neighborhood Economic Hardship Trajectories: ", TARGET_LABEL),
    subtitle = "LISA cluster transition 2013–2019 | Pooled z-score standardization",
    caption  = "Red = persistent/emerging hardship | Blue = stable/improving low-hardship areas"
  )
```

## Column {width=45%}

### Row {height=55%}

```{r}
#| label: map-change
#| title: "Economic Hardship Change (2013–2019)"
#| fig-height: 6
ggplot(change_df) +
  geom_sf(aes(fill = change_plot), color = NA) +
  scale_fill_gradient2(
    low      = "#0a4c6a",
    mid      = "#f5f5f5",
    high     = "#db2b27",
    midpoint = 0,
    limits   = c(-max_abs, max_abs),
    name     = "Economic\nHardship Change"
  ) +
  theme_urbn_dash(base_size = 10) +
  labs(
    title    = paste0("Economic Hardship Change (2013→2019): ", TARGET_LABEL),
    subtitle = "Blue = improved | Red = worsened | Winsorized at 2nd/98th percentile",
    caption  = "Pooled z-score standardization — scores are comparable across years"
  )
```

### Row {height=45%}

```{r}
#| label: vb-improved
#| content: valuebox
list(title = "Tracts Improved (2013→2019)",
     value = paste0(pct_improved, "%"),
     icon  = "arrow-down-circle", color = "success")
```

```{r}
#| label: vb-worsened
#| content: valuebox
list(title = "Tracts Worsened (2013→2019)",
     value = paste0(pct_worsened, "%"),
     icon  = "arrow-up-circle", color = "warning")
```

```{r}
#| label: vb-persistent
#| content: valuebox
list(title = "Persistently High Hardship",
     value = paste0(pct_persist_hh, "%"),
     icon  = "exclamation-circle", color = "danger")
```

```{r}
#| label: vb-emerging
#| content: valuebox
list(title = "Emerging Hot Spots",
     value = paste0(pct_emerg_hh, "%"),
     icon  = "arrow-up-right-circle", color = "warning")
```

# Policy Implications

## Column {width=50%}

::: {.card style="border-left: 5px solid #db2b27; margin-bottom: 12px;"}
### 🔴 Areas of Persistent Concern

**`r n_persist_hh` Persistent Hot Spot Tracts (`r pct_persist_hh`% of all tracts)**

These census tracts had statistically significant high-hardship clustering in **both** 2013 and 2019. The pattern is not random — a Global Moran's I of `r round(moran_global$estimate[1], 3)` confirms that hardship is spatially concentrated, not scattered. Tracts in this category share infrastructure deficits, limited employment access, and concentrated poverty that reinforce one another across neighborhood boundaries.

**Implication:** Individual-level interventions alone are unlikely to move the needle. Place-based, multi-sector investment is required.

*[TODO: Identify the specific geographic corridor in your county where these tracts are concentrated.]*
:::

::: {.card style="border-left: 5px solid #fdbf11; margin-bottom: 12px;"}
### 🟡 Early Warning Signals

**`r n_emerg_hh` Emerging Hot Spot Tracts (`r pct_emerg_hh`% of all tracts)**

These tracts were **not** significant hardship clusters in 2013 but became statistically significant by 2019 — representing the spatial expansion of hardship beyond historically distressed cores. This is an early warning signal that hardship is spreading, not contained.

**Displacement paradox:** Some "improving" tracts nearby may be gentrifying, pushing lower-income households outward into these emerging clusters. Declining hardship scores do not necessarily mean existing residents are better off.

*[TODO: Identify where emerging hot spots are forming in your county and what may be driving displacement.]*
:::

::: {.card style="border-left: 5px solid #55b748;"}
### 🟢 Signs of Progress — Read With Caution

**`r pct_improved`% of tracts showed improvement (2013→2019)**

The majority of tracts improved over the post-recession recovery window. However, aggregate improvement masks significant variation, and the data cannot distinguish genuine economic uplift from population turnover. A tract with a declining hardship index may simply have replaced lower-income residents with higher-income newcomers.

**Data limitation:** Before drawing conclusions from improving scores, ground-truth verification through community engagement and displacement tracking is essential.

*[TODO: Customize with your county's specific context.]*
:::

## Column {width=50%}

::: {.card}
### Recommendation 1
**TODO:** Write your first recommendation targeting the `r n_persist_hh` Persistent HH tracts.
Name the specific geographic corridor, cite the hardship index values, and propose a
concrete place-based intervention with a named responsible entity.

*[Your answer here — minimum 3 sentences]*
:::

::: {.card}
### Recommendation 2
**TODO:** Write your second recommendation targeting the `r n_emerg_hh` Emerging HH tracts.
Reference the displacement paradox, identify where these clusters are forming,
and propose an early-intervention or monitoring strategy.

*[Your answer here — minimum 3 sentences]*
:::

::: {.card}
### Recommendation 3
**TODO:** Write your third recommendation using the temporal evidence —
the trajectory map, change scores, and the Moran's I statistic of
`r round(moran_global$estimate[1], 3)`. Make a forward-looking
data-monitoring or cross-sector coordination argument.

*[Your answer here — minimum 3 sentences]*
:::
