---
title: "Economic Hardship in Maricopa County, AZ"
subtitle: "A Policy Brief"
author: "[Your Name]"
date: today
format:
  html:
    theme: readable
    toc: false
    code-fold: true
    code-summary: "Show code"
    self-contained: true
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
---

::: {.callout-important}
## Before You Begin
1. **Rename** this file `Policy_Brief_YourLastName.qmd`.
2. **Update the county block** (TARGET_STATE, TARGET_COUNTY, TARGET_LABEL).
   This controls all data pulls and figure titles — no other code changes needed.
3. **Knit the file** to see your county's results.
4. **Write your answers** in each [Your answer here] section.
5. **Submit** both `.qmd` and `.html` to Canvas.
:::

```{r}
#| label: setup
#| include: false
options(tigris_use_cache = TRUE)
options(tigris_progress_bar = FALSE)
set.seed(42)
```

```{r}
#| label: packages
#| include: false
if (!file.exists("~/.paf516_ready")) {
  if (!requireNamespace("renv", quietly = TRUE)) install.packages("renv")
  tmp <- tempfile(fileext = ".lock")
  download.file(
    url      = "https://raw.githubusercontent.com/AntJam-Howell/paf-516-labs/main/renv.lock",
    destfile = tmp, mode = "wb"
  )
  renv::restore(lockfile = tmp, prompt = FALSE)
  file.create("~/.paf516_ready")
}
library(tidyverse); library(tidycensus); library(sf); library(spdep)
library(viridis); library(patchwork); library(scales)
```

```{r}
#| label: county-config
TARGET_STATE  <- "AZ"
TARGET_COUNTY <- "Maricopa"
TARGET_LABEL  <- "Maricopa County, AZ"
```

```{r}
#| label: index-config
hardship_vars <- c(
  pov_below50   = "C17002_002",
  pov_50to99    = "C17002_003",
  pov_den       = "C17002_001",
  unemp_num     = "B23025_005",
  unemp_den     = "B23025_002",
  median_income = "B19013_001"
)

build_hardship_index <- function(raw_df) {
  raw_df %>%
    select(-moe) %>%
    pivot_wider(names_from = variable, values_from = estimate) %>%
    mutate(
      poverty_rate = (pov_below50 + pov_50to99) / pov_den,
      unemp_rate   = unemp_num / unemp_den
    ) %>%
    filter(!is.na(poverty_rate), !is.na(unemp_rate), !is.na(median_income)) %>%
    (function(df) {
      geom <- if (inherits(df, "sf")) st_geometry(df) else NULL
      tbl  <- st_drop_geometry(df)
      z_pov  <- as.numeric(scale(tbl$poverty_rate))
      z_unemp <- as.numeric(scale(tbl$unemp_rate))
      z_inc  <- -1 * as.numeric(scale(tbl$median_income))
      tbl$hardship_index <- rowMeans(cbind(z_pov, z_unemp, z_inc), na.rm = TRUE)
      if (!is.null(geom)) st_sf(tbl, geometry = geom) else tbl
    })()
}
```

```{r}
#| label: pull-local
#| cache: true
local_raw <- get_acs(
  geography = "tract", variables = hardship_vars,
  state = TARGET_STATE, county = TARGET_COUNTY,
  year = 2023, survey = "acs5", geometry = TRUE, progress_bar = FALSE
)
local_2023 <- build_hardship_index(local_raw)
```

```{r}
#| label: pull-temporal
#| cache: true
local_2013_raw <- get_acs(
  geography = "tract", variables = hardship_vars,
  state = TARGET_STATE, county = TARGET_COUNTY,
  year = 2013, survey = "acs5", geometry = TRUE, progress_bar = FALSE
)
local_2019_raw <- get_acs(
  geography = "tract", variables = hardship_vars,
  state = TARGET_STATE, county = TARGET_COUNTY,
  year = 2019, survey = "acs5", geometry = TRUE, progress_bar = FALSE
)
local_2016_raw <- get_acs(
  geography = "tract", variables = hardship_vars,
  state = TARGET_STATE, county = TARGET_COUNTY,
  year = 2016, survey = "acs5", geometry = FALSE, progress_bar = FALSE
)
```

```{r}
#| label: pooled-std
widen_and_rate <- function(df) {
  df %>%
    select(-moe) %>%
    pivot_wider(names_from = variable, values_from = estimate) %>%
    mutate(
      poverty_rate = (pov_below50 + pov_50to99) / pov_den,
      unemp_rate   = unemp_num / unemp_den
    ) %>%
    filter(!is.na(poverty_rate), !is.na(unemp_rate), !is.na(median_income))
}

local_2013_wide <- widen_and_rate(local_2013_raw)
local_2019_wide <- widen_and_rate(local_2019_raw)

r2013 <- local_2013_wide %>% st_drop_geometry() %>%
  select(GEOID, poverty_rate, unemp_rate, median_income)
r2019 <- local_2019_wide %>% st_drop_geometry() %>%
  select(GEOID, poverty_rate, unemp_rate, median_income)

vars_pool <- c("poverty_rate", "unemp_rate", "median_income")
pooled_stats <- map_dfr(vars_pool, function(v) {
  vals <- c(r2013[[v]], r2019[[v]])
  tibble(variable = v, mean = mean(vals, na.rm = TRUE), sd = sd(vals, na.rm = TRUE))
})

apply_pooled_z <- function(df, stats) {
  for (v in stats$variable) {
    df[[paste0("z_", v)]] <- (df[[v]] - stats$mean[stats$variable==v]) /
                              stats$sd[stats$variable==v]
  }
  df$z_median_income <- -1 * df$z_median_income
  df$hardship_index  <- rowMeans(
    df[, c("z_poverty_rate","z_unemp_rate","z_median_income")], na.rm = TRUE)
  df
}

r2013_z <- apply_pooled_z(r2013, pooled_stats)
r2019_z <- apply_pooled_z(r2019, pooled_stats)

change_df <- local_2019_wide %>%
  select(GEOID, NAME) %>%
  left_join(r2019_z %>% select(GEOID, hardship_2019 = hardship_index), by = "GEOID") %>%
  inner_join(r2013_z %>% select(GEOID, hardship_2013 = hardship_index), by = "GEOID") %>%
  mutate(hardship_change = hardship_2019 - hardship_2013)
```

```{r}
#| label: lisa-analysis
nb    <- poly2nb(change_df, queen = TRUE)
w     <- nb2listw(nb, style = "W", zero.policy = TRUE)
lisa  <- localmoran(change_df$hardship_2019, w, zero.policy = TRUE)
z19   <- scale(change_df$hardship_2019)[,1]
wz19  <- lag.listw(w, z19, zero.policy = TRUE)
p19   <- lisa[, "Pr(z != E(Ii))"]

change_df <- change_df %>%
  mutate(lisa_2019 = case_when(
    z19 > 0 & wz19 > 0 & p19 < 0.05 ~ "HH",
    z19 < 0 & wz19 < 0 & p19 < 0.05 ~ "LL",
    z19 > 0 & wz19 < 0 & p19 < 0.05 ~ "HL",
    z19 < 0 & wz19 > 0 & p19 < 0.05 ~ "LH",
    TRUE ~ "NS"
  ))

moran_global <- moran.test(change_df$hardship_2019, w, zero.policy = TRUE)
n_hh <- sum(change_df$lisa_2019 == "HH", na.rm = TRUE)
```

```{r}
#| label: trajectories
lisa_2013  <- localmoran(change_df$hardship_2013, w, zero.policy = TRUE)
z13  <- scale(change_df$hardship_2013)[,1]
wz13 <- lag.listw(w, z13, zero.policy = TRUE)
p13  <- lisa_2013[, "Pr(z != E(Ii))"]

local_2016_wide <- widen_and_rate(local_2016_raw %>% mutate(geometry = NULL))
r2016_z <- apply_pooled_z(
  local_2016_wide %>% select(GEOID, poverty_rate, unemp_rate, median_income),
  pooled_stats
)

change_df <- change_df %>%
  mutate(
    lisa_2013 = case_when(
      z13 > 0 & wz13 > 0 & p13 < 0.05 ~ "HH",
      z13 < 0 & wz13 < 0 & p13 < 0.05 ~ "LL",
      z13 > 0 & wz13 < 0 & p13 < 0.05 ~ "HL",
      z13 < 0 & wz13 > 0 & p13 < 0.05 ~ "LH",
      TRUE ~ "NS"
    ),
    trajectory = case_when(
      lisa_2013 == "HH" & lisa_2019 == "HH" ~ "Persistent HH",
      lisa_2013 != "HH" & lisa_2019 == "HH" ~ "Emerging HH",
      lisa_2013 == "HH" & lisa_2019 != "HH" ~ "Dissolving HH",
      lisa_2013 == "LL" & lisa_2019 == "LL" ~ "Persistent LL",
      lisa_2013 != "LL" & lisa_2019 == "LL" ~ "Emerging LL",
      lisa_2013 == "HL" | lisa_2019 == "HL" ~ "HL Outlier",
      TRUE                                   ~ "Stable NS"
    )
  ) %>%
  left_join(r2016_z %>% select(GEOID, hardship_2016 = hardship_index), by = "GEOID") %>%
  mutate(trend_type = case_when(
    hardship_2013 > hardship_2016 & hardship_2016 > hardship_2019 ~ "Consistently improving",
    hardship_2013 < hardship_2016 & hardship_2016 < hardship_2019 ~ "Consistently worsening",
    hardship_2013 > hardship_2016 & hardship_2016 < hardship_2019 ~ "V-shaped (improved mid)",
    hardship_2013 < hardship_2016 & hardship_2016 > hardship_2019 ~ "Inverted-V (worsened mid)",
    TRUE ~ "Flat / non-monotone"
  ))
```

```{r}
#| label: summary-stats
traj_colors <- c(
  "Persistent HH"="#D7191C", "Emerging HH"="#FDAE61", "Dissolving HH"="#FEE08B",
  "Persistent LL"="#2C7BB6", "Emerging LL"="#ABD9E9", "HL Outlier"="#9E5BB4",
  "Stable NS"="#CCCCCC"
)
lisa_colors <- c("HH"="#D7191C","LL"="#2C7BB6","HL"="#FDAE61","LH"="#ABD9E9","NS"="#CCCCCC")

pct_persist_hh <- round(100 * mean(change_df$trajectory == "Persistent HH", na.rm=TRUE), 1)
pct_emerg_hh   <- round(100 * mean(change_df$trajectory == "Emerging HH",   na.rm=TRUE), 1)
pct_dissolving <- round(100 * mean(change_df$trajectory == "Dissolving HH", na.rm=TRUE), 1)
```

::: {.callout-note}
## Executive Summary
**[Your answer here — 3-4 sentences]**
Name the county. State hardship level. Describe the spatial pattern.
State your top recommendation.
:::

## Background

This policy brief examines the spatial distribution and temporal trajectory of economic hardship across census tracts in `r TARGET_LABEL`. The analysis constructs a composite hardship index from three American Community Survey indicators — the poverty rate, the unemployment rate, and median household income (reverse-coded) — standardized and averaged into a single z-score measure. This index identifies neighborhoods where multiple dimensions of economic disadvantage overlap, providing a more nuanced picture than any single indicator alone.

The temporal analysis compares hardship levels in the 2009–2013 and 2015–2019 ACS 5-year windows, spanning the post-recession recovery decade. Using Local Indicators of Spatial Association (LISA), the analysis identifies statistically significant hot spots (clusters of high hardship) and cold spots (clusters of low hardship) at each time point, then classifies each tract's trajectory — persistent, emerging, or dissolving — to reveal where hardship is entrenching, spreading, or receding. A three-point check using the 2012–2016 midpoint distinguishes monotonic trends from V-shaped or inverted-V patterns. *If you selected a different county, review whether these temporal patterns match your local context.*

# Key Findings

## Finding 1 — Where Is Hardship Concentrated?

```{r}
#| label: fig-lisa
#| fig-width: 8
#| fig-height: 6
ggplot(change_df) +
  geom_sf(aes(fill = lisa_2019), color = NA) +
  scale_fill_manual(values = lisa_colors,
    labels = c("HH" = "Hot Spot (HH)", "LL" = "Cold Spot (LL)",
               "HL" = "High-Low Outlier", "LH" = "Low-High Outlier",
               "NS" = "Not Significant"),
    name = "LISA Cluster\n(2019, p<0.05)") +
  theme_void(base_size = 11) +
  labs(title    = paste0("Spatial Hardship Clusters: ", TARGET_LABEL),
       subtitle = "LISA analysis on ACS 2019 5-Year economic hardship index",
       caption  = paste0("Global Moran's I = ", round(moran_global$estimate[1], 3),
                         " (p < 0.001) | Queen contiguity, row-standardized weights"))
```

How many hot spot tracts? Where are they located? What does the Moran's I statistic tell you about the degree of spatial clustering? Reference specific numbers from the map.

*YOUR ANSWER HERE*

---

## Finding 2 — Which Neighborhoods Are Getting Worse?

```{r}
#| label: fig-trajectory
#| fig-width: 8
#| fig-height: 6
ggplot(change_df) +
  geom_sf(aes(fill = trajectory), color = NA) +
  scale_fill_manual(values = traj_colors,
    labels = c("Persistent HH" = "Persistent Hot Spot",
               "Emerging HH"   = "Emerging Hot Spot",
               "Dissolving HH" = "Dissolving Hot Spot",
               "Persistent LL" = "Persistent Cold Spot",
               "Emerging LL"   = "Emerging Cold Spot",
               "HL Outlier"    = "Spatial Outlier",
               "Stable NS"     = "No Dominant Pattern"),
    name = "Trajectory\n(2013→2019)") +
  theme_void(base_size = 11) +
  labs(title    = paste0("Neighborhood Hardship Trajectories: ", TARGET_LABEL),
       subtitle = "LISA cluster transition, 2013–2019 | Pooled z-score standardization",
       caption  = "Pooled mean/SD across both years — comparable change scores")
```

What share of tracts are persistent vs. emerging vs. dissolving hot spots? Where are emerging hot spots located? What does the pattern suggest about the direction of change in your county?

*YOUR ANSWER HERE*

---

## Finding 3 — The Hardship Profile

```{r}
#| label: fig-profile
#| fig-width: 7
#| fig-height: 4
traj_summary <- change_df %>%
  st_drop_geometry() %>%
  filter(trajectory != "Stable NS") %>%
  group_by(trajectory) %>%
  summarise(mean_hardship = mean(hardship_2019, na.rm = TRUE),
            n = n(), .groups = "drop") %>%
  arrange(desc(mean_hardship))

ggplot(traj_summary, aes(x = reorder(trajectory, mean_hardship),
                          y = mean_hardship, fill = trajectory)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(values = traj_colors) +
  coord_flip() +
  theme_minimal(base_size = 11) +
  labs(title    = paste0("Mean Hardship by Trajectory Type: ", TARGET_LABEL),
       subtitle = "Pooled z-score standardized hardship index (2019)",
       x = NULL, y = "Mean Hardship Index (2019)",
       caption  = paste0("Excludes Stable NS tracts (n = ",
                         sum(change_df$trajectory == "Stable NS", na.rm = TRUE), ")"))
```

Which trajectory type has the highest average hardship level? How do persistent hot spots compare to dissolving ones? What does this tell you about the pace of recovery?

*YOUR ANSWER HERE*

---

# Recommendations

::: {.callout-tip}
## Recommendation 1 — [TODO: Add a specific policy title]
**[Your answer here — 2-3 sentences]**
Ground in Finding 1 (spatial concentration). Name a specific intervention
type and how many tracts it would target. Name a responsible entity.
:::

::: {.callout-tip}
## Recommendation 2 — [TODO: Add a specific policy title]
**[Your answer here — 2-3 sentences]**
Ground in Finding 2 (trajectories). Focus on emerging hot spots
(early intervention) or dissolving hot spots (displacement risk).
Reference the displacement paradox from Module 5.
:::

::: {.callout-tip}
## Recommendation 3 — [TODO: Add a specific policy title]
**[Your answer here — 2-3 sentences]**
Ground in Finding 3 (hardship profile). Make a data-monitoring
recommendation — what should policymakers track, how often, and why?
:::

---

# Limitations

This analysis relies on ACS 5-year estimates, which carry margins of
error that can be substantial at the census tract level. The temporal
comparison (2013–2019) reflects the post-recession recovery decade and
does not capture COVID-era disruptions. Most importantly: a declining
hardship index in a tract may reflect genuine improvement for existing
residents — or it may reflect displacement, with lower-income households
replaced by higher-income newcomers. Census aggregate statistics cannot
distinguish these processes. This analysis identifies candidate
intervention areas; on-the-ground verification is essential before
resource allocation decisions are made.
